Vacuolar H+-ATPase Determines Daughter Cell Fates through Asymmetric Segregation of the Nucleosome Remodeling and Deacetylase Complex

Version 1

Abstract

Abstract

Asymmetric cell divisions (ACDs) generate two daughter cells with identical genetic information but distinct cell fates through epigenetic mechanisms. However, the process of partitioning different epigenetic information into daughter cells remains unclear. Here, we demonstrate that the nucleosome remodeling and deacetylase (NuRD) complex is asymmetrically segregated into the surviving daughter cell rather than the apoptotic one during ACDs in Caenorhabditis elegans. The absence of NuRD triggers apoptosis via the EGL-1-CED-9-CED-4-CED-3 pathway, while an ectopic gain of NuRD enables apoptotic daughter cells to survive. We identify the vacuolar H+–adenosine triphosphatase (V-ATPase) complex as a crucial regulator of NuRD’s asymmetric segregation. V-ATPase interacts with NuRD and is asymmetrically segregated into the surviving daughter cell. Inhibition of V-ATPase disrupts cytosolic pH asymmetry and NuRD asymmetry. We suggest that asymmetric segregation of V-ATPase may cause distinct acidification levels in the two daughter cells, enabling asymmetric epigenetic inheritance that specifies their respective life-versus-death fates.

Introduction

Asymmetric cell division (ACD) gives rise to two daughter cells that possess identical genetic material but distinct cell fates, playing a crucial role in both development and tissue homeostasis (Royall and Jessberger, 2021; Sunchu and Cabernard, 2020; Venkei and Yamashita, 2018; Wooten et al., 2020; Zion et al., 2020). Both extrinsic and intrinsic mechanisms determine distinct daughter cell fates after ACD. While extrinsic mechanisms, such as exposure to signaling cues from the local niche, have been extensively studied and are known to define stem cell fate (Morrison and Spradling, 2008), the intrinsic mechanisms are more complex and largely unresolved, despite several proteins, RNA molecules, and organelles having been implicated in the regulation of some types of ACD (Sunchu and Cabernard, 2020; Zion et al., 2020). Epigenetic mechanisms play a crucial role in guiding the two daughter cells towards establishing differential gene expression profiles, ultimately defining their unique cell identities (Allis and Jenuwein, 2016; Escobar et al., 2021; Stewart-Morgan et al., 2020). During the ACD of Drosophila male germline stem cells (GSCs), pre-existing and newly synthesized histones H3 and H4 are asymmetrically segregated towards the stem daughter cell and the differentiating daughter cell, respectively (Wooten et al., 2020). Nevertheless, the extent to which epigenetic information is asymmetrically inherited through ACD in other organisms and the mechanism by which this process occurs remains elusive.

Caenorhabditis elegans represents a valuable model for investigating ACD, given its invariant cell lineage and conserved mechanisms of ACD. During hermaphrodite development in C. elegans, 131 somatic cells undergo programmed cell death, with the majority produced through ACD that create a large cell programmed for survival and a small cell programmed to die (Sulston and Horvitz, 1977; Sulston et al., 1983a) The opposing cell fates of daughter cells, i.e., to live or die, offer a compelling experimental system for investigating how epigenetic inheritance determines life versus death decisions during ACD. It is noteworthy that 105 of the 131 apoptotic cells (Q.aa and Q.pp), respectively (Figure 1A).

Asymmetric segregation of NuRD during ACDs of C. elegans Q neuroblast.

(A) Schematic of the Q neuroblast lineages. QL or QR neuroblast each generates three neurons and two apoptotic cells (Q.aa/Q.pp, X). QL produces PQR, PVM, and SDQL. QR produces AQR, AVM, and SDQR.

(B) A model of the NuRD complex composition (Bracken et al., 2019; Lai and Wade, 2011).

(C) Protein domain structure of the GFP-tagged HDAC1/2 (HDA-1) or mNeonGreen-tagged RBBP4/7 (LIN-53). CAF1C: histone-binding protein RBBP4 or subunit C of CAF1 complex; WD: WD40 repeat tryptophan-aspartate domain. Scale bar: 100 amino acids.

(D) Representative images of endogenous HDA-1::GFP, LIN-53::mNeonGreen and overexpressed MYS-1::GFP during ACDs of QR.a. In each panel, the top row shows merged images, the middle row shows mCherry-tagged plasma membrane and histone, and the bottom row shows heatmaps of GFP/mNeonGreen fluorescence. The anterior of the cell is on the left. The GFP/mNeonGreen fluorescence intensity ratio between posterior and anterior chromatids or between QR.ap and QR.aa nuclei are shown in yellow at the lower right corner of heatmaps. Time is in minutes. Other frames are in Figure S3A, and the full movies are in Supplementary Movie S3-S5. Scale bar: 2 µm.

(E) Quantification of HDA-1::GFP, LIN-53::mNeonGreen and MYS-1::GFP fluorescence intensity ratio between QR.ap and QR.aa nuclei. Data are presented as mean ± SEM. N = 10 – 12.

(F) Quantification of HDA-1::GFP (magenta), LIN-53::mNeonGreen (orange) and MYS-1::GFP (gray) fluorescence intensity ratios between the posterior and anterior half of QR.a or between QR.ap and QR.aa. The anaphase onset is defined as the last frame without chromatids segregation. Data are mean ± SEM. N = 10 – 12.

(G) Quantification of HDA-1::GFP (left) and LIN-53::mNeonGreen (right) fluorescence intensity ratios between the large and small daughters of cells shown on the X-axis. Data are mean ± SEM. N = 7 – 14.

In the nematode, the classic apoptotic pathway initiates upon activation of the BH3-only protein EGL-1 in the cells that are fated to die. EGL-1 binds to the anti-apoptotic Bcl-2-like protein CED-9, which facilitates the release of the pro-apoptotic protein CED-4, leading to caspase CED-3 activation. Activated caspase promotes the exposure of phosphatidylserine (PS) on the surface of apoptotic cells, which triggers the phagocytosis of apoptotic cells via partially redundant signaling pathways such as the CED-1/CED-6/CED-7 pathway and the CED-2/CED-5/CED-12 pathway. Proper transcriptional regulation of the egl-1 gene is vital for life-versus-death decisions in C. elegans (Conradt and Horvitz, 1998; Nehme and Conradt, 2008). Although the cell-specific transcriptional regulator has been identified in certain cells, including the sexually dimorphic HSN and CEM neurons, M4 motor neuron, NSM neuron, P11.aaap cell, ABpl/rpppapp, and the tail spike cell (Hirose et al., 2010; Jiang and Wu, 2014; Nehme and Conradt, 2008), the general upstream regulators of egl-1 at the chromatin remodeling or epigenetic level are unclear. Two NSM neuroblasts undergo ACDs to each generate a larger NSM cell programed to survive and a smaller NSM sister cell programed to die (Sulston et al., 1983b). The PIG-1 kinase-dependent asymmetric partitioning of the Snail-like transcription factor, CES-1, represses egl-1 transcription in the larger NSM sister cells, thereby preventing their apoptosis (Hatzold and Conradt, 2008; Wei et al., 2020) However, the mechanism underlying the asymmetrical activation of egl-1 in other ACDs remains elusive.

In this study, we demonstrate the enrichment of the NuRD complex in cells that are predetermined to survive, and its role in suppressing the EGL-1-CED-9-CED-4-CED-3 apoptotic pathway through repression of the egl-1 gene. Furthermore, we report the interaction between the NuRD complex and the V-ATPase complex, and reveal that during asymmetric cell division, the localization of the V-ATPase complex and cytoplasmic pH are asymmetrical, thereby contributing to the polarized segregation of the NuRD complex.

Results

NuRD asymmetric segregation during neuroblast ACDs

In order to gain insights into the molecular distinctions between apoptotic and surviving cells in C. elegans embryos, we employed the SPLiT single-cell RNA sequencing platform to compare their transcriptomes (Figure S1A-S1C) (Rosenberg et al., 2018). The expression of the critical somatic apoptosis-inducing gene egl-1 was used to distinguish the apoptotic and surviving cells (Conradt et al., 2016). Intriguingly, we identified transcripts encoding subunits of the nucleosome remodeling and deacetylase (NuRD; also known as Mi-2) complex in cells where egl-1 expression was indiscernible (Figure S1D-S1F, Supplementary Table 1). The NuRD complex represents an evolutionarily conserved protein complex associated with chromatin that couples the activities of chromatin-remodeling ATPases with histone deacetylases (Figure 1B) (Bracken et al., 2019; Lai and Wade, 2011). NuRD-mediated alterations in chromatin structure are vital for appropriate transcriptional regulation during cell fate determination and lineage commitment (Bracken et al., 2019; Lai and Wade, 2011), suggesting that NuRD may serve as a candidate epigenetic factor to specify cell fate between survival and death.

To investigate whether NuRD components are distributed asymmetrically between apoptotic and surviving daughter cells, we expressed GFP-tagged NuRD subunits within the C. elegans Q neuroblast lineages. Q.a neuroblast undergoes ACD to generate a large surviving Q.ap cell and a small apoptotic Q.aa cell (Figure 1A). We found that GFP-tagged NuRD subunits, including the histone deacetylase HDA-1, the histone binding protein LIN-53, the nucleosome remodeling factor CHD-3, and the MEP-1 protein were enriched asymmetrically in the surviving QR.ap cell during QR.a division (Figure S2, Supplementary Movie S1 and S2). Notably, NuRD asymmetric distribution occurred within several minutes after metaphase, while GFP protein translation and chromophore maturation require approximately one hour in vivo (Heim et al., 1995; Iizuka et al., 2011), suggesting that the asymmetric enrichment of NuRD is likely the result of protein redistribution.

To investigate the dynamic distribution of endogenous NuRD during ACD, we generated a GFP knock-in (KI) strain for HDA-1 and an mNeonGreen (NG, green fluorescence) KI line for LIN-53 using CRISPR-Cas9 (Figure 1C). In QR.a cells, nuclear HDA-1 and LIN-53 were released into the cytoplasm and evenly distributed until metaphase (Figure 1D, Supplementary Movie S3 and S4). At anaphase, HDA-1 and LIN-53 became enriched in the posterior part of QR.a but became less detectable in the anterior part (Figure 1D, S3A-B, Supplementary Movie S3 and S4). By quantifying the fluorescence intensity ratio between posterior and anterior chromatids of QR.a or QR.ap and QR.aa nuclei, we found that nuclear HDA-1 or LIN-53 asymmetry gradually increased from 1.1-fold at the onset of anaphase to 1.5 or 1.8-fold at cytokinesis, respectively (Figure 1D-E). We measured the ratios of fluorescence intensities between the posterior and anterior halves of QR.a or between QR.ap and QR.aa (Figure S4A, see Materials and Methods). NuRD asymmetry became evident at ∼4 minutes and reached a plateau at ∼6 minutes after the anaphase onset (time 0 min, the last frame before the sister chromosome separation; Figure 1D, 1F, S3A and S4B). QR.a spent ∼6 minutes from anaphase to the completion of cytokinesis (Chai et al., 2012; Ou et al., 2010) indicating that QR.a cell establishes NuRD asymmetry during ACD.

Similar to QR.a, the dynamic and progressively developing NuRD asymmetry was also observed in the QL.a, QR.p and QL.p cell lineages, which generate apoptotic cells (Figure 1G, S3B and S4B). In contrast, an even distribution of HDA-1 and LIN-53 was observed in two surviving daughters of QR, QL, QL.pa and QR.pa cells, which generate two viable siblings (Figure 1G and S3B). To investigate whether other epigenetic factors are also asymmetrically segregated during Q cell ACDs, we tagged the MYST family histone acetyltransferase (MYS-1 in C. elegans) with GFP. We found that MYS-1::GFP was symmetrically segregated into the apoptotic and surviving daughter cells (Figure 1D-F, S3A and S4B, Supplementary Movie S5), indicating the specificity of the polarized NuRD partition. The total fluorescence of HDA-1, LIN-53, and MYS-1 remained constant during ACDs, suggesting that protein redistribution may establish NuRD asymmetry (Figure S4C). Inhibition of a PAR-1 family kinase PIG-1 led to symmetric Q cell divisions, producing extra neurons derived from some of their apoptotic daughters (Cordes et al., 2006). In pig-1 (gm344) mutants, HDA-1 and LIN-53 were evenly partitioned during ACDs (Figure S5, Supplementary Movie S6 and S7), indicating that asymmetric NuRD segregation depends on the PIG-1 kinase.

NuRD asymmetric segregation in embryonic cell lineages

Given that 113 out of 131 somatic apoptotic events in C. elegans occur during embryonic development (Sulston et al., 1983a), we used live imaging and automated cell lineage tracing algorithms (Du et al., 2014) to track NuRD asymmetry in embryos from the two- or four-cell stage up to the 350-cell stage For example, approximately 200 minutes after the first embryonic division, ABalaappp divides to generate two viable daughters while its sibling ABalaappa divides asymmetrically and produces one daughter that undergoes apoptosis (Figure 2A). Similar to what we observed in the Q lineage, HDA-1 was evenly distributed in the two viable daughters of ABalaappp, while it was asymmetrically segregated during the ACD of ABalaappa (Figure 2B). Overall, we found that HDA-1 and LIN-53 were enriched (average ratio > 1.5) in the surviving daughters from 15 out of 17 embryonic ACDs that produce apoptotic daughter cells (Figure 2C-D and Supplementary Table 2). There were two exceptions, ABaraaaap and MSpppaa, whose two daughter cells received similar amounts of NuRD (Figure 2C-D). While MSpppaa’s apoptotic daughter cell undergoes apoptosis about 400 minutes after its generation (Hsieh et al., 2012; Sulston et al., 1983b), consistent with the idea that a low concentration of NuRD may promote apoptosis, ABaraaaap’s apoptotic daughter cell completes apoptosis within 35 minutes (Sulston et al., 1983a), suggesting that a NuRD-independent mechanism may regulate apoptosis in this cell.

NuRD Asymmetry in C. elegans embryonic cell lineages.

(A) An x and y plane of 3D time-lapse imaging data at the ∼200 min after the first embryonic division (left) in the HDA-1::GFP knock-in animal. H2B::mCherry was expressed in the embryo ubiquitously. The bright green circle indicates the ABalaappa cell, whose division generates an apoptotic daughter cell. The cyan circle indicates the ABalaappp cell, whose division generates two surviving daughters. The orientation of the embryo is shown in the middle. The cell lineage of ABalaapp is shown on the right. Each vertical line represents a cell and each branch represents a cell division. X indicates an apoptotic cell. Scale bar: 5 μm.

(B) Representative images of dividing ABalaappp cell (top panel) and ABalaappa cell (bottom panel) from the lineage tracing data. GFP fluorescence are shown as heatmaps below merged images. Cyan or bright green lines mark the chromatids or nuclei of daughter cells according to the H2B::mCherry signal. The black and white arrowheads point to the anterior and posterior daughter cells, respectively. Scale bar: 5 µm.

(C) Tree visualization of HDA-1::GFP and LIN-53::mNeonGreen segregation between sister cells during embryonic development. Cells and their lineage relations are visualized as a tree structure, with vertical lines indicating cells and horizontal lines indicating cell divisions. Green vertical lines show cells with a significantly higher (>1.5-fold) nuclear HDA-1::GFP and LIN-53::mNeonGreen intensity than their sister cells. Red arrowheads point to apoptotic cells. Placement on the tree is according to the Sulston nomenclature. See also Supplementary Table 2.

(D) Quantifications of HDA-1::GFP and LIN-53::mNeonGreen fluorescence intensity ratios between nuclei of the live daughter cell and its apoptotic sister cell automatically identified and tracked in 3D time-lapse imaging of embryonic development (Supplementary Table 2). The lineage names of non-apoptotic sister cells of 17 cell pairs with apoptotic cells are shown on the X-axis. The apoptotic sister cell of MSpppaap (magenta circle) completes apoptosis approximately 400 minutes after birth. N = 3 - 4 in each measurement. Data are mean ± SEM. One-sample t-test compared to 1, *p< 0.05, **p< 0.01, ***p< 0.001, ns: not significant.

Loss of the deacetylation activity of NuRD causes ectopic apoptosis

To investigate the role of NuRD in determining cell fate, we reduced the deacetylation activity of NuRD using RNA-mediated interference (RNAi). The hda-1 and lin-53 RNAi efficacy was confirmed by the reduced green fluorescence in the germlines of HDA-1::GFP and LIN-53::mNeonGreen KI animals, as well as the increased acetylation level of histone H3K27 (Figure S6). As reported before, RNAi of either had-1 or lin-53 caused embryonic lethality (Shi and Mello, 1998). RNAi of lin-53 led to embryos arrested at gastrulation stage before the generation of apoptotic cells, preventing us from analyzing its roles in somatic apoptosis. In contrast, hda-1 RNAi embryos arrested between the late gastrulation stage and bean stage, allowing us to tracked some of apoptotic events. To quantify apoptotic events, we used a secreted Annexin V (sAnxV::GFP) sensor to label apoptotic cells with externalized phosphatidylserine (exPS) on the surface of the plasma membrane in ced-1(e1735) mutant embryos which inhibits engulfment of apoptotic cells (Conradt et al., 2016; Mapes et al., 2012; Zhou et al., 2001). The sAnxV::GFP labeled an average of 10 cells in ced-1(e1735) mutant embryos treated with control RNAi and an average of 14 cells in ced-1(e1735); hda-1 RNAi embryos (Figure 3A-B). To test whether the upregulated apoptosis in hda-1 RNAi embryos depends on the canonic apoptosis pathway, we introduced cell death-deficient mutations ced-3/ Caspase (n2433), ced-4/ APAF-1-like (n1162), ced-9/ BCL-2-like (n1950), and egl-1/ BH-3 only protein (n1084n3082) into ced-1 mutant, respectively. RNAi of hda-1 in these embryos only generated 0-2 exPS-positive cells (Figure 3A-B and S7A), indicating that the ectopic increase of apoptosis by HDA-1 inhibition requires the EGL-1-CED-9-CED-4-CED-3 pathway.

RNAi of hda-1 induces ectopic apoptosis and increases H3K27 acetylation of the egl-1 gene.

(A) Fluorescent images show Phsp::sAnxV::GFP from ced-1(e1735) and ced-1(e1735); ced-3(n2433) embryos between late gastrulation and bean stage, treated with control RNAi or hda-1 RNAi. Scale bars, 5 μm.

(B) Quantifications of cell corpse number in the embryos (N = 39 - 60) of indicated genotypes. Data are presented as mean ± SEM. Unpaired two-tailed t-test, compared with the control RNAi, ***p < 0.001.

(C) Scatter plot of the increased gene expression in both hda-1 RNAi and lin-53 RNAi animals. The egl-1 gene is marked in red. See also Supplementary Table 3.

(D) Quantitative real-time PCR (RT-PCR) measurement of egl-1 mRNA levels in the control, hda-1 or lin-53 RNAi embryos. Fold induction was calculated relative to levels in control RNAi embryos. Data represent the mean ± SEM of six biological replicates. Unpaired t-test, ***p < 0.001.

(E) Quantification of the number of cells expressing the Pegl-1::NLS::GFP reporter in embryos treated with control, hda-1 or lin-53 RNAi. Data are mean ± SEM and are compared by unpaired t-test, *p < 0.05, **p < 0.01, ***p < 0.001. N = 8 – 9.

(F) Normalized ChIP-Seq signal profiles of the H3K27 acetylation level in the control, hda-1 or lin-53 RNAi embryos at the egl-1 locus. The Y-axis shows the average sequencing coverage of bins per million reads of three biological replicates normalized to the input control sample. Ectopic H3K27ac enrichments at the egl-1 locus of hda-1 RNAi and lin-53 RNAi embryos are highlighted in magenta and cyan.

(G) ChIP-qPCR analyses using H3K27ac antibody or IgG (the negative control) at selected elements of egl-1 indicated in (F) by red lines. Results are shown as the percentage of input DNA. Data are mean ± SEM from three biological replicates. Unpaired t-test, *p < 0.05.

NuRD RNAi upregulates the egl-1 expression by increasing its H3K27 acetylation

To understand how NuRD regulates apoptotic cell fates, we performed RNA-Seq analyses on WT, hda-1 RNAi and lin-53 RNAi embryos. In both RNAi conditions, we observed 287 or 575 genes that were upregulated or downregulated (|log2FD| > 2 and FDR < 0.05) (Supplementary Table 3). Notably, we found that the apoptosis-inducing gene egl-1 was among the upregulated genes (Figure 3C). Our quantitative RT-PCR results further confirmed that egl-1 transcripts were more abundant in hda-1 RNAi and lin-53 RNAi embryos than in control RNAi embryos (Figure 3D). We also used a Pegl-1::NLS::GFP transcriptional reporter to examine the egl-1 transcription activity during embryonic development, which allowed us to detect egl-1 expression while avoiding the adverse effects of egl-1 overexpression on apoptosis. We found that RNAi of hda-1 or lin-53 caused an increased number of Pegl-1::NLS::GFP positive cells (Figure 3E and S7B). These results indicate that the loss of NuRD abnormally activates the egl-1 gene expression.

Next, we investigated whether the loss of histone deacetylase activity in NuRD increased the H3K27ac level at the egl-1 locus, as the histone H3K27ac level is associated with transcription activation (Heinz et al., 2015). To test this, we performed ChIP-Seq experiments with the anti-H3K27ac antibody and found that the transcription start site (TSS) and the gene body region of egl-1 had higher levels of H3K27ac in hda-1 or lin-53 RNAi embryos compared to WT (Figure 3F). This was validated by ChIP-qPCR in hda-1 RNAi embryos (Figure 3G). Therefore, inhibiting NuRD increased the H3K27ac level at the egl-1 locus, leading to its upregulation and subsequent activation of apoptosis.

V-ATPase regulates asymmetric segregation of NuRD during somatic ACDs

To investigate how NuRD is asymmetrically segregated during ACD, we used affinity purification with an anti-GFP antibody and mass spectrometry (MS) to isolate NuRD binding partners from the lysate of HDA-1::GFP KI animals. Our analysis revealed all the previously known NuRD subunits from HDA-1::GFP KI animals (Figure S8, Supplementary Table 4) (Bracken et al., 2019; Lai and Wade, 2011), validating our experimental method. Interestingly, our co-immunoprecipitation and MS with HDA-1::GFP identified 12 subunits of the vacuolar-type H+-ATPase (V-ATPase) (Figure 4A-B), consistent with a previous report that used FLAG-tagged LIN-53 to identify 4 V-ATPase subunits (Muthel et al., 2019). We further confirmed the binding of HDA-1 to V-ATPase subunits in C. elegans by co-immunoprecipitation and western blot analysis with anti-V1A/VHA-13 antibodies (Figure 4C). V-ATPase is a proton pump that comprises a transmembrane domain (V0) responsible for proton transport across membranes and a peripheral ATP-hydrolytic domain (V1) that catalyzes ATP hydrolysis (Figure 4B). V-ATPases play a critical role in intracellular pH homeostasis and regulate numerous physiological processes, including apoptosis (Abbas et al., 2020; Chen et al., 2022; Ernstrom et al., 2012; Vasanthakumar and Rubinstein, 2020). However, the role of V-ATPase in the asymmetric segregation of epigenetic factors during ACD remains unknown.

HAD-1 interacts with subunits of V-ATPase.

(A) The plot compares counts of proteins co-precipitated with HDA-1::GFP with those with the control ACT-4 (actin)::GFP. The PSM (Peptide-Spectrum Match) is the number of identified peptide spectra matched for the protein. Blue dots represent the subunits of V-ATPase. See also Supplementary Table 4.

(B) Schematic model of the V-ATPase complex. The known worm subunits are indicated. The mean PSM of the encoded protein from two co-IP and MS repeats is shown in blue after the gene name.

(C) Western blot (WB) showing co-immunoprecipitation (co-IP) of V-ATPase V1 domain A subunit (V1A) with the HDA-1 from worm lysates.

To investigate the role of V-ATPase in NuRD asymmetric segregation in Q neuroblast, we used a pharmacologic inhibitor, bafilomycin A1 (Baf A1), to inhibit V-ATPase proton pumping activity (Furuchi et al., 1993; Wang et al., 2021; Yoshimori et al., 1991). To assess the inhibitory effect of Baf A1 on the proton translocation activity of V-ATPase in Q neuroblast, we monitored the cytosolic pH dynamics in dividing QR.a using the super-ecliptic pHluorin. The pHluorin is a pH-sensitive GFP reporter whose fluorescence can be quenched by the acidic pH (Miesenbock et al., 1998; Sankaranarayanan et al., 2000). In DMSO-treated animals, the pHlourin fluorescence intensity remained constant in the posterior portion that forms QR.ap but was significantly reduced in the anterior portion that forms QR.aa, indicating that the cytoplasm of the future apoptotic cell became more acidic (Figure 5A-B and Supplementary Movie S8). This observation is consistent with the recognition that cytosolic acidification is a common feature of both death receptor-mediated and mitochondria-dependent apoptosis (Lagadic-Gossmann et al., 2004; Matsuyama et al., 2000; Matsuyama and Reed, 2000). Notably, Baf A1 treatment reduced the pHluorin fluorescence intensity ratio between QR.ap and QR.aa from 1.6-fold to 1.3-fold (Figure 5A-B and Supplementary Movie S8), suggesting that Baf A1 may disrupt the cytosolic pH asymmetry in dividing QR.a cells by inhibiting V-ATPase activity.

V-ATPase regulates NuRD asymmetric segregation and cell fates.

(A) Dynamics of the cytosolic pH indicated by super-ecliptic pHluorin during QR.a division in DMSO control or Baf A1-treated animals. In each panel, the top row shows mCherry-tagged plasma membrane and histone, and the bottom row shows heatmaps of super-ecliptic pHluorin signal. The anterior of the cell is on the left. Time 0 min is the onset of anaphase: scale bar, 2 µm. See also Supplementary Movie S8.

(B) The super-ecliptic pHluorin fluorescence intensity ratio between QR.ap and QR.aa. Data are mean ± SEM. N =11-15. Unpaired two-tailed t-test, **p < 0.01, ns: not significant.

(C) Images of HDA-1::GFP distribution during QR.a cell division in DMSO control or Baf A1-treated animals. In each panel, the top row shows mCherry-tagged plasma membrane and histone, and the bottom row shows heatmaps of HDA-1::GFP fluorescence. The GFP fluorescence intensity ratios between the posterior and anterior chromatids or between QR.ap and QR.aa nuclei are shown in green at the lower left corner of heatmaps. Anterior of the cell is left. Scale bar: 2 µm. See also Supplementary Movie S9.

(D) Quantification of HDA-1::GFP fluorescence intensity ratios between the large and small daughter cell nuclei and the cell size ratios between the large and small daughters. The names of mother cells are shown on the X-axis. Data are mean ± SEM. N = 9-12. Unpaired two-tailed t-test, **p < 0.01, ***p < 0.001, ns: not significant.

(E) Representative images showing fates of QR.aa after DMSO (top), Baf A1 (middle) and Baf A1 plus AID treatment (bottom). The Q cell plasma membrane and chromosome are labeled by mCherry. HDA-1::GFP is shown as inverted images. Yellow arrow head shows a short neurite-like outgrowth of QR.aa in Baf A1 treated larvae. Frequencies of QR.aa survival and HDA-1 maintenance are showed on the right. Time 0 min is the birth of QR.aa. Scale bar: 5 µm.

We observed that treatment with Baf A1 resulted in the symmetric segregation of HDA-1 during QR.a division (Figure 5C-D and Supplementary Movie S9). Notably, neither DMSO nor Baf A1 treatment affected the asymmetry in daughter cell size (Figure 5C-D and Supplementary Movie S9), suggesting that intracellular acidification, regulated by V-ATPase activity, specifically affects NuRD asymmetric segregation without influencing daughter cell size asymmetry. We also investigated whether the small QR.aa cell carrying ectopic NuRD could escape apoptosis. In DMSO-treated animals, QR.aa cells underwent apoptosis, and 12 out of 13 QR.aa cells were engulfed and degraded by a neighboring epithelial cell within 120 min after birth. However, in Baf A1-treated animals, QR.aa inherited similar levels of HDA-1::GFP as its sister cell, and 11 out of 12 QR.aa cells carrying ectopic NuRD survived for over 120 min and formed a short neurite-like outgrowth (Figure 5E). To confirm whether the long-lived QR.aa was a consequence of ectopic gain of NuRD, we depleted HDA-1 within Q cell lineages using the auxin-inducible protein degradation (AID) system (Zhang et al., 2015). By introducing the degron sequence into the hda-1 locus and expressing the TIR1 F-box protein under the Q cell-specific egl-17 promoter, we showed that HDA-1::GFP fluorescence in Q cells was substantially decreased under auxin treatment (Figure 5E). Administration of auxin restored QR.aa apoptosis in Baf A1-treated animals (Figure 5E). These results suggest that V-ATPase activity-dependent NuRD asymmetric segregation contributes to the specification of the live-death fate.

To understand how V-ATPase regulates NuRD asymmetric segregation, we generated a transgenic strain expressing wrmScarlet-tagged VHA-17, which is the e subunit of the V0 domain. Using this strain, we were able to examine the dynamic distribution of V-ATPase during QR.a cell division. Like HDA-1, V-ATPase was also asymmetrically enriched in the surviving QR.ap portion (Figure 6A-B and Supplementary Movie S10). This observation suggests that V-ATPase and NuRD may co-segregate asymmetrically during ACD. We also found that Baf A1 treatment disrupted V-ATPase asymmetric distribution (Figure 6A-B and Supplementary Movie S10), indicating the importance of the proton-pumping activity of V-ATPase in its asymmetric segregation. Therefore, our results suggest that V-ATPase may facilitate the asymmetric distribution of NuRD through its proton pumping activity.

V-ATPase distribution during ACDs and a model.

(A) Dynamics of VHA-17::wrmScarlet during QR.a cell division in DMSO control or Baf A1-treated animals. VHA-17::wrmScarlet fluorescence is shown as heatmaps. Time 0 min is the onset of anaphase. Scale bar: 2 µm.

(B) Quantification of VHA-17::wrmScarlet (magenta) and HDA-1::GFP (blue) fluorescence intensity ratios between the posterior and anterior half of QR.a or between QR.ap and QR.aa. Data are mean ± SEM. N = 10 – 14. One-sample t-test compare to 1, ***p < 0.001, ns: not significant. See also Supplementary Movie S10.

(C) A proposed model. Asymmetric segregation of V-ATPase mediated the enrichment of its associated NuRD in the large daughter cell, where high level of NuRD deacetylates egl-1 and suppressed its expression.

Discussion

Our findings indicate that the asymmetric segregation of the NuRD complex during ACD is regulated in a V-ATPase-dependent manner, and is critical for the differential expression of apoptosis activator egl-1 and the life-versus-death fate decision (Figure 6C). Specifically, we have provided evidence that the reduced level of NuRD in the apoptotic daughter cell leads to an increased level of H3K27ac, an epigenetic modification known to be associated with active gene expression, at the egl-1 locus, resulting in the upregulation of egl-1 expression and subsequent induction of apoptosis. Based on our results, we propose a dual-function model of V-ATPase in NuRD asymmetric segregation: (i) V-ATPase interacts with NuRD, both of which are asymmetrically segregated into the surviving cell; (ii) the asymmetrical activity of V-ATPase generates a more acidic cytoplasmic environment in the future apoptotic cell portion compared to the surviving portion, which might contribute to the asymmetric segregation of V-ATPase and NuRD. This model sheds light on the mechanisms underlying the regulation of apoptosis during ACD, and opens up new avenues for further investigation into the intricate interplay between V-ATPase, NuRD, and epigenetic modifications in the context of cell fate decisions.

The intrinsic mechanisms governing binary cell fate decisions involve asymmetric cortical localization of cell fate determinants, polarized partitioning of RNA species, unequal segregation of organelles, and biased distribution of damaged proteins or protein aggregates (Sunchu and Cabernard, 2020; Venkei and Yamashita, 2018). Despite this, little is known about the asymmetric segregation of epigenetic modification enzymes during ACDs, nor the functions of polarized V-ATPase distribution and cytoplasmic proton asymmetry during this process. Our findings provide new insights into the asymmetric segregation of cell-intrinsic factors and the previously unrecognized roles of V-ATPase and cytosolic acidification in this process.

Two mechanisms have been proposed to explain the V-ATPase-dependent asymmetric segregation of NuRD. Firstly, a polarized intracellular transportation system could selectively deliver NuRD-V-ATPase-containing organelles to the surviving daughter cell. Although V-ATPase is primarily known for its localization and function in the late endosome and lysosome, recent evidence suggests that V-ATPase subunits are synthesized and assembled on the endoplasmic reticulum (ER) (Abbas et al., 2020; Graham et al., 2003; Wang et al., 2020). Our findings demonstrate that V-ATPase colocalizes with the ER marker during ACDs, suggesting that NuRD-V-ATPase may be delivered as cargo on the ER or ER-derived vesicles (Figure S9). As reported in previous studies, a polarized microtubule transportation system, governed by specific motor proteins and the microtubule tracks, is responsible for the asymmetric segregation of signaling endosomes during ACDs of Drosophila sensory organ precursors (Derivery et al., 2015). It is possible that a similar biased transportation system plays a role in the asymmetric segregation of NuRD-V-ATPase-containing organelles.

Secondly, the surviving daughter cell chromatids may have a greater ability to recruit NuRD than their apoptotic sister. This could be due to a higher number of yet unknown NuRD recruiting factors on chromatids or specific posttranslational modifications of chromosomal proteins that enhance the chromosomal recruitment of NuRD. Despite having an identical DNA sequence, sister chromatids can be distinguished by chromosomal proteins or modifications (Ranjan et al., 2019; Tran et al., 2012), which likely provides the specific molecular cues for asymmetric NuRD recruitment. The NuRD asymmetry on chromosomes (1.5∼1.8-fold between QR.ap and QR.aa nuclei) is greater than that between the two daughter cells (approximately 1.3-fold) (Figure 1D-1F), suggesting that polarized NuRD-V-ATPase transportation and distinct NuRD affinity from sister chromatids may act in concert to establish NuRD asymmetry.

The observation of asymmetric segregation of the NuRD complex during cell divisions suggests that the mother cell may play a role in initiating the fate specification of its daughter cells. This notion is supported by studies conducted on Drosophila male germline stem cells (GSCs) which demonstrated that the old and new histone H3-H4 or unequal amounts of histone H3 variant CENP-A were incorporated into sister chromatids prior to cell division (Ranjan et al., 2019; Tran et al., 2012). Although these patterns were not observed in cell types other than GSCs, asymmetric NuRD segregation was found to be common in C. elegans (Figure 1D-G, Figure 2, and Figure S3). Among the 395 cell lineages that were tracked in C. elegans embryos and generated two surviving daughter cells, NuRD was equally distributed to both daughters in 390 instances (Figure 2C-2D, Supplementary Table 2). However, in the remaining five cell divisions, both HDA-1 and LIN-53 exhibited >1.5-fold differences between the two surviving siblings (Figure 2C-2D, Supplementary Table 2). This observation suggests that biased NuRD segregation may contribute to binary cell fate decisions beyond apoptosis. In the embryonic cerebral cortex of mammals, approximately 30% of newborn cells undergo programmed cell death (Voiculescu et al., 2000), which is a significantly higher rate than observed in C. elegans where only 12% of somatically born cells undergo apoptosis (131 apoptotic cells among 1090 somatically born cells). Thus, it is worth exploring whether brain stem cell divisions employ asymmetric NuRD segregation to specify daughter cell fates. Further studies on polarity establishment hold promise in elucidating the mechanisms underlying the asymmetric inheritance of epigenetic information.

Materials and Methods Worm strains and culture

C. elegans were maintained as described in the standard protocol (Brenner, 1974). All strains were cultivated at 20 °C on nematode growth medium (NGM) agar plates seeded with Escherichia coli OP50 or HT115 (feeding RNAi assay). The wild-type strain was Bristol N2. Some strains were provided by the Caenorhabditis Genetics Center (CGC), funded by the NIH Office of Research Infrastructure Programs (P40 OD010440). The strains used in this study were listed in Extended Data Table 1.

Molecular biology

We performed genome editing experiments in C. elegans following the established methods (Dickinson et al., 2013; Friedland et al., 2013). We used the CRISPR design tool (https://zlab.bio/guide-design-resources) to select the target sites. The sgRNA sequences (Extended Data Table 2) were inserted into the pDD162 vector (Addgene #47549) by linearizing the vector with 15 base pairs (bp) overlapped primers using Phusion high-fidelity DNA polymerase (New England Biolabs, cat. no. MO531L). PCR products were digested using Dpn I (Takara, cat. no. 1235A) for 2 hours at 37°C and transformed into Trans5 α bacterial chemically competent cells (TransGen Biotech, Cat no. CD201-01). The linearized PCR products with 15 bp overlapping ends were cyclized to generate plasmids by spontaneous recombination in bacteria. Homology recombination (HR) templates were constructed by cloning the 1.5 kb upstream and downstream homology arms into the pPD95.77 plasmid using the In-Fusion Advantage PCR cloning kit (Clontech, Cat. #639621). Fluorescence tags were inserted into the constructs with a flexible linker before the stop codons. Synonymous mutations were introduced to Cas9 target sites to avoid the cleavage of the homologous repair template by Cas9. The plasmids were listed in Extended Data Table 3.

Constructs that express GFP-tagged NuRD components were generated using the PCR SOEing method(Hobert, 2002). N2 genomic sequences (1.5-2 kb promoter plus coding region) were linked with gfp::unc-54 3’ UTR DNA fragments. Constructs that express wrmScarlet-tagged V-ATPase components with gfp::unc-54 3’ UTR were generated using the In-Fusion Advantage PCR cloning kit (Clontech, Cat. #639621). The primers were listed in Extended Data Table 3.

Genome editing and transgenesis

To generate a knock-in strain, we purified the sgRNA construct and the repair template plasmids with the PureLink Quick PCR purification Kit (Invitrogen, #K310001) and co-injected them into N2 animals with the pRF4 [rol-6 (su1006)] and Podr-1::dsRed selection markers. The F1 transgenic progenies were singled and screened by PCR and Sanger sequencing. Transgenic C. elegans was generated by the germline microinjection of DNA plasmids or PCR products into N2 with Podr-1::gfp or Pegl-17::myri-mCherry and Pegl-17::mCherry::his-24 plasmids. We maintained at least two independent transgenic lines with a constant transmission rate (>50%). Concentrations of DNA constructs used for generating knock-in were 50 ng/µl, and for transgenesis were 20 ng/µl. All the strains, primers, plasmids, and PCR products were listed in Extended Data Table 1-3, respectively.

Live-Cell Imaging and quantification

Live imaging of C. elegans embryos, larvae, or young adults followed our established protocols (Chai et al., 2012; Zhang et al., 2017). L1 larvae or young-adult worms were anesthetized with 0.1 mmol/L levamisole in M9 buffer, mounted on 3% (wt/vol) agarose pads maintained at 20°C. Our imaging system includes an Axio Observer Z1 microscope (Carl Zeiss MicroImaging) equipped with a Zeiss 100×/1.46 numerical aperture (NA) objective, an Andor iXon+ EM-CCD camera, and 488 and 561-nm lines of a Sapphire CW CDRH USB Laser System attached to a spinning disk confocal scan head (Yokogawa CSU-X1 Spinning Disk Unit). Images were acquired by μManager (https://www.micro-manager.org). Time-lapse images of Q cell divisions were acquired with an exposure time of 200 msec every 1 min. To follow the cell fate of QR.aa in WT, Baf A1- or Baf A1 and IAA-treated worms, time-lapse images were acquired with 300 msec for 561 nm and 100 msec for 488 nm exposure time at 5min intervals. To image NuRD dynamics during oocyte meiosis, we used an exposure time of 300 msec every 30 sec. To image Phsp-16.41::sAnxV::GFP positive cells in WT, mutant, or RNAi animals, we incubated the eggs at 33 °C for 30 min to induce the reporter expression 3 hours post-egg-laying. After we kept the eggs at 20 °C for 2 hours, we imaged them with an exposure time of 200 msec. To acquire time-lapse images of Pegl-1::NLS::GFP positive cells during embryonic development, we dissected 2-4 cell-stage embryos from gravid adults and imaged them with an exposure time of 300 msec every 5 min. ImageJ software (http://rsbweb.nih.gov/ij/) was used to perform image analysis and measurement. Image stacks were z-projected using the average projection for quantification and using the maximum projection for display.

The quantifications of cellular fluorescence intensity ratios in Q cell lineages were described in Figure S4A. We used the mCherry-labeled plasma membrane to circumscribe Q cells (Region of Interest). To determine the ratios of fluorescence intensities in the posterior to anterior half (P/A) of Q.a lineages or A/P of Q.p lineages, the cell in the mean intensity projection was divided into posterior and anterior halves. ImageJ software was used to measure the fluorescence intensities of two halves with the background subtracted. The background-subtracted mean fluorescence intensities in the two halves were then normalized by the area and the two resulting values were divided. The same procedure was used to determine the ratios of fluorescence intensities between two daughter cells. Total fluorescence intensity was the sum of the posterior and anterior fluorescence intensities or the sum of fluorescence intensities from two daughter cells without area normalization. The ROIs used in the measurement and calculation of the nuclear intensity ratios were determined as regions with mCherry-tagged histone in each cell.

Feeding RNAi

RNAi bacteria were grown in LB media containing carbenicillin (50 µg/mL) and tetracycline hydrochloride (12.5 µg/mL) at 37°C overnight. RNAi bacteria were then seeded on NGM plates supplemented with 50 µg/mL carbenicillin and 1 mM Isopropyl β-D-thiogalactopyranoside (IPTG). We grew the seeded RNAi plates for 2 days at room temperature, allowing double-stranded RNA expression. To score phenotypes during embryogenesis or meiosis, we transferred the synchronized young adult worms to the culture plate containing the induced E. coli HT115 strains expressing the control (luciferase) or specific gene-targeted dsRNAs. We incubated them for 16-24 hours at 20°C and then scored the phenotypes. As for RT-qPCR and RNA-Seq experiments, parental animals of RNAi were used for laying eggs, and the progeny was collected 5 hours post-egg-laying. In ChIP-Seq experiments, we used worms of mixed stages under RNAi treatment for 24-36 hours. RNAi bacterial strains were from the Vidal library.

Immunoprecipitation and mass spectrometry

GFP transgenic or knock-in worms were raised on one hundred 90-mm NGM plates. Animals were collected and washed with M9 buffer three times. For each replicate, the lysate was made from 1∼2 ml packed worms with 3∼4 ml of 0.5-mm diameter glass beads using FastPrep-24 (MP Biomedicals) in lysis buffer [pH 7.4, 150 mM NaCl, 25 mM Tris-HCl, 10% glycerol, 1% NP-40, 1x cocktail of protease inhibitors from Roche (Complete, EDTA free), 40 mM NaF, 5 mM Na3VO4]. Worm lysates were then cleared by centrifugation at 14,000g for 30 min at 4 °C. For anti-GFP immunoprecipitation (IP), the supernatant was incubated with GFP-Trap A beads (Chromoteck, GTA20) for 1 hr. Beads were then washed three times with lysis buffer. Proteins were eluted with 300 μl 0.1 M glycine-HCl (pH 2.5) into 15 μl 1.5 M Tris-HCl (pH 8.8), samples were further processed for SDS–PAGE and western blot, or followed by precipitation by 100 μl trichloroacetic acid and re-dissolution in 60 μl 8 M Urea, 100 mM Tris-HCl (pH 8.5) for MS. MS Samples were then reduced in 5 mM TCEP, alkylated by 10 mM iodoacetamide, and diluted four-fold with 100 mM Tris-HCl (pH 8.5). The 0.2 μg trypsin was used to digest 10 μg protein in 1 mM CaCl2 and 20 mM methylamine overnight at 37°C. The resultant peptides were desalted by ZipTip pipette tips (Merck Millipore). For liquid chromatography-tandem mass spectrometry analysis, peptides were separated by an EASY-nLCII integrated nano-HPLC system (Proxeon, Odense, Denmark) directly interfaced to a Thermo Scientific Q Exactive mass spectrometer (Thermo Fisher Scientific). Peptides were loaded on an analytical fused-silica capillary column (150 mm in length, 75 mm in internal diameter; Upchurch) packed with 5 mm, 300 Å C-18 resin (Varian). Mobile phase A consisted of 0.1% formic acid, while mobile phase B consisted of 0.1% formic acid and 100% acetonitrile. The Q Exactive mass spectrometer was operated through Xcalibur 2.1.2 software in data-dependent acquisition mode. A single full-scan mass spectrum was used in the orbitrap (400–1,800 m/z in mass range, 60,000 in resolution) followed by ten data-dependent tandem mass spectrometry scans at 27% normalized collision energy (HCD). The tandem mass spectrometry spectra results were searched against the C. elegans proteome database by the Proteome Discoverer (version PD1.4; Thermo Fisher Scientific).

Generation of C. elegans embryonic cell suspensions for SPLiT-Seq

We prepared single cells from a C. elegans strain (smIs89) that expressed a Pegl-1::NLS::GFP reporter using the established protocols with minor modifications(Bianchi and Driscoll, 2006; Zhang et al., 2011). Synchronized gravid adults were transferred to NGM plates to lay eggs for 1 hour, and about 50 μl eggs pellet were collected 3 hours post laid. We washed eggs three times with 1 ml sterile egg buffer [118 mM NaCl, 48 mM KCl, 3 mM CaCl2, 3 mM MgCl2, 5 mM HEPES (pH 7.2)]. We rinsed the pellet with 250 μl 1 U/ml Chitinase three times. Eggshells were digested with 250 μl of 1 U/ml chitinase for 45 min on a rotter. We stopped the reaction by adding 750 µl of L-15/FBS (L-15 medium containing 10% fetal bovine serum). Eggs were pelleted by centrifugation at 900 g for 3 min. The supernatant was displaced with 500 µl L-15/FBS. Embryos were dissociated by gently pipetting up and down 60 times using an insulin syringe (29 G needle). The dissociated cells were separated from undissociated cell clumps by filtering through a 5 µm mesh filter, and we washed the filter with 500 µl L-15/FBS. Cells were collected by centrifuging 1 ml filtered suspension at 3220 g for 3 min at 4°C. We removed the supernatant and gently resuspended cells with 2 ml 1X PBS containing 5 μl SUPERaseIn RNase Inhibitor (Invitrogen) and 2.5 μl RiboLock RNase Inhibitor (Thermo Scientific).

SPLiT-Seq library preparation and sequencing

We performed SPLiT-Seq as previously described (Rosenberg et al., 2018) with minor modifications. After cell fixation, centrifugations were performed at 3220 g for 3 min according to the C. elegans cell size. The RT primers were poly-dT and random hexamer mixture with UMIs (Unique Molecular Identifiers). All the barcodes used in this study were described in the published protocol(Rosenberg et al., 2018). The first 24 barcodes were used in the first round of barcoding, named “Round1_01” to “Round1_24”, and 48 barcodes were used in the third round of barcoding, called “Round3_49” to “Round3_96”. Each cell barcode came from three rounds of barcoding. TruePrep DNA Library Prep Kit V2 constructed the sequencing library for Illumina (Vazyme Biotech #TD502) according to the manufacturer’s instructions. The library was sequenced on HiSeq systems (Illumina) using 150 nucleotides (nts) kits and paired-end sequencing.

SPLiT-Seq data processing

According to the results of FastQC, adaptors or low-quality nucleotides were trimmed by Trim Galore (v0.5.2) using the default parameters. For each paired-end sequencing read, a 10-bp UMI sequence and a 24-bp cell barcode were extracted from the Read 2 file by the tool “preprocess_splitseq.pl” of zUMIs (v0.0.6). The Read 1 was split by different cell barcodes in Read 2 and mapped to the modified C. elegans genome (WS263 with an artificial chromosome containing the C. elegans version of the GFP sequence) by zUMIs (v0.0.6) and STAR (v2.6.0c). Unique mapping reads were kept. The duplicated reads from the same transcript were excluded based on the UMI information in Read 2. Low-quality transcriptomes were removed from the analysis, yielding 442 single-cell transcriptomes, including transcript counts from 8629 genes. Among these transcriptomes, 41 transcriptomes were assigned to the putative apoptotic cell based on the expression of egl-1 or gfp. Among the total 8624 genes detected in SPLiT-seq, transcripts from 6487 genes were not detected in the putative apoptotic cell population. Enriched GO terms of those 6487 genes were identified using Gene Ontology knowledgebase (Ashburner et al., 2000; Gene Ontology, 2021) (http://geneontology.org/).

Cell lineage tracing and quantification of reporter expression

Embryo mounting, live imaging, lineage tracing, and expression quantification were performed according to a previously described procedure with minor modifications (Bao and Murray, 2011; Du et al., 2014; Murray et al., 2008). Two-to four-cell stage embryos were collected from young adult worms and mounted between two coverslips in the egg buffer containing 20-30 20-μm polystyrene microspheres and sealed with melted Vaseline (Bao and Murray, 2011). 3D time-lapse imaging was performed at 20 °C ambient temperature using a spinning disk confocal microscope (Revolution XD) for 300-time points at a 75-second interval. For each time point, embryos were scanned for 30 Z focal planes with 1 µm spacing. 3D tiff stack images were processed with the StarryNite software for automated cell identification and tracing to reconstruct embryonic cell lineages using the ubiquitously–expressed mCherry fluorescence. The raw results of cell identification and tracing were subjected to extensive manual inspection and editing using the AceTree program to ensure high accuracy (Du et al., 2014). For each mentioned strain, the cell lineage was traced from the two- or four-cell stage to the 350-cell stage for three or four embryos (experimental replicates). Regions of the chromatids or nuclei are determined by the H2B::mCherry signal. The fluorescent intensity of HDA-1::GFP or LIN-53::mNeonGreen in each traced nucleus at each time point was measured as the average fluorescent intensity of all the pixels within each nucleus and then with the average fluorescent intensity of the local background subtracted (Murray et al., 2008).

RNA sequencing

Synchronized young adult worms were cultured on RNAi plates for 24 hours. Gravid adults were transferred to new RNAi plates to lay eggs for 1 hour. Eggs were harvested 5 hours post-laid and then lysed with TRIzol reagent (Invitrogen). We extracted the total RNA following the manufacturer’s protocol. The Qubit RNA High Sensitivity Assay Kit (Invitrogen) was used to quantify RNA concentration. The Agilent 2100 bioanalyzer system was used for the assessment of RNA quality. We used samples with an RNA integrity number (RIN) above 6.0 for sequencing library construction. We used identical input total RNA (50 ng to 500 ng) between control and samples for library preparation using the KAPA RNA HyperPrep Kit (KAPA Biosystems, Wilmington, MA, USA). Libraries were analyzed by Agilent 2100 bioanalyzer system for quality control. The library samples were sequenced on an Illumina NovaSeq 6000 platform. Approximately 5 GB of raw data were generated from each sample with 150 bp paired-end read lengths.

RNA-Seq data analysis

FastQC assessed the quality score, adaptor content, and duplication rates of sequencing reads. Trim_galore (v0.6.0) was used to remove the low-quality bases and adaptor sequences with default parameters. After trimming, paired-end reads with at least 20 nucleotides in length were aligned to the C. elegans reference genome (WS263) using STAR (2.5.4b) with default parameters. Read counts representing gene expression levels were calculated with HTSeq (version 0.9.1) using default parameters. Uniquely mapped reads were included. Gene names were annotated on the Wormbase website (https://wormbase.org/tools/mine/simplemine.cgi). DESeq2 package in R programming language was used for differential expression analysis.

Differentially expressed genes were defined as upregulated genes (log2-transformed fold change greater than 1) or downregulated genes (log2-transformed fold change less than -1) with a false discovery rate (FDR) less than 0.05. We used the ggplot2 package of R to plot figures.

Reverse transcription and quantitative real-time PCR

Young adult worms were cultured on RNAi plates for 24 hours. We transferred the gravid adults to new RNAi plates to lay eggs for 1 hour. Eggs were harvested 5 hours post-laid and then lysed with TRIzol reagent (Invitrogen). The RNeasy Mini kit (QIAGEN) prepared the total RNA from ∼20 μl egg pellets for each biological replicate. cDNA from the same amount of RNA between RNAi and control samples were synthesized in a 20 μl reaction volume by the PrimeScript™ RT reagent Kit with gDNA Eraser (TAKARA, Code No. RR047A). One μl of cDNA was used as the template in a 10 μl reaction volume of the PowerUp SYBR Green Master Mix (Applied Biosystems) with four technical replicates. Quantitative real-time PCR was performed in two independent experiments with three biological replicates each time using the Applied Biosystems QuantStudio 1 Real-Time PCR System and normalized to the cell division cycle-related GTPase encoding gene cdc-42. Data were analyzed using the standard curve method. Primers used in real-time PCR assays were listed in Extended Data Table 3.

Chromatin immunoprecipitation

For ChIP assays, chromatin immunoprecipitation was performed as previously described(Mukhopadhyay et al., 2008) with modifications. Mixture stage worms on RNAi plates were harvested and washed three times with M9 buffer. 0.4 ml of packed worms were obtained and frozen into small balls with liquid nitrogen for each ChIP replicate. We quickly ground the little worm balls into powders in liquid nitrogen. Worm powder was crosslinked in 4 ml crosslinking buffer (1.1% formaldehyde in PBS with protease and phosphatase inhibitors) with constant rotation for 15 min at room temperature. Fixed worm samples were quenched by incubation with 0.125 M glycine for 5 min and then washed in cold PBS-PIC (PBS buffer with protease inhibitor cocktail) three times. The tissue pellet resuspended in PBS-PIC was homogenized by 2-3 strokes in a Dounce homogenizer. The cell suspension was centrifuged at 2,000 g for 5 min at 4°C, and we removed the supernatant.

We performed nuclei preparation, chromatin digestion, chromatin immunoprecipitation, and DNA Purification using SimpleChIP® Enzymatic Chromatin IP Kit (Magnetic Beads) (Cell Signaling Technology, #9003). Samples were sonicated using a Qsonica sonicator (Qsonica Q800R) at 70% amplitude for 5 cycles of 30 sec on and 30 sec off with 4 repeats. The chromatin was digested using 6 μl of micrococcal nuclease (MNase) (CST catalog number 10011S) in 300 μl of buffer B (CST catalog number 7007) containing 0.5 mM DTT for 8 min at 37 °C. The quality of chromatin digestion was analyzed by Agilent 2100 bioanalyzer system. Before IP, the chromatin sample concentration was assessed by the BCA Protein Assay Kit (Tiangen, cat no. PA115-02). 1 μg H3K27ac antibody (Abcam ab4729) or 2 μg IgG was used per IP with 0.5-1 mg input chromatin protein.

According to the manufacturer, we used 200 ng purified DNA samples to construct the ChIP–seq library using the VAHTS Universal DNA Library Prep Kit for Illumina V3 (Vazyme ND607). DNA libraries were sequenced on an Illumina NovaSeq 6000 platform, with 150-bp paired-end sequencing. Three biological replicates were collected for two independent ChIP–seq experiments. For ChIP–qPCR analysis of H3K27ac enrichment in the egl-1 locus, the same volume of cDNA from the IP experiment described above was used for qPCR analysis. Extended Data Table 3 listed the primers.

ChIP-Seq data processing

150-bp double-end reads were aligned to the C. elegans genome version WBcel235 using Bwa-mem2 version 2.1 with default parameters. Only uniquely mapped reads were kept. Samples were normalized by bins per million mapped reads (BPM) and averaged, and then treated samples were subtracted by input samples to calculate enrichment using deepTools2 (Ramirez et al., 2016). Data were visualized using IGV(Robinson et al., 2011).

Immunofluorescence

We performed immunofluorescence of C. elegans eggs using the freeze-cracking method. In brief, eggs were washed with M9 buffer three times and dripped onto poly-L-lysine coated slides. After freezing in liquid nitrogen for 5 minutes, the coverslip was swiftly removed, and embryos were fixed in methanol at -20 °C for 15 min. We washed the slides twice with PBST (PBS+ 0.05% Tween 20). We incubated slides with 1% BSA in PBST for 30 min at room temperature to block the unspecific bindings. The primary and secondary antibodies were diluted in PBST plus 1% BSA by 1:500-1:2,000 for primary antibodies (ab10799 for histone H3 and ab4729 for H3K27ac) and 1:5,000 for the secondary antibodies (Alexa-488 anti-mouse and Alexa546 anti-rabbit from Invitrogen). The primary antibody was incubated overnight at 4°C in a moisture chamber, and the secondary antibody was incubated for 1 hour at room temperature. After washing with PBST, we applied 1-2 drops of the Fluoroshield Mounting Medium containing DAPI (Abcam, ab104139) and coverslips to the samples.

Pharmacological and chemical treatments

To block V-ATPase proton pump activity in the L1-L2 larvae, which have cuticle layers that may compromise agent penetration, the Bafilomycin A1 treatment was designed based on the published protocols (Kumsta et al., 2017; Papandreou and Tavernarakis, 2017; Pivtoraiko et al., 2010). Briefly, gravid worms were placed in 1 drop of ddH2O containing 3% (v/v) pelleted OP50 and 100 μM bafilomycin A1 (Abcam, dissolved in DMSO) for 16 h at 20 ℃. After incubation, L1 larvae were washed with M9 buffer and recovered for 5 h on NG agar plates before imaging.

To achieve Q cell conditional HDA-1 depletion, the auxin-inducible degradation (AID) was performed as described previously (Zhang et al., 2015). The natural auxin (indole-3-acetic acid, IAA) was purchased from Alfa Aesar (#A10556). A 400 mM solution in ethanol was stored at 4°C as stock. Before imaging, L1 larvae were transferred to the S basal buffer supplemented with 3% (v/v) pelleted OP50 and 4 mM of auxin for a 1.5 h treatment.

Coimmunoprecipitation and Western blot

For immunoprecipitation of endogenous V-ATPase A subunit with HDA-1, HDA-1-GFP knock-in worms were lysed in ice-cold lysis buffer [pH 7.4, 150 mM NaCl, 25 mM Tris-HCl, 10% glycerol, 1% NP-40, 2x cocktail of protease inhibitors from Roche (Complete, EDTA free), 40 mM NaF, 5 mM Na3VO4]. The soluble fractions from worm lysates were immunoprecipitated with anti-GFP GFP-Trap A beads (Chromoteck, GTA20) for 1 hr at 4 °C. Immunoprecipitates were washed three times with lysis buffer containing 1x cocktail of protease inhibitors from Roche (Complete, EDTA free) and were then transferred to new tubes. Beads were washed again with 100 mM PB (pH 6.0, 6.5, 7.0, or 7.5) and boiled in 1X SDS loading buffer under 95 ℃ for 5min. The protein samples were analyzed by Western blotting.

Protein samples were resolved by SDS–PAGE and transferred to PVDF membranes. Each membrane was divided into two parts to be incubated with GFP (Abcam ab290, 1:4,000) or ATP6V1A (Abcam ab199326, 1:2,000) primary antibodies, followed by HRP-conjugated Goat Anti-Rabbit IgG (H&L) (Easybio, 1:5,000) secondary antibodies. Quantification of the endogenous V-ATPase A subunit interacting with HDA-1 was performed using HDA-1::GFP bands as a calibration standard. Quantitative densitometry of chemiluminescent bands was performed using ImageJ software.

Data availability

The data sets generated and analyzed in this study are available in the NCBI Gene Expression Omnibus (GEO, http://www.ncbi.nlm.nih.gov/geo) under accession number GSE167379 with the secure token uxwleayqnjkvnkf.

Supporting information

eLife Assessment

The authors propose that the asymmetric segregation of the NuRD complex in C. elegans is regulated in a V-ATPase-dependent manner, that this plays a crucial role in determining the differential expression of the apoptosis activator egl-1 and that it is therefore critical for the life/death fate decision in this species. The proposed model is interesting and the work could be important if proven correct. However, the current evidence is inadequate to support the major claims.

Joint Public Review:

Throughout the study, there is insufficient information about how experiments were performed and how often (imaging, pull-downs etc), how data was acquired, modified and analysed (especially imaging data, see below), how statistical analyses were done and what is presented in the figures (single planes or maximum intensity projections etc). This makes it difficult to evaluate the data and results.

There is insufficient information about tools and reporters used. This is misleading and impacts the conclusions that can be made from the results presented. To give an example, in Figure 1D-F, the authors present data that HDA-1::GFP and LIN-53::mNeonGreen (both components of the nucleosome remodeling and deacetylation complex) but not the histone acetyltransferase MYS-1::GFP are 'asymmetrically segregated' during QR.a division. However, the authors do not mention that HDA-1::GFP and LIN-53::mNeonGreen are expressed at endogenous levels (they are CRISPR alleles) whereas MYS-1::GFP is overexpressed (integration of a multi-copy extrachromosomal array). The difference in 'segregation' could therefore be a consequence of different levels of expression rather than different modes of segregation ('asymmetric' versus 'symmetric').

There is insufficient information about the phenotypes of the animals used (RNAi knock-downs of hda-1, lin-53 RNAi, pig-1 etc). Again this is misleading and impacts the conclusions that can be made. To give some examples, (1) in Figure 3A-G, control RNAi embryos are compared to hda-1 RNAi and lin-53 RNAi embryos. What the authors do not mention is that hda-1 RNAi and lin-53 RNAi embryos have severe developmental defects and essentially cannot be compared to control RNAi embryos. The differences between the embryos can be seen in Figure S7B where bright-field images of control RNAi, hda-1 RNAi and lin-53 RNAi embryos are shown. At the 350 min time point, a normal embryo is visible for the control, a 'ball of cells' embryo for hda-1 RNAi and an embryo that seems to have arrested at an earlier developmental stage (and therefore have much larger cells) for lin-53 RNAi. Because of these pleiotropic phenotypes, it is unclear whether differences seen for example in sAnxV::GFP positive cells (Figure 3A) are the result of a direct effect of hda-1(RNAi) on cell death or whether they are the result of global changes in development and cell fate induced by hda-1(RNAi). hda-1(RNAi) and lin-53(RNAi) embryos are also used for the data shown in Figures S6 and S7, raising the same concerns; (2) the authors do not mention what the impact of Baf A1 treatment is on animals; however, the images provided in Figure 5E indicate that Baf A1 treatment causes pleiotropic effects in L1 larvae.

There is a lack of adequate controls. Because of this, some of the data presented must be considered as preliminary. To give some examples: (1) controls are lacking for the data shown in Figure 3D-G (i.e. genes other than egl-1). Since hda-1 RNAi has a pleiotropic effect and most likely affects H3K27 acetylation genome-wide, this is critical. Based on what is shown, it is unclear whether the results presented are specific to egl-1 or not; (2) the co-IP and mass spec data shown in Figure 4A, C and Figure S8 also lack a critical control, which is GFP only. Because of this, it is unclear whether subunits of the V-ATPase bind to HDA-1 or GFP. The co-IP and mass spec data forms the basis of Figures 5 and 6 as well as Figure S9. Data presented in these figures therefore has to be considered preliminary as well.

Inappropriate methods are used. For this reason, some of the data again must be considered preliminary. To give some examples: (1) in Figure 5A, B, the authors used super-ecliptic pHluorin to look at changes in pH in the daughter cells. However, the authors used quenching of super-ecliptic pHluorin fluorescence rather than a ratio-metric method to 'measure' changes in pH. Because of this, it is unclear whether the changes in fluorescence observed are due to changes in pH or changes in the amount of pHluorin protein. Figure 5A, B forms the basis for the experiments presented in the remaining parts of Figure 5 as well as in Figure 6 and Figure S9; (2) the authors' description of how some images were modified before quantitative analysis raises concerns. The figures of concern are particularly Figure 1 and Figure S4, where background subtraction with denoising and deconvolution was used. Background subtraction, with denoising and deconvolution is an image manipulation that enhances the contrast between background and what looks like foreground. Therefore, background subtraction should be applied primarily in experiments involving image segmentation not fluorescence intensity measurement. Not being provided any information by the authors about the kind of subtraction that was made, this processing could lead to an uneven subtraction across the image, which can easily lead to artefacts. Since the fluorescence intensity in the smaller daughter cell is lower, and thus closer to background, the algorithm the authors used may have misinterpreted the grey value information in the smaller daughter cell pixels. This could have led to an asymmetric subtraction of background in the two daughter cells, leading to a stronger subtraction in the smaller daughter cell. Ultimately, their processing could have artificially increased the intensity asymmetry between the two daughter cells in all their results.

The imaging data is of low quality (for example Figures 1, 2, 5, 6; Figures S2, S3, S5, S6, S9). Since much of the study and the findings are based on imaging, this is a major concern. Critical parameters are not mentioned (number of sections in z-stack, size of the field-of-view, laser power used etc), which makes it difficult to understand what was done and what one is looking at. To give some specific examples, (1) the images shown in Figure 2B are of very low quality with severe background from neighbouring cells. In addition, the outline of the cells (plasma membrane) or the nuclei of the daughter cells is unknown. Based on this it is not clear how the authors could have measured 'Fluorescence intensity ratio between sister nuclei' in an accurate and unbiased way (what is clear from these images is that there is an increase in HDA-1::GFP signal in ALL surviving daughters (asymmetric and symmetric divisions) post cytokinesis but not in the daughter cell that is about to die (asymmetric and unequal division); (2) the images in Figure 6A and Figure S9A on VHA-17 segregation and its colocalization to ER and lysosome segregation during QR.a division are of very low quality and it is unclear to the reviewer how such images were used to obtain the quantitative data shown.

In some cases, there is a discrepancy between what is shown in figures and what the authors state in the text. To give some examples: (1) on page 7, the authors state "..., we found that nuclear HDA-1 or LIN-53 asymmetry gradually increased from 1.1-fold at the onset of anaphase to 1.5 or 1.8-fold at cytokinesis, respectively (Figure 1D-E)." Looking at the images for HDA-1 and LIN-53 in Figure 1D, the increase in the ratio mainly occurs between 4 min and 6 min, which is post cytokinesis and NOT prior to cytokinesis; (2) these images (Figure 1D) also show that there is an increase in the HDA-1 and LIN-53 signals in the larger daughter cells (QR.ap), which suggests that the increase in ratios (Figure 1E) is the result of increased HDA-1 and LIN-53 synthesis post cytokinesis. However, on top of page 8, the authors state "The total fluorescence of HDA-1, LIN-53 and MYS-1 remained constant during ACDs, suggesting that protein redistribution may establish NuRD asymmetry (Figure S4C)." In Figure S4C, the authors present straight lines for 'relative total fluorescence' for imaging (probably z-stacks) that was done every min over the course of 7 min. If there was no increase in material as the authors claim, they should have seen significant photobleaching over the course of the 7 min and therefore reduced level of 'relative total fluorescence' over time. How the data presented in Figure S4C was generated is therefore unclear. (Despite the fact that the authors claim that the asymmetry seen is not due to new synthesis in the larger daughter cell post cytokinesis, it would be more consistent with the first experiment presented in this study (Figure S1) that shows that there is more hda-1 mRNA in egl-1(-) cells compared to egl-1(+) cells); (3) On page 12, the authors state "..., in Baf A1-treated animals, QRaa inherited similar levels of HDA-1::GFP as its sister cell,...". However, looking at the image provided in Figure 5E (0 min), there seems to be a similar ratio of HDA-1::GFP between the daughter cells in DMSO and Baf A1-treated animals.

Author Response

The following is the authors’ response to the original reviews.

Public Reviews:

Throughout the study, there is insufficient information about how experiments were performed and how often (imaging, pull-downs etc), how data was acquired, modified and analysed (especially imaging data, see below), how statistical analyses were done and what is presented in the figures (single planes or maximum intensity projections etc). This makes it difficult to evaluate the data and results.

We have incorporated additional experimental details to the Materials and Methods section: "Recent advancements in optical and camera technologies permit the acquisition of Z-stacks without perturbing Q cell division or overall animal development. Z-stack images were acquired over a range of -1.6 to +1.6 μm from the focal plane, at intervals of 0.8 μm. The field-of-view spanned 160 μm × 160 μm, and the laser power, as measured at the optical fiber, was approximately 1 mW. ImageJ software (http://rsbweb.nih.gov/ij/) was used to perform image analysis and measurement. Image stacks were z-projected using the average projection for quantification and using the maximum projection for visual display. "

The majority of our experimental procedures adhere to methodologies delineated in our prior publications and other scientific literature. We were pioneers in the development of fluorescence time-lapse live microscopy techniques for capturing Q cell migration and asymmetric division (Ou and Vale, Journal of Cell Biology, 2009; Ou et al., Science, 2010; Chai et al., Nature Protocols, 2012). Our innovative imaging protocol uncovered a novel mode of polarized, non-muscle myosin-II-dependent asymmetric cell division (Ou et al., Science, 2010). Subsequently, we unveiled another previously uncharacterized mechanism of asymmetric cell division dependent on polarized actin polymerization (Chai et al., Cell Discovery, 2022). In the present study, we have significantly refined our imaging and quantification protocols. Different from the single-focal-plane imaging employed in our earlier study by Ou et al. 2009, advancements in optical technologies and camera resolution now enable us to undertake time-lapse imaging across multiple focal planes and track signal differences between the anterior and posterior segments of dividing cells.

There is insufficient information about tools and reporters used. This is misleading and impacts the conclusions that can be made from the results presented. To give an example, in Figure 1D-F, the authors present data that HDA-1::GFP and LIN-53::mNeonGreen (both components of the nucleosome remodeling and deacetylation complex) but not the histone acetyltransferase MYS-1::GFP are 'asymmetrically segregated' during QR.a division. However, the authors do not mention that HDA-1::GFP and LIN-53::mNeonGreen are expressed at endogenous levels (they are CRISPR alleles) whereas MYS-1::GFP is overexpressed (integration of a multi-copy extrachromosomal array). The difference in 'segregation' could therefore be a consequence of different levels of expression rather than different modes of segregation ('asymmetric' versus 'symmetric').

Figure S2 shows overexpressed HDA-1, LIN-53 and CHD-3 are also asymmetrically segregated during ACD of QR.a, which indicates that different levels of expression do not affect the modes of segregation, at least for the NuRD subunits. In the main text, however, we presented the asymmetric segregation of HDA-1::GFP and LIN-53::mNeonGreen using their CRISPR KI alleles.

There is insufficient information about the phenotypes of the animals used (RNAi knock-downs of hda-1, lin-53 RNAi, pig-1 etc). Again this is misleading and impacts the conclusions that can be made. To give some examples,

1. In Figure 3A-G, control RNAi embryos are compared to hda-1 RNAi and lin-53 RNAi embryos. What the authors do not mention is that hda-1 RNAi and lin-53 RNAi embryos have severe developmental defects and essentially cannot be compared to control RNAi embryos. The differences between the embryos can be seen in Figure S7B where bright-field images of control RNAi, hda-1 RNAi and lin-53 RNAi embryos are shown. At the 350 min time point, a normal embryo is visible for the control, a 'ball of cells' embryo for hda-1 RNAi and an embryo that seems to have arrested at an earlier developmental stage (and therefore have much larger cells) for lin-53 RNAi. Because of these pleiotropic phenotypes, it is unclear whether differences seen for example in sAnxV::GFP positive cells (Figure 3A) are the result of a direct effect of hda-1(RNAi) on cell death or whether they are the result of global changes in development and cell fate induced by hda-1(RNAi). hda-1(RNAi) and lin-53(RNAi) embryos are also used for the data shown in Figures S6 and S7, raising the same concerns;

In the submitted manuscript, we mentioned that hda-1 RNAi and lin-53 RNAi caused embryonic lethality and that we could track some of the apoptotic events in hda-1 RNAi embryos arrested between the late gastrulation stage and bean stage. We agree with the reviewers that because of the pleiotropic phenotypes, we cannot distinguish whether sAnxV::GFP positive cells (Figure 3A) are the result of a direct effect of hda-1 (RNAi) on cell death or whether they are the result of global changes in development and cell fate induced by hda-1 (RNAi). We added the sentence to page 9 line 26: “Considering the pleiotropic phenotypes caused by loss of HDA-1, we cannot exclude the possibility that ectopic cell death might result from global changes in development, even though HDA-1 may directly contribute to the life-versus-death fate determination.”

1. The authors do not mention what the impact of Baf A1 treatment is on animals; however, the images provided in Figure 5E indicate that Baf A1 treatment causes pleiotropic effects in L1 larvae.

We have carefully checked the BafA1 treated animals, but have not been able to detect any visible defect in Baf A1 treated animals under a 25× dissection microscope at the given dosage and duration of treatment. We also searched for the published images or literature and did not find pleiotropic effects on the animal level at this dosage and duration; however, we agree with the reviewers that perturbation of pH homeostasis in lysosomes by BafA1 will certainly generate pleiotropic cellular defects. We discussed the issue below:

"Although BafA1-mediated disruption of lysosomal pH homeostasis is recognized to elicit a wide array of intracellular abnormalities, we found no evidence of such pleiotropic effects at the organismal level with the dosage and duration of treatment employed in this study."

There is a lack of adequate controls. Because of this, some of the data presented must be considered as preliminary. To give some examples:

1. Controls are lacking for the data shown in Figure 3D-G (i.e. genes other than egl-1). Since hda-1 RNAi has a pleiotropic effect and most likely affects H3K27 acetylation genome-wide, this is critical. Based on what is shown, it is unclear whether the results presented are specific to egl-1 or not;

In figure 3F, we added F23B12.1 and sru-43 as the controls of egl-1. We added “while the H3K27ac level of genes adjacent to egl-1 showed no significant changes” to Page 10 line 22 in the revised text.

1. The co-IP and mass spec data shown in Figure 4A, C and Figure S8 also lack a critical control, which is GFP only. Because of this, it is unclear whether subunits of the V-ATPase bind to HDA-1 or GFP. The co-IP and mass spec data forms the basis of Figures 5 and 6 as well as Figure S9. Data presented in these figures therefore has to be considered preliminary as well.

In the co-IP and mass spec shown in Figure 4A, we used ACT-4::GFP as the negative control, which can preclude V-ATPase subunits that bind to GFP. In Figure 4C, we used anti-V1A (V-ATPase V1 domain A subunit) antibody to confirm the interaction between V1A and HDA-1. In Figure S8B, we also used ACT-4::GFP as a control, showing other NuRD subunits bind to HDA-1 rather than GFP.

Inappropriate methods are used. For this reason, some of the data again must be considered preliminary. To give some examples:

1. In Figure 5A, B, the authors used super-ecliptic pHluorin to look at changes in pH in the daughter cells. However, the authors used quenching of super-ecliptic pHluorin fluorescence rather than a ratio-metric method to 'measure' changes in pH. Because of this, it is unclear whether the changes in fluorescence observed are due to changes in pH or changes in the amount of pHluorin protein. Figure 5A, B forms the basis for the experiments presented in the remaining parts of Figure 5 as well as in Figure 6 and Figure S9;

Bafilomycin A1 inhibits the activity of V-ATPase, presumably preventing the pumping of protons into the apoptotic daughter cell. It is more likely that the apoptotic daughter cell becomes less acidic and more neutral after the treatment of Baf1A, although we cannot exclude the possibility that the changes in fluorescence could be due to changes in the amount of pHluorin protein. A ratio-metric method to measure changes in pH will be further used to distinguish the two possibilities.

We added “although we cannot exclude the possibility that the changes in fluorescence could be due to changes in the amount of pHluorin protein.” to Page 12 line 12 in the revised text.

1. The authors' description of how some images were modified before quantitative analysis raises concerns. The figures of concern are particularly Figure 1 and Figure S4, where background subtraction with denoising and deconvolution was used. Background subtraction, with denoising and deconvolution is an image manipulation that enhances the contrast between background and what looks like foreground. Therefore, background subtraction should be applied primarily in experiments involving image segmentation not fluorescence intensity measurement. Not being provided any information by the authors about the kind of subtraction that was made, this processing could lead to an uneven subtraction across the image, which can easily lead to artefacts. Since the fluorescence intensity in the smaller daughter cell is lower, and thus closer to background, the algorithm the authors used may have misinterpreted the grey value information in the smaller daughter cell pixels. This could have led to an asymmetric subtraction of background in the two daughter cells, leading to a stronger subtraction in the smaller daughter cell. Ultimately, their processing could have artificially increased the intensity asymmetry between the two daughter cells in all their results.

As mentioned earlier, the imaging and quantification methods of this manuscript have been routinely used in our previous publications or studies from many other labs (Gräbnitz F, et al., Cell Rep. 2023; Herrero E, et al., Genetics. 2020; Roubinet C, et al., Curr Biol. 2021). Background subtraction is a standard procedure to quantify cellular fluorescence intensities. The fluorescence intensity of the slide background was measured from a region without worm bodies, of the same size as the region of interest. We have added how we measured the background to page 19 Line 24: “The fluorescence intensity of the slide background was measured from a region without worm bodies, of the same size as the region of interest.”

The imaging data is of low quality (for example Figures 1, 2, 5, 6; Figures S2, S3, S5, S6, S9). Since much of the study and the findings are based on imaging, this is a major concern. Critical parameters are not mentioned (number of sections in z-stack, size of the field-of-view, laser power used etc), which makes it difficult to understand what was done and what one is looking at.

Fluorescence images of neuroblast asymmetric cell division in developing C. elegans larvae has historically presented considerable challenges. Our recent methodological advancements have facilitated live imaging in this intricate system with improved resolution. In the revised manuscript, we have elucidated the specific z-stack parameters, field-of-view dimensions, and laser power settings employed: "Z-stack images were acquired over a range of -1.6 to +1.6 μm from the focal plane, at intervals of 0.8 μm. The field-of-view spaned 160 μm × 160 μm, and the laser power, as measured at the optical fiber, was approximately 1 mW."

To give some specific examples,

1. The images shown in Figure 2B are of very low quality with severe background from neighbouring cells. In addition, the outline of the cells (plasma membrane) or the nuclei of the daughter cells is unknown. Based on this it is not clear how the authors could have measured 'Fluorescence intensity ratio between sister nuclei' in an accurate and unbiased way (what is clear from these images is that there is an increase in HDA-1::GFP signal in ALL surviving daughters (asymmetric and symmetric divisions) post cytokinesis but not in the daughter cell that is about to die (asymmetric and unequal division));

We employed live-cell imaging in conjunction with automated cell lineage tracing algorithms (Du et al., Cell, 2014) to scrutinize NuRD asymmetry in embryos from the two- or four-cell stage up to the 350-cell stage. This sophisticated approach was initially pioneered by Dr. Zhirong Bao at Sloan Kettering and subsequently refined by Dr. Zhuo Du during Dr. Du's postdoctoral training in Dr. Bao's laboratory. This advanced imaging pipeline enables the scientific community to quantify cellular fluorescence intensity in an automated fashion, thereby substantially mitigating manual intervention and bias.

1. The images in Figure 6A and Figure S9A on VHA-17 segregation and its colocalization to ER and lysosome segregation during QR.a division are of very low quality and it is unclear to the reviewer how such images were used to obtain the quantitative data shown.

In some cases, there is a discrepancy between what is shown in figures and what the authors state in the text. To give some examples:

1. On page 7, the authors state "..., we found that nuclear HDA-1 or LIN-53 asymmetry gradually increased from 1.1-fold at the onset of anaphase to 1.5 or 1.8-fold at cytokinesis, respectively (Figure 1D-E)." Looking at the images for HDA-1 and LIN-53 in Figure 1D, the increase in the ratio mainly occurs between 4 min and 6 min, which is post cytokinesis and NOT prior to cytokinesis;

Thank the reviewer for pointing out this. The nuclear HDA-1 or LIN-53 asymmetry increased to 1.5 or 1.8-fold 6 min after the onset of anaphase, when QR.a just completes cytokinesis. Therefore, We change the sentence “we found that nuclear HDA-1 or LIN-53 asymmetry gradually increased from 1.1-fold at the onset of anaphase to 1.5 or 1.8-fold at cytokinesis, respectively (Figure 1D-E).” to “we found that nuclear HDA-1 or LIN-53 asymmetry gradually increased from 1.1-fold at the onset of anaphase to 1.5 or 1.8-fold upon the completion of cytokinesis, respectively (Figure 1D-E).”

However, nuclear HDA-1 or LIN-53 asymmetry initiates prior to cytokinesis. We started to see the nuclear HDA-1 or LIN-53 asymmetry (1.4 fold for HDA-1 and 1.2 fold for LIN-53 ) 2 min after the onset of anaphase (Figure 1D).

1. These images (Figure 1D) also show that there is an increase in the HDA-1 and LIN-53 signals in the larger daughter cells (QR.ap), which suggests that the increase in ratios (Figure 1E) is the result of increased HDA-1 and LIN-53 synthesis post cytokinesis. However, on top of page 8, the authors state "The total fluorescence of HDA-1, LIN-53 and MYS-1 remained constant during ACDs, suggesting that protein redistribution may establish NuRD asymmetry (Figure S4C)." In Figure S4C, the authors present straight lines for 'relative total fluorescence' for imaging (probably z-stacks) that was done every min over the course of 7 min. If there was no increase in material as the authors claim, they should have seen significant photobleaching over the course of the 7 min and therefore reduced level of 'relative total fluorescence' over time. How the data presented in Figure S4C was generated is therefore unclear. (Despite the fact that the authors claim that the asymmetry seen is not due to new synthesis in the larger daughter cell post cytokinesis, it would be more consistent with the first experiment presented in this study (Figure S1) that shows that there is more hda-1 mRNA in egl-1(-) cells compared to egl-1(+) cells);

Regarding the concern of photo-bleaching, we have meticulously calibrated our imaging system over the past several years. Rigorous controls, qualification, and analyses were scrupulously undertaken during the development of our fluorescence time-lapse imaging system for the investigation of Q cell dynamics, initially established by Dr. Guangshuo Ou in Ron Vale's laboratory—a renowned hub for avant-garde imaging techniques (Ou & Vale, Journal of Cell Biology, 2009; Ou et al., Science, 2010). Remarkably, no discernible photobleaching was observed even during two to three-hour imaging.

We agree that protein turnover, involving both degradation and synthesis, may occur. However, NuRD asymmetric distribution occurred within several minutes after metaphase and QR.a completes cytokinesis ~6min after the onset of anaphase, while GFP protein translation and maturation require ~ 30 min in Q neuroblast (Ou & Vale, Journal of Cell Biology, 2009). Even if hda-1::gfp mRNA is translated during cell division, the nascent GFP-tagged protein will mature long after the completion of cytokinesis. Consequently, we postulate that the influence of newly synthesized GFP-tagged protein during Q cell division is negligible for quantification purposes. It is plausible that the asymmetry in HAD-1 protein distribution is independent of hda-1 mRNA asymmetry.

1. On page 12, the authors state "..., in Baf A1-treated animals, QRaa inherited similar levels of HDA-1::GFP as its sister cell,...". However, looking at the image provided in Figure 5E (0 min), there seems to be a similar ratio of HDA-1::GFP between the daughter cells in DMSO and Baf A1-treated animals.

We have adjusted the images in Figure 5E to show the asymmetry in DMSO-treated control animals. We acknowledge variations among animals. Our quantifications from more than 10 animals show the HDA-1 asymmetry in DMSO-treated animals in Figure 5B.

Recommendations for the authors:

Conclusion 1

"Here, we demonstrate that the nucleosome remodeling and deacetylase (NuRD) complex is asymmetrically segregated into the surviving daughter cell rather than the apoptotic one during ACDs in Caenorhabditis elegans" (Abstract)

Results described on pages 6-9 ("NuRD asymmetric segregation during neuroblast ACDs" and "NuRD asymmetric segregation in embryonic cell lineages") and data shown in Figure S1, Figure 1, Figures S2, S3, S4, S5, Figure 2.

Conclusion 1 is not supported by the results as numerous concerns exist about the data in many of these figures (see above, major weaknesses). A more likely explanation for the authors' observations is that there is synthesis of NuRD post cytokinesis and that asymmetries in the amounts of NuRD observed in the two daughter cells is a consequence of their different cell sizes (QR.ap is 3x as large as QR.aa). This is supported by the finding that the loss of pig-1, which causes 'equal' division resulting in two daughter cells of similar sizes, abolishes the differences in NuRD seen between the daughter cells.

As discussed earlier, GFP protein translation and maturation require ~ 30 min in Q neuroblast (Ou & Vale, Journal of Cell Biology, 2009). Even if there is the synthesis of NuRD post cytokinesis, the nascent GFP-tagged protein will not mature within our imaging timeframe, Therefore, NuRD asymmetry is unlikely to be a result of the synthesis of NuRD post cytokinesis. In addition, We found that MYS-1::GFP was symmetrically segregated into the small apoptotic daughter cells and big surviving daughter cells, suggesting NuRD asymmetry may be irrelevant to cell size asymmetry.

Interestingly, despite the fact that the loss of pig-1 causes 100% of the divisions to be equal by size and symmetric with respect to NuRD amounts, it only causes about 30% of QR.aa cells to inappropriately survive. This demonstrates that there is a correlation between NuRD asymmetry and daughter cell size asymmetry but NOT between NuRD asymmetry and cell death. This also demonstrates that loss of 'NuRD asymmetry' and presence of NuRD in the daughter that should die is NOT sufficient to block its death.

Cordes et al. 2006 (DOI: 10.1242/dev.02447) reported that in pig-1 loss-of-function mutants, <40% of Q.p lineages produce extra neurons because Q.pp cells inappropriately survive. Noticeably, only 30% and 5% Q.p lineages produce extra neurons in ced-3 and egl-1 loss of function single mutant, respectively. pig-1 ced-3 double mutant or pig-1 egl-1 double mutants show a dramatically stronger phenotype than either single mutant, resulting in about 80% of Q.p lineages producing extra neurons. These results suggest that pig-1 functions in parallel to the EGL-1-CED-9-CED-4-CED-3 cell death pathway to promote Q cell apoptosis.

We agree with the reviewer that “loss of 'NuRD asymmetry' and presence of NuRD in the daughter that should die is NOT sufficient to block its death” in pig-1 loss-of-function mutants. However, these results do not rule out the correlation between NuRD asymmetry and cell death. In the pig-1 mutant, the concentration of NuRD in Q.pp might not be high enough to completely block the death pathway. Alternatively, NuRD may be one but not the only factor blocking the cell death pathway.

Lastly, it is imperative to underscore that cellular aberrations observed during early developmental stages frequently undergo compensatory correction during subsequent developmental stages or even initial aging stages. For example, in certain cell migration mutants exhibiting early migration defects, the initial penetrance exceeds 80%; however, the penetrance is mitigated to a mere 30% in adults. Such observations have been corroborated in our prior publications focusing on cell migration dynamics (Wang et al., PNAS, 2013; Zhu et al., Dev Cell, 2016). This appears to be a pervasive phenomenon, echoed by several laboratories specializing in neural development. Sengupta and Blacque’s labs has reported that early aging can ameliorate ciliary phenotypes in C. elegans mutants with compromised intraflagellar transport mechanisms. Accordingly, late developmental stages may act as a compensatory buffer for antecedent developmental abnormalities.

Conclusion 2

"The absence of NuRD triggers apoptosis via the EGL-1-CED-9-CED-4-CED-3 pathway, while an ectopic gain of NuRD enables apoptotic cells to survive." (Abstract) Results described on pages 8-10 ("Loss of the deacetylation activity of NuRD causes ectopic apoptosis" and "NuRD RNAi upregulates the egl-1 expression by increasing its H3K27 aceylation") and data shown in Figure S6, Figure 3, Figure S7 and data shown in Figure 5.

Because of the various concerns raised above (major weaknesses) about the data presented in Figure S6, Figure 3, Figure S7 (pleiotropic phenotypes of hda-1 and lin-53 RNAi animals, lack of controls etc), there is no evidence that NuRD has a specific and/or direct effect on egl-1 expression in cells programmed to die or that loss of NuRD causes ectopic egl-1-dependent cell death. The claim that "ectopic gain of NuRD enables apoptotic cells to survive." is based on Figure 5E, where the authors show that Baf A1 treatment causes symmetric NuRD segregation in 11/12 animals and that QR.aa survives in 11/12 animals. However, those data are unconvincing. As mentioned above (major weaknesses), from the low-quality images provided, it is not clear whether there is 'symmetric NuRD segregation' in Baf A1 treated animals, and the conditions of the animals are a concern because of pleiotropic effects of blocking V-ATPase. (I am not convinced I am actually looking at the same region of an L1 larvae in the three animals because the HDA-1::GFP signal seems inconsistent across them.) One process that is affected by a block of V-ATPase is engulfment. The fact that the authors observe that 130 min post-cytokinesis, QR.aa still persists in Baf A1 treated animals could therefore be the result of a delay in engulfment rather than a block in cell death. In addition, the claim that ectopic gain of NuRD enables apoptotic cells to survive contradicts their findings on loss of pig-1 described about ('Conclusion 1').

We acknowledge the limitations of our imaging system; however, as we pointed out earlier that we developed imaging methods and kept improving them. We have tried our best to obtain images from developing C. elegans larvae. On the other hand, we showed that hda-1 RNAi and lin-53 RNAi increase the expression of a subset of genes, including egl-1, either directly or indirectly (Fig. 3C). Figure 3B shows the ectopic cell death caused by loss of NuRD is dependent on EGL-1-CED-9-CED-4-CED-3 pathway. While we cannot exclude several other possibilities while addressing such a complex problem in such a challenging model system, these results provide some evidence supporting that our claim can be one of the possibilities.

Conclusion(s) 3

"We identified the vacuolar H+-adenosine triphosphatase (V-ATPase) complex as a crucial regulator of NuRD's asymmetric segregation. V-ATPase interacts with NuRD and is asymmetrically segregated into the surviving daughter cell. Inhibition of V-ATPase disrupts cytosolic pH asymmetry and NuRD asymmetry" (Abstract)

Results described on pages 10-13 ("V-ATPase regulates asymmetric segregation of NuRD during somatic ACDs") and data shown in Figures 4, 5, 6, Figures S8, S9.

As outlined above (major weaknesses), the evidence that HDA-1 interacts with the V-ATPase complex is preliminary (no GFP control), and I consider the imaging data showing that V-ATPase asymmetrically segregates very low quality and unconvincing (Figure 6). The data on pH changes are also preliminary as the experiment was not done the way it should have (quenching rather than ratiometric). Finally, there are concerns about the results that apparently demonstrate that inhibiting V-ATPase activity disrupts pH asymmetry and NuRD asymmetry (impact of Baf A1 treatment).

As discussed earlier, Bafilomycin A1 inhibits the activity of V-ATPase, presumably preventing the pumping of protons into apoptotic daughter cells. It is more likely that the apoptotic daughter cell becomes less acidic and more neutral after the treatment of Baf1A, although we cannot exclude the possibility that the changes in fluorescence could be due to changes in the amount of pHluorin protein. A ratio-metric method to measure changes in pH will be further used to distinguish the two possibilities.

We added “although we cannot exclude the possibility that the changes in fluorescence could be due to changes in the amount of pHluorin protein.” to Page 12 line 12 in the revised text.

Conclusion 4

"We suggest that asymmetric segregation of V-ATPase may cause distinct acidification levels in the two daughter cells, enabling asymmetric epigenetic inheritance that specifies their respective life-versus-death fates." (Abstract) Discussion and model Figure 6C.

I consider the model premature and not based on any convincing data. In addition, the role of V-ATPase and acidification does not make sense. V-ATPase is involved in the acidification of the lysosomal system (lumen), and it is thought that cytosolic acidification in apoptotic cells is caused by lysosomal leakage. This is not consistent with the authors' model.

This manuscript lacks a section describing details of statistical analyses and the rationale for the chosen test, sample sizes, exclusion criteria, and replication details. Although the sample size is relatively smaller (less than 30), the authors used "unpaired t-test" for most of the tests. They should describe which type of t-test they used (parametric or non-parametric test). They also should provide replication details for non-statistical data set, for example Fig 3F and Fig 4C.

We used the Unpaired two-tailed parametric t-test. We have now added the information for statistic tests in the revised methods and figure legends.

Version 2

Abstract

Abstract

Asymmetric cell divisions (ACDs) generate two daughter cells with identical genetic information but distinct cell fates through epigenetic mechanisms. However, the process of partitioning different epigenetic information into daughter cells remains unclear. Here, we demonstrate that the nucleosome remodeling and deacetylase (NuRD) complex is asymmetrically segregated into the surviving daughter cell rather than the apoptotic one during ACDs in Caenorhabditis elegans. The absence of NuRD triggers apoptosis via the EGL-1-CED-9-CED-4-CED-3 pathway, while an ectopic gain of NuRD enables apoptotic daughter cells to survive. We identify the vacuolar H+–adenosine triphosphatase (V-ATPase) complex as a crucial regulator of NuRD’s asymmetric segregation. V-ATPase interacts with NuRD and is asymmetrically segregated into the surviving daughter cell. Inhibition of V-ATPase disrupts cytosolic pH asymmetry and NuRD asymmetry. We suggest that asymmetric segregation of V-ATPase may cause distinct acidification levels in the two daughter cells, enabling asymmetric epigenetic inheritance that specifies their respective life-versus-death fates.

Introduction

Asymmetric cell division (ACD) gives rise to two daughter cells that possess identical genetic material but distinct cell fates, playing a crucial role in both development and tissue homeostasis (Royall and Jessberger, 2021; Sunchu and Cabernard, 2020; Venkei and Yamashita, 2018; Wooten et al., 2020; Zion et al., 2020). Both extrinsic and intrinsic mechanisms determine distinct daughter cell fates after ACD. While extrinsic mechanisms, such as exposure to signaling cues from the local niche, have been extensively studied and are known to define stem cell fate (Morrison and Spradling, 2008), the intrinsic mechanisms are more complex and largely unresolved, despite several proteins, RNA molecules, and organelles having been implicated in the regulation of some types of ACD (Sunchu and Cabernard, 2020; Zion et al., 2020). Epigenetic mechanisms play a crucial role in guiding the two daughter cells towards establishing differential gene expression profiles, ultimately defining their unique cell identities (Allis and Jenuwein, 2016; Escobar et al., 2021; Stewart-Morgan et al., 2020). During the ACD of Drosophila male germline stem cells (GSCs), pre-existing and newly synthesized histones H3 and H4 are asymmetrically segregated towards the stem daughter cell and the differentiating daughter cell, respectively (Wooten et al., 2020). Nevertheless, the extent to which epigenetic information is asymmetrically inherited through ACD in other organisms and the mechanism by which this process occurs remains elusive.

Caenorhabditis elegans represents a valuable model for investigating ACD, given its invariant cell lineage and conserved mechanisms of ACD. During hermaphrodite development in C. elegans, 131 somatic cells undergo programmed cell death, with the majority produced through ACD that create a large cell programmed for survival and a small cell programmed to die (Sulston and Horvitz, 1977; Sulston et al., 1983a) The opposing cell fates of daughter cells, i.e., to live or die, offer a compelling experimental system for investigating how epigenetic inheritance determines life versus death decisions during ACD. It is noteworthy that 105 of the 131 apoptotic cells arise from neuronal lineages, of which the Q neuroblast represents a tractable system for studying ACD and apoptosis at single-cell resolution (Ellis and Horvitz, 1986; Hedgecock et al., 1983; Ou et al., 2010; Sulston and Horvitz, 1977). During the first larval stage, Q neuroblasts on the left (QL) and right (QR) undergo three rounds of asymmetric divisions, which produce three different neurons and two apoptotic cells (Q.aa and Q.pp), respectively (Figure 1A).

Asymmetric segregation of NuRD during ACDs of C. elegans Q neuroblast.

(A) Schematic of the Q neuroblast lineages. QL or QR neuroblast each generates three neurons and two apoptotic cells (Q.aa/Q.pp, X). QL produces PQR, PVM, and SDQL. QR produces AQR, AVM, and SDQR.

(B) A model of the NuRD complex composition (Bracken et al., 2019; Lai and Wade, 2011).

(C) Protein domain structure of the GFP-tagged HDAC1/2 (HDA-1) or mNeonGreen-tagged RBBP4/7 (LIN-53). CAF1C: histone-binding protein RBBP4 or subunit C of CAF1 complex; WD: WD40 repeat tryptophan-aspartate domain. Scale bar: 100 amino acids.

(D) Representative images of endogenous HDA-1::GFP, LIN-53::mNeonGreen and overexpressed MYS-1::GFP during ACDs of QR.a. In each panel, the top row shows merged images, the middle row shows mCherry-tagged plasma membrane and histone, and the bottom row shows heatmaps of GFP/mNeonGreen fluorescence. The anterior of the cell is on the left. The GFP/mNeonGreen fluorescence intensity ratio between posterior and anterior chromatids or between QR.ap and QR.aa nuclei are shown in yellow at the lower right corner of heatmaps. Time is in minutes. Other frames are in Figure S3A, and the full movies are in Supplementary Movie S3-S5. Scale bar: 2 µm.

(E) Quantification of HDA-1::GFP, LIN-53::mNeonGreen and MYS-1::GFP fluorescence intensity ratio between QR.ap and QR.aa nuclei. Data are presented as mean ± SEM. N = 10 – 12.

(F) Quantification of HDA-1::GFP (magenta), LIN-53::mNeonGreen (orange) and MYS-1::GFP (gray) fluorescence intensity ratios between the posterior and anterior half of QR.a or between QR.ap and QR.aa. The anaphase onset is defined as the last frame without chromatids segregation. Data are mean ± SEM. N = 10 – 12.

(G) Quantification of HDA-1::GFP (left) and LIN-53::mNeonGreen (right) fluorescence intensity ratios between the large and small daughters of cells shown on the X-axis. Statistical significance for the fluorescence intensity ratio is determined using a one-sample t-test compared with the 1. Data are mean ± SEM, N = 7 – 14, ** p < 0.01; *** p < 0.001.

In the nematode, the classic apoptotic pathway initiates upon activation of the BH3-only protein EGL-1 in the cells that are fated to die. EGL-1 binds to the anti-apoptotic Bcl-2-like protein CED-9, which facilitates the release of the pro-apoptotic protein CED-4, leading to caspase CED-3 activation. Activated caspase promotes the exposure of phosphatidylserine (PS) on the surface of apoptotic cells, which triggers the phagocytosis of apoptotic cells via partially redundant signaling pathways such as the CED-1/CED-6/CED-7 pathway and the CED-2/CED-5/CED-12 pathway. Proper transcriptional regulation of the egl-1 gene is vital for life-versus-death decisions in C. elegans (Conradt and Horvitz, 1998; Nehme and Conradt, 2008). Although the cell-specific transcriptional regulator has been identified in certain cells, including the sexually dimorphic HSN and CEM neurons, M4 motor neuron, NSM neuron, P11.aaap cell, ABpl/rpppapp, and the tail spike cell (Hirose et al., 2010; Jiang and Wu, 2014; Nehme and Conradt, 2008), the general upstream regulators of egl-1 at the chromatin remodeling or epigenetic level are unclear. Two NSM neuroblasts undergo ACDs to each generate a larger NSM cell programed to survive and a smaller NSM sister cell programed to die (Sulston et al., 1983b). The PIG-1 kinase-dependent asymmetric partitioning of the Snail-like transcription factor, CES-1, represses egl-1 transcription in the larger NSM sister cells, thereby preventing their apoptosis (Hatzold and Conradt, 2008; Wei et al., 2020) However, the mechanism underlying the asymmetrical activation of egl-1 in other ACDs remains elusive.

In this study, we demonstrate the enrichment of the NuRD complex in cells that are predetermined to survive, and its role in suppressing the EGL-1-CED-9-CED-4-CED-3 apoptotic pathway through repression of the egl-1 gene. Furthermore, we report the interaction between the NuRD complex and the V-ATPase complex, and reveal that during asymmetric cell division, the localization of the V-ATPase complex and cytoplasmic pH are asymmetrical, thereby contributing to the polarized segregation of the NuRD complex.

Results

NuRD asymmetric segregation during neuroblast ACDs

In order to gain insights into the molecular distinctions between apoptotic and surviving cells in C. elegans embryos, we employed the SPLiT single-cell RNA sequencing platform to compare their transcriptomes (Figure S1A-S1C) (Rosenberg et al., 2018). The expression of the critical somatic apoptosis-inducing gene egl-1 was used to distinguish the apoptotic and surviving cells (Conradt et al., 2016). Intriguingly, we identified transcripts encoding subunits of the nucleosome remodeling and deacetylase (NuRD; also known as Mi-2) complex in cells where egl-1 expression was indiscernible (Figure S1D-S1F, Supplementary Table 1). The NuRD complex represents an evolutionarily conserved protein complex associated with chromatin that couples the activities of chromatin-remodeling ATPases with histone deacetylases (Figure 1B) (Bracken et al., 2019; Lai and Wade, 2011). NuRD-mediated alterations in chromatin structure are vital for appropriate transcriptional regulation during cell fate determination and lineage commitment (Bracken et al., 2019; Lai and Wade, 2011), suggesting that NuRD may serve as a candidate epigenetic factor to specify cell fate between survival and death.

To investigate whether NuRD components are distributed asymmetrically between apoptotic and surviving daughter cells, we expressed GFP-tagged NuRD subunits within the C. elegans Q neuroblast lineages. Q.a neuroblast undergoes ACD to generate a large surviving Q.ap cell and a small apoptotic Q.aa cell (Figure 1A). We found that GFP-tagged NuRD subunits, including the histone deacetylase HDA-1, the histone binding protein LIN-53, the nucleosome remodeling factor CHD-3, and the MEP-1 protein were enriched asymmetrically in the surviving QR.ap cell during QR.a division (Figure S2, Supplementary Movie S1 and S2). Notably, NuRD asymmetric distribution occurred within several minutes after metaphase, while GFP protein translation and chromophore maturation require approximately 30 min in Q neuroblast (Ou and Vale, 2009), suggesting that the asymmetric enrichment of NuRD is likely the result of protein redistribution.

To investigate the dynamic distribution of endogenous NuRD during ACD, we generated a GFP knock-in (KI) strain for HDA-1 and an mNeonGreen (NG, green fluorescence) KI line for LIN-53 using CRISPR-Cas9 (Figure 1C). In QR.a cells, nuclear HDA-1 and LIN-53 were released into the cytoplasm and evenly distributed until metaphase (Figure 1D, Supplementary Movie S3 and S4). At anaphase, HDA-1 and LIN-53 became enriched in the posterior part of QR.a but became less detectable in the anterior part (Figure 1D, S3A-B, Supplementary Movie S3 and S4). By quantifying the fluorescence intensity ratio between posterior and anterior chromatids of QR.a or QR.ap and QR.aa nuclei, we found that nuclear HDA-1 or LIN-53 asymmetry gradually increased from 1.1-fold at the onset of anaphase to 1.5 or 1.8-fold upon completion of cytokinesis, respectively (Figure 1D-E). We measured the ratios of fluorescence intensities between the posterior and anterior halves of QR.a or between QR.ap and QR.aa (Figure S4A, see Materials and Methods). NuRD asymmetry became evident at ∼4 minutes and reached a plateau at ∼6 minutes after the anaphase onset (time 0 min, the last frame before the sister chromosome separation; Figure 1D, 1F, S3A and S4B). QR.a spent ∼6 minutes from anaphase to the completion of cytokinesis (Chai et al., 2012; Ou et al., 2010) indicating that QR.a cell establishes NuRD asymmetry during ACD.

Similar to QR.a, the dynamic and progressively developing NuRD asymmetry was also observed in the QL.a, QR.p and QL.p cell lineages, which generate apoptotic cells (Figure 1G, S3B and S4B). In contrast, an even distribution of HDA-1 and LIN-53 was observed in two surviving daughters of QR, QL, QL.pa and QR.pa cells, which generate two viable siblings (Figure 1G and S3B). To investigate whether other epigenetic factors are also asymmetrically segregated during Q cell ACDs, we tagged the MYST family histone acetyltransferase (MYS-1 in C. elegans) with GFP. We found that MYS-1::GFP was symmetrically segregated into the apoptotic and surviving daughter cells (Figure 1D-F, S3A and S4B, Supplementary Movie S5), indicating the specificity of the polarized NuRD partition. The total fluorescence of HDA-1, LIN-53, and MYS-1 remained constant during ACDs, suggesting that protein redistribution may establish NuRD asymmetry (Figure S4C). Inhibition of a PAR-1 family kinase PIG-1 led to symmetric Q cell divisions, producing extra neurons derived from some of their apoptotic daughters (Cordes et al., 2006). In pig-1 (gm344) mutants, HDA-1 and LIN-53 were evenly partitioned during ACDs (Figure S5, Supplementary Movie S6 and S7), indicating that asymmetric NuRD segregation depends on the PIG-1 kinase.

NuRD asymmetric segregation in embryonic cell lineages

Given that 113 out of 131 somatic apoptotic events in C. elegans occur during embryonic development (Sulston et al., 1983a), we used live imaging and automated cell lineage tracing algorithms (Du et al., 2014) to track NuRD asymmetry in embryos from the two- or four-cell stage up to the 350-cell stage For example, approximately 200 minutes after the first embryonic division, ABalaappp divides to generate two viable daughters while its sibling ABalaappa divides asymmetrically and produces one daughter that undergoes apoptosis (Figure 2A). Similar to what we observed in the Q lineage, HDA-1 was evenly distributed in the two viable daughters of ABalaappp, while it was asymmetrically segregated during the ACD of ABalaappa (Figure 2B). Overall, we found that HDA-1 and LIN-53 were enriched (average ratio > 1.5) in the surviving daughters from 15 out of 17 embryonic ACDs that produce apoptotic daughter cells (Figure 2C-D and Supplementary Table 2). There were two exceptions, ABaraaaap and MSpppaa, whose two daughter cells received similar amounts of NuRD (Figure 2C-D). While MSpppaa’s apoptotic daughter cell undergoes apoptosis about 400 minutes after its generation (Hsieh et al., 2012; Sulston et al., 1983b), consistent with the idea that a low concentration of NuRD may promote apoptosis, ABaraaaap’s apoptotic daughter cell completes apoptosis within 35 minutes (Sulston et al., 1983a), suggesting that a NuRD-independent mechanism may regulate apoptosis in this cell.

NuRD Asymmetry in C. elegans embryonic cell lineages.

(A) An x and y plane of 3D time-lapse imaging data at the ∼200 min after the first embryonic division (left) in the HDA-1::GFP knock-in animal. H2B::mCherry was expressed in the embryo ubiquitously. The bright green circle indicates the ABalaappa cell, whose division generates an apoptotic daughter cell. The cyan circle indicates the ABalaappp cell, whose division generates two surviving daughters. The orientation of the embryo is shown in the middle. The cell lineage of ABalaapp is shown on the right. Each vertical line represents a cell and each branch represents a cell division. X indicates an apoptotic cell. Scale bar: 5 μm.

(B) Representative images of dividing ABalaappp cell (top panel) and ABalaappa cell (bottom panel) from the lineage tracing data. GFP fluorescence is shown as heatmaps below merged images. Cyan or bright green lines mark the chromatids or nuclei of daughter cells according to the H2B::mCherry signal and the published algorithm to automatically track cells in 3D pictures (Du et al., 2014). The black and white arrowheads point to the anterior and posterior daughter cells, respectively. Scale bar: 5 µm.

(C) Tree visualization of HDA-1::GFP and LIN-53::mNeonGreen segregation between sister cells during embryonic development. Cells and their lineage relations are visualized as a tree structure, with vertical lines indicating cells and horizontal lines indicating cell divisions. Green vertical lines show cells with a significantly higher (>1.5-fold) nuclear HDA-1::GFP and LIN-53::mNeonGreen intensity than their sister cells. Red arrowheads point to apoptotic cells. Placement on the tree is according to the Sulston nomenclature. See also Supplementary Table 2.

(D) Quantifications of HDA-1::GFP and LIN-53::mNeonGreen fluorescence intensity ratios between nuclei of the live daughter cell and its apoptotic sister cell automatically identified and tracked in 3D time-lapse imaging of embryonic development (Supplementary Table 2). The lineage names of non-apoptotic sister cells of 17 cell pairs with apoptotic cells are shown on the X-axis. The apoptotic sister cell of MSpppaap (magenta circle) completes apoptosis approximately 400 minutes after birth. N = 3 - 4 in each measurement. Data are mean ± SEM. One-sample t-test compared to 1, *p< 0.05, **p< 0.01, ***p< 0.001, ns: not significant.

Loss of the deacetylation activity of NuRD causes ectopic apoptosis

To investigate the role of NuRD in determining cell fate, we reduced the deacetylation activity of NuRD using RNA-mediated interference (RNAi). The hda-1 and lin-53 RNAi efficacy was confirmed by the reduced green fluorescence in the germlines of HDA-1::GFP and LIN-53::mNeonGreen KI animals, as well as the increased acetylation level of histone H3K27 (Figure S6). As reported before, RNAi of either hda-1 or lin-53 caused embryonic lethality (Shi and Mello, 1998). RNAi of lin-53 led to embryos arrested at gastrulation stage before the generation of apoptotic cells, preventing us from analyzing its roles in somatic apoptosis. In contrast, hda-1 RNAi embryos arrested between the late gastrulation stage and bean stage, allowing us to tracked some of apoptotic events. To quantify apoptotic events, we used a secreted Annexin V (sAnxV::GFP) sensor to label apoptotic cells with externalized phosphatidylserine (exPS) on the surface of the plasma membrane in ced-1(e1735) mutant embryos which inhibits engulfment of apoptotic cells (Conradt et al., 2016; Mapes et al., 2012; Zhou et al., 2001). The sAnxV::GFP labeled an average of 10 cells in ced-1(e1735) mutant embryos treated with control RNAi and an average of 14 cells in ced-1(e1735); hda-1 RNAi embryos (Figure 3A-B). To test whether the upregulated apoptosis in hda-1 RNAi embryos depends on the canonic apoptosis pathway, we introduced cell death-deficient mutations ced-3/ Caspase (n2433), ced-4/ APAF-1-like (n1162), ced-9/ BCL-2-like (n1950), and egl-1/ BH-3 only protein (n1084n3082) into ced-1 mutant, respectively. RNAi of hda-1 in these embryos only generated 0-2 exPS-positive cells (Figure 3A-B and S7A), indicating that the ectopic increase of apoptosis by HDA-1 inhibition requires the EGL-1-CED-9-CED-4-CED-3 pathway. Considering the pleiotropic phenotypes caused by loss of HDA-1, we cannot exclude the possibility that ectopic cell death might result from global changes in development, even though HDA-1 may directly contribute to the life-versus-death fate determination.

RNAi of hda-1 induces ectopic apoptosis and increases H3K27 acetylation of the egl-1 gene.

(A) Fluorescent images show Phsp::sAnxV::GFP from ced-1(e1735) and ced-1(e1735); ced-3(n2433) embryos between late gastrulation and bean stage, treated with control RNAi or hda-1 RNAi. Scale bars, 5 μm.

(B) Quantifications of cell corpse number in the embryos (N = 39 - 60) of indicated genotypes. Data are presented as mean ± SEM. Unpaired two-tailed parametric t-test, compared with the control RNAi, ***p < 0.001.

(C) Scatter plot of the increased gene expression in both hda-1 RNAi and lin-53 RNAi animals. The egl-1 gene is marked in red. See also Supplementary Table 3.

(D) Quantitative real-time PCR (RT-PCR) measurement of egl-1 mRNA levels in the control, hda-1 or lin-53 RNAi embryos. Fold induction was calculated relative to levels in control RNAi embryos. Data represent the mean ± SEM of six biological replicates. Unpaired parametric t-test, ***p < 0.001.

(E) Quantification of the number of cells expressing the Pegl-1::NLS::GFP reporter in embryos treated with control, hda-1 or lin-53 RNAi. Data are mean ± SEM and are compared by unpaired parametric t-test, *p < 0.05, **p < 0.01, ***p < 0.001. N = 8 – 9.

(F) Normalized ChIP-Seq signal profiles of the H3K27 acetylation level in the control, hda-1 or lin-53 RNAi embryos at the egl-1 locus. The Y-axis shows the average sequencing coverage of bins per million reads of three biological replicates normalized to the input control sample. Ectopic H3K27ac enrichments at the egl-1 locus of hda-1 RNAi and lin-53 RNAi embryos are highlighted in magenta and cyan.

(G) ChIP-qPCR analyses using H3K27ac antibody or IgG (the negative control) at selected elements of egl-1 indicated in (F) by red lines. Results are shown as the percentage of input DNA. Data are mean ± SEM from three biological replicates. Unpaired parametric t-test, *p < 0.05.

NuRD RNAi upregulates the egl-1 expression by increasing its H3K27 acetylation

To understand how NuRD regulates apoptotic cell fates, we performed RNA-Seq analyses on WT, hda-1 RNAi and lin-53 RNAi embryos. In both RNAi conditions, we observed 287 or 575 genes that were upregulated or downregulated (|log2FD| > 2 and FDR < 0.05) (Supplementary Table 3). Notably, we found that the apoptosis-inducing gene egl-1 was among the upregulated genes (Figure 3C). Our quantitative RT-PCR results further confirmed that egl-1 transcripts were more abundant in hda-1 RNAi and lin-53 RNAi embryos than in control RNAi embryos (Figure 3D). We also used a Pegl-1::NLS::GFP transcriptional reporter to examine the egl-1 transcription activity during embryonic development, which allowed us to detect egl-1 expression while avoiding the adverse effects of egl-1 overexpression on apoptosis. We found that RNAi of hda-1 or lin-53 caused an increased number of Pegl-1::NLS::GFP positive cells (Figure 3E and S7B). These results indicate that the loss of NuRD abnormally activates the egl-1 gene expression.

Next, we investigated whether the loss of histone deacetylase activity in NuRD increased the H3K27ac level at the egl-1 locus, as the histone H3K27ac level is associated with transcription activation (Heinz et al., 2015). To test this, we performed ChIP-Seq experiments with the anti-H3K27ac antibody and found that the transcription start site (TSS) and the gene body region of egl-1 had higher levels of H3K27ac in hda-1 or lin-53 RNAi embryos compared to WT, while the H3K27ac level of genes adjacent to egl-1 showed no significant changes (Figure 3F). This was validated by ChIP-qPCR in hda-1 RNAi embryos (Figure 3G). Therefore, inhibiting NuRD increased the H3K27ac level at the egl-1 locus, leading to its upregulation and subsequent activation of apoptosis.

V-ATPase regulates asymmetric segregation of NuRD during somatic ACDs

To investigate how NuRD is asymmetrically segregated during ACD, we used affinity purification with an anti-GFP antibody and mass spectrometry (MS) to isolate NuRD binding partners from the lysate of HDA-1::GFP KI animals. Our analysis revealed all the previously known NuRD subunits from HDA-1::GFP KI animals (Figure S8, Supplementary Table 4) (Bracken et al., 2019; Lai and Wade, 2011), validating our experimental method. Interestingly, our co-immunoprecipitation and MS with HDA- 1::GFP identified 12 subunits of the vacuolar-type H+-ATPase (V-ATPase) (Figure 4A-B), consistent with a previous report that used FLAG-tagged LIN-53 to identify 4 V-ATPase subunits (Muthel et al., 2019). We further confirmed the binding of HDA-1 to V-ATPase subunits in C. elegans by co-immunoprecipitation and western blot analysis with anti-V1A/VHA-13 antibodies (Figure 4C). V-ATPase is a proton pump that comprises a transmembrane domain (V0) responsible for proton transport across membranes and a peripheral ATP-hydrolytic domain (V1) that catalyzes ATP hydrolysis (Figure 4B). V-ATPases play a critical role in intracellular pH homeostasis and regulate numerous physiological processes, including apoptosis (Abbas et al., 2020; Chen et al., 2022; Ernstrom et al., 2012; Vasanthakumar and Rubinstein, 2020). However, the role of V-ATPase in the asymmetric segregation of epigenetic factors during ACD remains unknown.

HAD-1 interacts with subunits of V-ATPase.

(A) The plot compares counts of proteins co-precipitated with HDA-1::GFP with those with the control ACT-4 (actin)::GFP. The PSM (Peptide-Spectrum Match) is the number of identified peptide spectra matched for the protein. Blue dots represent the subunits of V-ATPase. See also Supplementary Table 4.

(B) Schematic model of the V-ATPase complex. The known worm subunits are indicated. The mean PSM of the encoded protein from two co-IP and MS repeats is shown in blue after the gene name.

(C) Western blot (WB) showing co-immunoprecipitation (co-IP) of V-ATPase V1 domain A subunit (V1A) with the HDA-1 from worm lysates. Assay was performed using three biological replicates. Three independent biological replicates of the experiment were conducted with similar results

To investigate the role of V-ATPase in NuRD asymmetric segregation in Q neuroblast, we used a pharmacologic inhibitor, bafilomycin A1 (Baf A1), to inhibit V-ATPase proton pumping activity (Furuchi et al., 1993; Wang et al., 2021; Yoshimori et al., 1991). To assess the inhibitory effect of Baf A1 on the proton translocation activity of V-ATPase in Q neuroblast, we monitored the cytosolic pH dynamics in dividing QR.a using the super-ecliptic pHluorin. The pHluorin is a pH-sensitive GFP reporter whose fluorescence can be quenched by the acidic pH (Miesenbock et al., 1998; Sankaranarayanan et al., 2000). Although BafA1-mediated disruption of lysosomal pH homeostasis is recognized to elicit a wide array of intracellular abnormalities, we found no evidence of such pleiotropic effects at the organismal level with the dosage and duration of treatment employed in this study. In DMSO-treated animals, the pHlourin fluorescence intensity remained constant in the posterior portion that forms QR.ap but was significantly reduced in the anterior portion that forms QR.aa, indicating that the cytoplasm of the future apoptotic cell became more acidic (Figure 5A-B and Supplementary Movie S8). This observation is consistent with the recognition that cytosolic acidification is a common feature of both death receptor-mediated and mitochondria-dependent apoptosis (Lagadic-Gossmann et al., 2004; Matsuyama et al., 2000; Matsuyama and Reed, 2000). Notably, Baf A1 treatment reduced the pHluorin fluorescence intensity ratio between QR.ap and QR.aa from 1.6-fold to 1.3-fold (Figure 5A-B and Supplementary Movie S8), suggesting that Baf A1 may disrupt the cytosolic pH asymmetry in dividing QR.a cells by inhibiting V-ATPase activity, although we cannot exclude the possibility that the changes in fluorescence could be due to changes in the amount of pHluorin protein.

V-ATPase regulates NuRD asymmetric segregation and cell fates.

(A) Dynamics of the cytosolic pH indicated by super-ecliptic pHluorin during QR.a division in DMSO control or Baf A1-treated animals. In each panel, the top row shows mCherry-tagged plasma membrane and histone, and the bottom row shows heatmaps of super-ecliptic pHluorin signal. The anterior of the cell is on the left. Time 0 min is the onset of anaphase: scale bar, 2 µm. See also Supplementary Movie S8.

(B) The super-ecliptic pHluorin fluorescence intensity ratio between QR.ap and QR.aa. Data are mean ± SEM. N =11-15. Unpaired two-tailed parametric t-test, **p < 0.01, ns: not significant.

(C) Images of HDA-1::GFP distribution during QR.a cell division in DMSO control or Baf A1-treated animals. In each panel, the top row shows mCherry-tagged plasma membrane and histone, and the bottom row shows heatmaps of HDA-1::GFP fluorescence. The GFP fluorescence intensity ratios between the posterior and anterior chromatids or between QR.ap and QR.aa nuclei are shown in green at the lower left corner of heatmaps. Anterior of the cell is left. Scale bar: 2 µm. See also Supplementary Movie S9.

(D) Quantification of HDA-1::GFP fluorescence intensity ratios between the large and small daughter cell nuclei and the cell size ratios between the large and small daughters. The names of mother cells are shown on the X-axis. Data are mean ± SEM. N = 9-12. Unpaired two-tailed parametric t-test, **p < 0.01, ***p < 0.001, ns: not significant.

(E) Representative images showing fates of QR.aa after DMSO (top), Baf A1 (middle) and Baf A1 plus AID treatment (bottom). The Q cell plasma membrane and chromosome are labeled by mCherry. HDA-1::GFP is shown as inverted images. Yellow arrow head shows a short neurite-like outgrowth of QR.aa in Baf A1 treated larvae. Frequencies of QR.aa survival and HDA-1 maintenance are showed on the right. Time 0 min is the birth of QR.aa. Scale bar: 5 µm.

We observed that treatment with Baf A1 resulted in the symmetric segregation of HDA-1 during QR.a division (Figure 5C-D and Supplementary Movie S9). Notably, neither DMSO nor Baf A1 treatment affected the asymmetry in daughter cell size (Figure 5C-D and Supplementary Movie S9), suggesting that intracellular acidification, regulated by V-ATPase activity, specifically affects NuRD asymmetric segregation without influencing daughter cell size asymmetry. We also investigated whether the small QR.aa cell carrying ectopic NuRD could escape apoptosis. In DMSO-treated animals, QR.aa cells underwent apoptosis, and 12 out of 13 QR.aa cells were engulfed and degraded by a neighboring epithelial cell within 120 min after birth. However, in Baf A1-treated animals, QR.aa inherited similar levels of HDA-1::GFP as its sister cell, and 11 out of 12 QR.aa cells carrying ectopic NuRD survived for over 120 min and formed a short neurite-like outgrowth (Figure 5E). To confirm whether the long-lived QR.aa was a consequence of ectopic gain of NuRD, we depleted HDA-1 within Q cell lineages using the auxin-inducible protein degradation (AID) system (Zhang et al., 2015). By introducing the degron sequence into the hda-1 locus and expressing the TIR1 F-box protein under the Q cell-specific egl-17 promoter, we showed that HDA-1::GFP fluorescence in Q cells was substantially decreased under auxin treatment (Figure 5E). Administration of auxin restored QR.aa apoptosis in Baf A1-treated animals (Figure 5E). These results suggest that V-ATPase activity-dependent NuRD asymmetric segregation contributes to the specification of the live-death fate.

To understand how V-ATPase regulates NuRD asymmetric segregation, we generated a transgenic strain expressing wrmScarlet-tagged VHA-17, which is the e subunit of the V0 domain. Using this strain, we were able to examine the dynamic distribution of V-ATPase during QR.a cell division. Like HDA-1, V-ATPase was also asymmetrically enriched in the surviving QR.ap portion (Figure 6A-B and Supplementary Movie S10). This observation suggests that V-ATPase and NuRD may co-segregate asymmetrically during ACD. We also found that Baf A1 treatment disrupted V-ATPase asymmetric distribution (Figure 6A-B and Supplementary Movie S10), indicating the importance of the proton-pumping activity of V-ATPase in its asymmetric segregation. Therefore, our results suggest that V-ATPase may facilitate the asymmetric distribution of NuRD through its proton pumping activity.

: V-ATPase distribution during ACDs and a model.

(A) Dynamics of VHA-17::wrmScarlet during QR.a cell division in DMSO control or Baf A1-treated animals. VHA-17::wrmScarlet fluorescence is shown as heatmaps. Time 0 min is the onset of anaphase. Scale bar: 2 µm.

(B) Quantification of VHA-17::wrmScarlet (magenta) and HDA-1::GFP (blue) fluorescence intensity ratios between the posterior and anterior half of QR.a or between QR.ap and QR.aa. Data are mean ± SEM. N = 10 – 14. One-sample t-test compare to 1, ***p < 0.001, ns: not significant. See also Supplementary Movie S10.

(C) A proposed model. Asymmetric segregation of V-ATPase mediated the enrichment of its associated NuRD in the large daughter cell, where high level of NuRD deacetylates egl-1 and suppressed its expression.

Discussion

Our findings indicate that the asymmetric segregation of the NuRD complex during ACD is regulated in a V-ATPase-dependent manner, and is critical for the differential expression of apoptosis activator egl-1 and the life-versus-death fate decision (Figure 6C). Specifically, we have provided evidence that the reduced level of NuRD in the apoptotic daughter cell leads to an increased level of H3K27ac, an epigenetic modification known to be associated with active gene expression, at the egl-1 locus, resulting in the upregulation of egl-1 expression and subsequent induction of apoptosis. Based on our results, we propose a dual-function model of V-ATPase in NuRD asymmetric segregation: (i) V-ATPase interacts with NuRD, both of which are asymmetrically segregated into the surviving cell; (ii) the asymmetrical activity of V-ATPase generates a more acidic cytoplasmic environment in the future apoptotic cell portion compared to the surviving portion, which might contribute to the asymmetric segregation of V-ATPase and NuRD. This model sheds light on the mechanisms underlying the regulation of apoptosis during ACD, and opens up new avenues for further investigation into the intricate interplay between V-ATPase, NuRD, and epigenetic modifications in the context of cell fate decisions.

The intrinsic mechanisms governing binary cell fate decisions involve asymmetric cortical localization of cell fate determinants, polarized partitioning of RNA species, unequal segregation of organelles, and biased distribution of damaged proteins or protein aggregates (Sunchu and Cabernard, 2020; Venkei and Yamashita, 2018). Despite this, little is known about the asymmetric segregation of epigenetic modification enzymes during ACDs, nor the functions of polarized V-ATPase distribution and cytoplasmic proton asymmetry during this process. Our findings provide new insights into the asymmetric segregation of cell-intrinsic factors and the previously unrecognized roles of V-ATPase and cytosolic acidification in this process.

Two mechanisms have been proposed to explain the V-ATPase-dependent asymmetric segregation of NuRD. Firstly, a polarized intracellular transportation system could selectively deliver NuRD-V-ATPase-containing organelles to the surviving daughter cell. Although V-ATPase is primarily known for its localization and function in the late endosome and lysosome, recent evidence suggests that V-ATPase subunits are synthesized and assembled on the endoplasmic reticulum (ER) (Abbas et al., 2020; Graham et al., 2003; Wang et al., 2020). Our findings demonstrate that V-ATPase colocalizes with the ER marker during ACDs, suggesting that NuRD-V-ATPase may be delivered as cargo on the ER or ER-derived vesicles (Figure S9). As reported in previous studies, a polarized microtubule transportation system, governed by specific motor proteins and the microtubule tracks, is responsible for the asymmetric segregation of signaling endosomes during ACDs of Drosophila sensory organ precursors (Derivery et al., 2015). It is possible that a similar biased transportation system plays a role in the asymmetric segregation of NuRD-V-ATPase-containing organelles.

Secondly, the surviving daughter cell chromatids may have a greater ability to recruit NuRD than their apoptotic sister. This could be due to a higher number of yet unknown NuRD recruiting factors on chromatids or specific posttranslational modifications of chromosomal proteins that enhance the chromosomal recruitment of NuRD. Despite having an identical DNA sequence, sister chromatids can be distinguished by chromosomal proteins or modifications (Ranjan et al., 2019; Tran et al., 2012), which likely provides the specific molecular cues for asymmetric NuRD recruitment. The NuRD asymmetry on chromosomes (1.5∼1.8-fold between QR.ap and QR.aa nuclei) is greater than that between the two daughter cells (approximately 1.3-fold) (Figure 1D-1F), suggesting that polarized NuRD-V-ATPase transportation and distinct NuRD affinity from sister chromatids may act in concert to establish NuRD asymmetry.

The observation of asymmetric segregation of the NuRD complex during cell divisions suggests that the mother cell may play a role in initiating the fate specification of its daughter cells. This notion is supported by studies conducted on Drosophila male germline stem cells (GSCs) which demonstrated that the old and new histone H3-H4 or unequal amounts of histone H3 variant CENP-A were incorporated into sister chromatids prior to cell division (Ranjan et al., 2019; Tran et al., 2012). Although these patterns were not observed in cell types other than GSCs, asymmetric NuRD segregation was found to be common in C. elegans (Figure 1D-G, Figure 2, and Figure S3). Among the 395 cell lineages that were tracked in C. elegans embryos and generated two surviving daughter cells, NuRD was equally distributed to both daughters in 390 instances (Figure 2C-2D, Supplementary Table 2). However, in the remaining five cell divisions, both HDA-1 and LIN-53 exhibited >1.5-fold differences between the two surviving siblings (Figure 2C-2D, Supplementary Table 2). This observation suggests that biased NuRD segregation may contribute to binary cell fate decisions beyond apoptosis. In the embryonic cerebral cortex of mammals, approximately 30% of newborn cells undergo programmed cell death (Voiculescu et al., 2000), which is a significantly higher rate than observed in C. elegans where only 12% of somatically born cells undergo apoptosis (131 apoptotic cells among 1090 somatically born cells). Thus, it is worth exploring whether brain stem cell divisions employ asymmetric NuRD segregation to specify daughter cell fates. Further studies on polarity establishment hold promise in elucidating the mechanisms underlying the asymmetric inheritance of epigenetic information.

Materials and Methods

Worm strains and culture

C. elegans were maintained as described in the standard protocol (Brenner, 1974). All strains were cultivated at 20 °C on nematode growth medium (NGM) agar plates seeded with Escherichia coli OP50 or HT115 (feeding RNAi assay). The wild-type strain was Bristol N2. Some strains were provided by the Caenorhabditis Genetics Center (CGC), funded by the NIH Office of Research Infrastructure Programs (P40 OD010440). The strains used in this study were listed in Extended Data Table 1.

Molecular biology

We performed genome editing experiments in C. elegans following the established methods (Dickinson et al., 2013; Friedland et al., 2013). We used the CRISPR design tool (https://zlab.bio/guide-design-resources) to select the target sites. The sgRNA sequences (Extended Data Table 2) were inserted into the pDD162 vector (Addgene #47549) by linearizing the vector with 15 base pairs (bp) overlapped primers using Phusion high-fidelity DNA polymerase (New England Biolabs, cat. no. MO531L). PCR products were digested using Dpn I (Takara, cat. no. 1235A) for 2 hours at 37°C and transformed into Trans5 α bacterial chemically competent cells (TransGen Biotech, Cat no. CD201-01). The linearized PCR products with 15 bp overlapping ends were cyclized to generate plasmids by spontaneous recombination in bacteria. Homology recombination (HR) templates were constructed by cloning the 1.5 kb upstream and downstream homology arms into the pPD95.77 plasmid using the In-Fusion Advantage PCR cloning kit (Clontech, Cat. #639621). Fluorescence tags were inserted into the constructs with a flexible linker before the stop codons. Synonymous mutations were introduced to Cas9 target sites to avoid the cleavage of the homologous repair template by Cas9. The plasmids were listed in Extended Data Table 3.

Constructs that express GFP-tagged NuRD components were generated using the PCR SOEing method(Hobert, 2002). N2 genomic sequences (1.5-2 kb promoter plus coding region) were linked with gfp::unc-54 3’ UTR DNA fragments. Constructs that express wrmScarlet-tagged V-ATPase components with gfp::unc-54 3’ UTR were generated using the In-Fusion Advantage PCR cloning kit (Clontech, Cat. #639621). The primers were listed in Extended Data Table 3.

Genome editing and transgenesis

To generate a knock-in strain, we purified the sgRNA construct and the repair template plasmids with the PureLink Quick PCR purification Kit (Invitrogen, #K310001) and co-injected them into N2 animals with the pRF4 [rol-6 (su1006)] and Podr-1::dsRed selection markers. The F1 transgenic progenies were singled and screened by PCR and Sanger sequencing. Transgenic C. elegans was generated by the germline microinjection of DNA plasmids or PCR products into N2 with Podr-1::gfp or Pegl-17::myri-mCherry and Pegl-17::mCherry::his-24 plasmids. We maintained at least two independent transgenic lines with a constant transmission rate (>50%). Concentrations of DNA constructs used for generating knock-in were 50 ng/µl, and for transgenesis were 20 ng/µl. All the strains, primers, plasmids, and PCR products were listed in Extended Data Table 1-3, respectively.

Live-Cell Imaging and quantification

Live imaging of C. elegans embryos, larvae, or young adults followed our established protocols (Chai et al., 2012; Zhang et al., 2017). L1 larvae or young-adult worms were anesthetized with 0.1 mmol/L levamisole in M9 buffer, mounted on 3% (wt/vol) agarose pads maintained at 20°C. Our imaging system includes an Axio Observer Z1 microscope (Carl Zeiss MicroImaging) equipped with a Zeiss 100×/1.46 numerical aperture (NA) objective, an Andor iXon+ EM-CCD camera, and 488 and 561-nm lines of a Sapphire CW CDRH USB Laser System attached to a spinning disk confocal scan head (Yokogawa CSU-X1 Spinning Disk Unit). Images were acquired by μManager (https://www.micro-manager.org). Time-lapse images of Q cell divisions were acquired with an exposure time of 200 msec every 1 min. Recent advancements in optical and camera technologies permit the acquisition of Z-stacks without perturbing Q cell division or overall animal development. Z-stack images were acquired over a range of -1.6 to +1.6 μm from the focal plane, at intervals of 0.8 μm. The field-of-view spaned 160 μm × 160 μm, and the laser power, as measured at the optical fiber, was approximately 1 mW. ImageJ software (http://rsbweb.nih.gov/ij/) was used to perform image analysis and measurement. Image stacks were z-projected using the average projection for quantification and using the maximum projection for visual display. To follow the cell fate of QR.aa in WT, Baf A1- or Baf A1 and IAA-treated worms, time-lapse images were acquired with 300 msec for 561 nm and 100 msec for 488 nm exposure time at 5-minute intervals. To image Phsp-16.41::sAnxV::GFP positive cells in WT, mutant, or RNAi animals, we incubated the eggs at 33 °C for 30 min to induce the reporter expression 3 hours post-egg-laying. After we kept the eggs at 20 °C for 2 hours, we imaged them with an exposure time of 200 msec. To acquire time-lapse images of Pegl-1::NLS::GFP positive cells during embryonic development, we dissected 2-4 cell-stage embryos from gravid adults and imaged them with an exposure time of 300 msec every 5 min.

The quantifications of cellular fluorescence intensity ratios in Q cell lineages were described in Figure S4A. We used the mCherry-labeled plasma membrane to circumscribe Q cells (Region of Interest). To determine the ratios of fluorescence intensities in the posterior to anterior half (P/A) of Q.a lineages or A/P of Q.p lineages, the cell in the mean intensity projection was divided into posterior and anterior halves. ImageJ software was used to measure the fluorescence intensities of two halves with the background subtracted. The fluorescence intensity of the slide background was measured from a region without worm bodies, of the same size as the region of interest. The background-subtracted mean fluorescence intensities in the two halves were then normalized by the area and the two resulting values were divided. The same procedure was used to determine the ratios of fluorescence intensities between two daughter cells. Total fluorescence intensity was the sum of the posterior and anterior fluorescence intensities or the sum of fluorescence intensities from two daughter cells without area normalization. The ROIs used in the measurement and calculation of the nuclear intensity ratios were determined as regions with mCherry-tagged histone in each cell.

Feeding RNAi

RNAi bacteria were grown in LB media containing carbenicillin (50 µg/mL) and tetracycline hydrochloride (12.5 µg/mL) at 37°C overnight. RNAi bacteria were then seeded on NGM plates supplemented with 50 µg/mL carbenicillin and 1 mM Isopropyl β-D-thiogalactopyranoside (IPTG). We grew the seeded RNAi plates for 2 days at room temperature, allowing double-stranded RNA expression. To score phenotypes during embryogenesis or meiosis, we transferred the synchronized young adult worms to the culture plate containing the induced E. coli HT115 strains expressing the control (luciferase) or specific gene-targeted dsRNAs. We incubated them for 16-24 hours at 20°C and then scored the phenotypes. As for RT-qPCR and RNA-Seq experiments, parental animals of RNAi were used for laying eggs, and the progeny was collected 5 hours post-egg-laying. In ChIP-Seq experiments, we used worms of mixed stages under RNAi treatment for 24-36 hours. RNAi bacterial strains were from the Vidal library.

Immunoprecipitation and mass spectrometry

GFP transgenic or knock-in worms were raised on one hundred 90-mm NGM plates. Animals were collected and washed with M9 buffer three times. For each replicate, the lysate was made from 1∼2 ml packed worms with 3∼4 ml of 0.5-mm diameter glass beads using FastPrep-24 (MP Biomedicals) in lysis buffer [pH 7.4, 150 mM NaCl, 25 mM Tris-HCl, 10% glycerol, 1% NP-40, 1x cocktail of protease inhibitors from Roche (Complete, EDTA free), 40 mM NaF, 5 mM Na3VO4]. Worm lysates were then cleared by centrifugation at 14,000g for 30 min at 4 °C. For anti-GFP immunoprecipitation (IP), the supernatant was incubated with GFP-Trap A beads (Chromoteck, GTA20) for 1 hr. Beads were then washed three times with lysis buffer. To detect the interaction between the V-ATPase subunit and HDA-1, beads are boiled with 1X SDS-PAGE sample loading buffer for SDS–PAGE and western blot. Otherwise, proteins were eluted from beads with 300 μl 0.1 M glycine-HCl (pH 2.5) Add 15 μl 1.5 M Tris-HCl (pH 8.8) to the eluted protein, followed by precipitation using 100 μl trichloroacetic acid and re-dissolution in 60 μl 8 M Urea, 100 mM Tris-HCl (pH 8.5) for MS. MS Samples were then reduced in 5 mM TCEP, alkylated by 10 mM iodoacetamide, and diluted four-fold with 100 mM Tris-HCl (pH 8.5). The 0.2 μg trypsin was used to digest 10 μg protein in 1 mM CaCl2 and 20 mM methylamine overnight at 37°C. The resultant peptides were desalted by ZipTip pipette tips (Merck Millipore). For liquid chromatography-tandem mass spectrometry analysis, peptides were separated by an EASY-nLCII integrated nano-HPLC system (Proxeon, Odense, Denmark) directly interfaced to a Thermo Scientific Q Exactive mass spectrometer (Thermo Fisher Scientific). Peptides were loaded on an analytical fused-silica capillary column (150 mm in length, 75 mm in internal diameter; Upchurch) packed with 5 mm, 300 Å C-18 resin (Varian). Mobile phase A consisted of 0.1% formic acid, while mobile phase B consisted of 0.1% formic acid and 100% acetonitrile. The Q Exactive mass spectrometer was operated through Xcalibur 2.1.2 software in data-dependent acquisition mode. A single full-scan mass spectrum was used in the orbitrap (400–1,800 m/z in mass range, 60,000 in resolution) followed by ten data-dependent tandem mass spectrometry scans at 27% normalized collision energy (HCD). The tandem mass spectrometry spectra results were searched against the C. elegans proteome database by the Proteome Discoverer (version PD1.4; Thermo Fisher Scientific).

Generation of C. elegans embryonic cell suspensions for SPLiT-Seq

We prepared single cells from a C. elegans strain (smIs89) that expressed a Pegl-1::NLS::GFP reporter using the established protocols with minor modifications(Bianchi and Driscoll, 2006; Zhang et al., 2011). Synchronized gravid adults were transferred to NGM plates to lay eggs for 1 hour, and about 50 μl eggs pellet were collected 3 hours post laid. We washed eggs three times with 1 ml sterile egg buffer [118 mM NaCl, 48 mM KCl, 3 mM CaCl2, 3 mM MgCl2, 5 mM HEPES (pH 7.2)]. We rinsed the pellet with 250 μl 1 U/ml Chitinase three times. Eggshells were digested with 250 μl of 1 U/ml chitinase for 45 min on a rotter. We stopped the reaction by adding 750 µl of L-15/FBS (L-15 medium containing 10% fetal bovine serum). Eggs were pelleted by centrifugation at 900 g for 3 min. The supernatant was displaced with 500 µl L-15/FBS. Embryos were dissociated by gently pipetting up and down 60 times using an insulin syringe (29 G needle). The dissociated cells were separated from undissociated cell clumps by filtering through a 5 µm mesh filter, and we washed the filter with 500 µl L-15/FBS. Cells were collected by centrifuging 1 ml filtered suspension at 3220 g for 3 min at 4°C. We removed the supernatant and gently resuspended cells with 2 ml 1X PBS containing 5 μl SUPERaseIn RNase Inhibitor (Invitrogen) and 2.5 μl RiboLock RNase Inhibitor (Thermo Scientific).

SPLiT-Seq library preparation and sequencing

We performed SPLiT-Seq as previously described (Rosenberg et al., 2018) with minor modifications. After cell fixation, centrifugations were performed at 3220 g for 3 min according to the C. elegans cell size. The RT primers were poly-dT and random hexamer mixture with UMIs (Unique Molecular Identifiers). All the barcodes used in this study were described in the published protocol(Rosenberg et al., 2018). The first 24 barcodes were used in the first round of barcoding, named “Round1_01” to “Round1_24”, and 48 barcodes were used in the third round of barcoding, called “Round3_49” to “Round3_96”. Each cell barcode came from three rounds of barcoding. TruePrep DNA Library Prep Kit V2 constructed the sequencing library for Illumina (Vazyme Biotech #TD502) according to the manufacturer’s instructions. The library was sequenced on HiSeq systems (Illumina) using 150 nucleotides (nts) kits and paired-end sequencing.

SPLiT-Seq data processing

According to the results of FastQC, adaptors or low-quality nucleotides were trimmed by Trim Galore (v0.5.2) using the default parameters. For each paired-end sequencing read, a 10-bp UMI sequence and a 24-bp cell barcode were extracted from the Read 2 file by the tool “preprocess_splitseq.pl” of zUMIs (v0.0.6). The Read 1 was split by different cell barcodes in Read 2 and mapped to the modified C. elegans genome (WS263 with an artificial chromosome containing the C. elegans version of the GFP sequence) by zUMIs (v0.0.6) and STAR (v2.6.0c). Unique mapping reads were kept. The duplicated reads from the same transcript were excluded based on the UMI information in Read 2. Low-quality transcriptomes were removed from the analysis, yielding 442 single-cell transcriptomes, including transcript counts from 8629 genes. Among these transcriptomes, 41 transcriptomes were assigned to the putative apoptotic cell based on the expression of egl-1 or gfp. Among the total 8624 genes detected in SPLiT-seq, transcripts from 6487 genes were not detected in the putative apoptotic cell population. Enriched GO terms of those 6487 genes were identified using Gene Ontology knowledgebase (Ashburner et al., 2000; Gene Ontology, 2021) (http://geneontology.org/).

Cell lineage tracing and quantification of reporter expression

Embryo mounting, live imaging, lineage tracing, and expression quantification were performed according to a previously described procedure with minor modifications (Bao and Murray, 2011; Du et al., 2014; Murray et al., 2008). Two- to four-cell stage embryos were collected from young adult worms and mounted between two coverslips in the egg buffer containing 20-30 20-μm polystyrene microspheres and sealed with melted Vaseline (Bao and Murray, 2011). 3D time-lapse imaging was performed at 20 °C ambient temperature using a spinning disk confocal microscope (Revolution XD) for 300-time points at a 75-second interval. For each time point, embryos were scanned for 30 Z focal planes with 1 µm spacing. 3D tiff stack images were processed with the StarryNite software for automated cell identification and tracing to reconstruct embryonic cell lineages using the ubiquitously–expressed mCherry fluorescence. The raw results of cell identification and tracing were subjected to extensive manual inspection and editing using the AceTree program to ensure high accuracy (Du et al., 2014). For each mentioned strain, the cell lineage was traced from the two- or four-cell stage to the 350-cell stage for three or four embryos (experimental replicates). Regions of the chromatids or nuclei are determined by the H2B::mCherry signal. The fluorescent intensity of HDA-1::GFP or LIN-53::mNeonGreen in each traced nucleus at each time point was measured as the average fluorescent intensity of all the pixels within each nucleus and then with the average fluorescent intensity of the local background subtracted (Murray et al., 2008).

RNA sequencing

Synchronized young adult worms were cultured on RNAi plates for 24 hours. Gravid adults were transferred to new RNAi plates to lay eggs for 1 hour. Eggs were harvested 5 hours post-laid and then lysed with TRIzol reagent (Invitrogen). We extracted the total RNA following the manufacturer’s protocol. The Qubit RNA High Sensitivity Assay Kit (Invitrogen) was used to quantify RNA concentration. The Agilent 2100 bioanalyzer system was used for the assessment of RNA quality. We used samples with an RNA integrity number (RIN) above 6.0 for sequencing library construction. We used identical input total RNA (50 ng to 500 ng) between control and samples for library preparation using the KAPA RNA HyperPrep Kit (KAPA Biosystems, Wilmington, MA, USA). Libraries were analyzed by Agilent 2100 bioanalyzer system for quality control. The library samples were sequenced on an Illumina NovaSeq 6000 platform. Approximately 5 GB of raw data were generated from each sample with 150 bp paired-end read lengths.

RNA-Seq data analysis

FastQC assessed the quality score, adaptor content, and duplication rates of sequencing reads. Trim_galore (v0.6.0) was used to remove the low-quality bases and adaptor sequences with default parameters. After trimming, paired-end reads with at least 20 nucleotides in length were aligned to the C. elegans reference genome (WS263) using STAR (2.5.4b) with default parameters. Read counts representing gene expression levels were calculated with HTSeq (version 0.9.1) using default parameters. Uniquely mapped reads were included. Gene names were annotated on the Wormbase website (https://wormbase.org/tools/mine/simplemine.cgi). DESeq2 package in R programming language was used for differential expression analysis. Differentially expressed genes were defined as upregulated genes (log2-transformed fold change greater than 1) or downregulated genes (log2-transformed fold change less than -1) with a false discovery rate (FDR) less than 0.05. We used the ggplot2 package of R to plot figures.

Reverse transcription and quantitative real-time PCR

Young adult worms were cultured on RNAi plates for 24 hours. We transferred the gravid adults to new RNAi plates to lay eggs for 1 hour. Eggs were harvested 5 hours post-laid and then lysed with TRIzol reagent (Invitrogen). The RNeasy Mini kit (QIAGEN) prepared the total RNA from ∼20 μl egg pellets for each biological replicate. cDNA from the same amount of RNA between RNAi and control samples were synthesized in a 20 μl reaction volume by the PrimeScript™ RT reagent Kit with gDNA Eraser (TAKARA, Code No. RR047A). One μl of cDNA was used as the template in a 10 μl reaction volume of the PowerUp SYBR Green Master Mix (Applied Biosystems) with four technical replicates. Quantitative real-time PCR was performed in two independent experiments with three biological replicates each time using the Applied Biosystems QuantStudio 1 Real-Time PCR System and normalized to the cell division cycle-related GTPase encoding gene cdc-42. Data were analyzed using the standard curve method. Primers used in real-time PCR assays were listed in Extended Data Table 3.

Chromatin immunoprecipitation

For ChIP assays, chromatin immunoprecipitation was performed as previously described(Mukhopadhyay et al., 2008) with modifications. Mixture stage worms on RNAi plates were harvested and washed three times with M9 buffer. 0.4 ml of packed worms were obtained and frozen into small balls with liquid nitrogen for each ChIP replicate. We quickly ground the little worm balls into powders in liquid nitrogen. Worm powder was crosslinked in 4 ml crosslinking buffer (1.1% formaldehyde in PBS with protease and phosphatase inhibitors) with constant rotation for 15 min at room temperature. Fixed worm samples were quenched by incubation with 0.125 M glycine for 5 min and then washed in cold PBS-PIC (PBS buffer with protease inhibitor cocktail) three times. The tissue pellet resuspended in PBS-PIC was homogenized by 2-3 strokes in a Dounce homogenizer. The cell suspension was centrifuged at 2,000 g for 5 min at 4°C, and we removed the supernatant.

We performed nuclei preparation, chromatin digestion, chromatin immunoprecipitation, and DNA Purification using SimpleChIP® Enzymatic Chromatin IP Kit (Magnetic Beads) (Cell Signaling Technology, #9003). Samples were sonicated using a Qsonica sonicator (Qsonica Q800R) at 70% amplitude for 5 cycles of 30 sec on and 30 sec off with 4 repeats. The chromatin was digested using 6 μl of micrococcal nuclease (MNase) (CST catalog number 10011S) in 300 μl of buffer B (CST catalog number 7007) containing 0.5 mM DTT for 8 min at 37 °C. The quality of chromatin digestion was analyzed by Agilent 2100 bioanalyzer system. Before IP, the chromatin sample concentration was assessed by the BCA Protein Assay Kit (Tiangen, cat no. PA115-02). 1 μg H3K27ac antibody (Abcam ab4729) or 2 μg IgG was used per IP with 0.5-1 mg input chromatin protein.

According to the manufacturer, we used 200 ng purified DNA samples to construct the ChIP–seq library using the VAHTS Universal DNA Library Prep Kit for Illumina V3 (Vazyme ND607). DNA libraries were sequenced on an Illumina NovaSeq 6000 platform, with 150-bp paired-end sequencing. Three biological replicates were collected for two independent ChIP–seq experiments. For ChIP–qPCR analysis of H3K27ac enrichment in the egl-1 locus, the same volume of cDNA from the IP experiment described above was used for qPCR analysis. Extended Data Table 3 listed the primers.

ChIP-Seq data processing

150-bp double-end reads were aligned to the C. elegans genome version WBcel235 using Bwa-mem2 version 2.1 with default parameters. Only uniquely mapped reads were kept. Samples were normalized by bins per million mapped reads (BPM) and averaged, and then treated samples were subtracted by input samples to calculate enrichment using deepTools2 (Ramirez et al., 2016). Data were visualized using IGV(Robinson et al., 2011).

Immunofluorescence

We performed immunofluorescence of C. elegans eggs using the freeze-cracking method. In brief, eggs were washed with M9 buffer three times and dripped onto poly-L-lysine coated slides. After freezing in liquid nitrogen for 5 minutes, the coverslip was swiftly removed, and embryos were fixed in methanol at -20 °C for 15 min. We washed the slides twice with PBST (PBS+ 0.05% Tween 20). We incubated slides with 1% BSA in PBST for 30 min at room temperature to block the unspecific bindings. The primary and secondary antibodies were diluted in PBST plus 1% BSA by 1:500-1:2,000 for primary antibodies (ab10799 for histone H3 and ab4729 for H3K27ac) and 1:5,000 for the secondary antibodies (Alexa-488 anti-mouse and Alexa546 anti-rabbit from Invitrogen). The primary antibody was incubated overnight at 4°C in a moisture chamber, and the secondary antibody was incubated for 1 hour at room temperature. After washing with PBST, we applied 1-2 drops of the Fluoroshield Mounting Medium containing DAPI (Abcam, ab104139) and coverslips to the samples.

Pharmacological and chemical treatments

To block V-ATPase proton pump activity in the L1-L2 larvae, which have cuticle layers that may compromise agent penetration, the Bafilomycin A1 treatment was designed based on the published protocols (Kumsta et al., 2017; Papandreou and Tavernarakis, 2017; Pivtoraiko et al., 2010). Briefly, gravid worms were placed in 1 drop of ddH2O containing 3% (v/v) pelleted OP50 and 100 μM bafilomycin A1 (Abcam, dissolved in DMSO) for 16 h at 20 ℃. After incubation, L1 larvae were washed with M9 buffer and recovered for 5 h on NG agar plates before imaging.

To achieve Q cell conditional HDA-1 depletion, the auxin-inducible degradation (AID) was performed as described previously (Zhang et al., 2015). The natural auxin (indole-3-acetic acid, IAA) was purchased from Alfa Aesar (#A10556). A 400 mM solution in ethanol was stored at 4°C as stock. Before imaging, L1 larvae were transferred to the S basal buffer supplemented with 3% (v/v) pelleted OP50 and 4 mM of auxin for a 1.5 h treatment.

Coimmunoprecipitation and Western blot

For immunoprecipitation of endogenous V-ATPase A subunit with HDA-1, HDA-1-GFP knock-in worms were lysed in ice-cold lysis buffer [pH 7.4, 150 mM NaCl, 25 mM Tris-HCl, 10% glycerol, 1% NP-40, 2x cocktail of protease inhibitors from Roche (Complete, EDTA free), 40 mM NaF, 5 mM Na3VO4]. The soluble fractions from worm lysates were immunoprecipitated with anti-GFP GFP-Trap A beads (Chromoteck, GTA20) for 1 hr at 4 °C. Immunoprecipitates were washed three times with lysis buffer containing 1x cocktail of protease inhibitors from Roche (Complete, EDTA free) and were then transferred to new tubes. Beads were washed again with 100 mM PB (pH 6.0, 6.5, 7.0, or 7.5) and boiled in 1X SDS loading buffer under 95 ℃ for 5min. The protein samples were analyzed by Western blotting.

Protein samples were resolved by SDS–PAGE and transferred to PVDF membranes. Each membrane was divided into two parts to be incubated with GFP (Abcam ab290, 1:4,000) or ATP6V1A (Abcam ab199326, 1:2,000) primary antibodies, followed by HRP-conjugated Goat Anti-Rabbit IgG (H&L) (Easybio, 1:5,000) secondary antibodies. Quantification of the endogenous V-ATPase A subunit interacting with HDA-1 was performed using HDA-1::GFP bands as a calibration standard. Quantitative densitometry of chemiluminescent bands was performed using ImageJ software.

Data availability

The data sets generated and analyzed in this study are available in the NCBI Gene Expression Omnibus (GEO, http://www.ncbi.nlm.nih.gov/geo) under accession number GSE167379 with the secure token uxwleayqnjkvnkf.

Supporting information

eLife Assessment

The authors propose that the asymmetric segregation of the NuRD complex in C. elegans is regulated in a V-ATPase-dependent manner, that this plays a crucial role in determining the differential expression of the apoptosis activator egl-1, and that it is therefore critical for the life/death fate decision in this species. If proven, the proposed model of the V-ATPase-NuRD-EGL-1-Apoptosis cascade would shed light onto the mechanisms underlying the regulation of apoptosis fate during asymmetric cell division, and stimulate further investigation into the intricate interplay between V-ATPase, NuRD, and epigenetic modifications. However, the strength of evidence for this is currently incomplete.

Joint Public Review:

Xie et al. propose that the asymmetric segregation of the NuRD complex is regulated in a V-ATPase-dependent manner, and plays a crucial role in determining the differential expression of the apoptosis activator egl-1 and thus critical for the life/death fate decision.

While the model is very intriguing, the reviewers raised concerns regarding the rigor of the method. One issue is with statistics (either insufficient information or inadequate use of statistics), and second is the concern that the asymmetry observed may be caused by one cell dying (resulting in protein degradation, RNA degradation etc). We recommend that the authors address these issues.

Major #1:

There are still many misleading statements/conclusions that are not rigorously tested or that are logically flawed. These issues must be thoroughly addressed for this manuscript to be solid.

1. Asymmetry detected by scRNA seq vs. imaging may not represent the same phenomenon, thus should not be discussed as two supporting pieces of evidence for the authors' model, and importantly each method has its own flaw. First, for scRNA seq, when cells become already egl-1 positive, those cells may be already dying, and thus NuRD complex's transcripts' asymmetry may not have any significance. The data presented in FigS1D, E show that there are lots of genes (6487 out of 8624) that are decreased in dying cells. Thus, it is not convincing to claim that NuRD asymmetry is regulated by differential RNA amount.

2. Regarding NuRD protein's asymmetry, there are still multiple issues. Most likely explanation of their asymmetry is purely daughter size asymmetry. Because one cell is much bigger than the other (3 times larger), NuRD components, which are not chromatin associated, would be inherited to the bigger cell 3 times more than the smaller daughter. Then, upon nuclear envelope reformation, NuRD components will enter the nucleus, and there will be 3 times more NuRD components in the bigger daughter cell. It is possible that this is actually the underling mechanism to regulate gene expression differentially, but this possibility is not properly acknowledged. Currently, the authors use chromatin associated protein (Mys-1) as 'symmetric control', but this is not necessarily a fair comparison. For NuRD asymmetry to be meaningful, an example of protein is needed that is non-chromatin associated in mitosis, distributed to daughter cells proportional to daughter cell size, and re-enter nucleus after nuclear envelope formation to show symmetric distribution. And if daughter size asymmetry is the cause of NuRD asymmetry, other lineages that do not undergo apoptosis but exhibit daughter size asymmetry would also show NuRD asymmetry. The authors should comment on this (if such examples exist, it is fine in that in those cell types, NuRD asymmetry may be used for differential gene expression, not necessarily to induce cell death, but such comparison provides the explanation for NuRD asymmetry, and puts the authors finding in a better context).

3. For the analysis of protein asymmetry between two daughters in Fig S4C, the method of calibration is unclear, making it difficult to interpret the results.

4. As for pHluorin experiments, the authors were asked to test the changes in fluorescence observed are due to changes in pH or changes in the amount of pHluorin protein. They need to add a ratio-metric method in this manuscript. A brief mention to Page 12 line 12 is insufficient to clarify this issue.

Major #2:

Some issues surrounding statistics must be resolved.

1. Fig. 1FG, 2D, 3BDEG, 5BD and 6B used either one-sample t-test or unpaired two-tailed parametric t-test for statistical comparison. These t-tests require a verification of each sample fitting to a normal distribution. The authors need to describe a statistical test used to verify a normal distribution of each sample.

2. Fig. 2D, 3D, and 3G have very small sample size (N=3-4, N=6, N=3, respectively), it is possible that a normal distribution cannot be verified. How can the authors justify the use of one-sample t-test and unpaired parametric t-test ?

3. Statistical comparison in Fig. 2D and Fig. 6B should be re-assessed. For Fig. 2D, the authors need to compare the intensity ratio of HDA-1/LIN53 between sister cells dying within 35 min and those over 400 min. For Fig. 6B, they need to compare the intensity ratio of VHA-17 between DMSO- and BafA1- treated cells at the same time point after anaphase.

Author response:

The following is the authors’ response to the current reviews.

Joint Public Review:

Xie et al. propose that the asymmetric segregation of the NuRD complex is regulated in a V-ATPase-dependent manner, and plays a crucial role in determining the differential expression of the apoptosis activator egl-1 and thus critical for the life/death fate decision.

Remaining concerns are the following:

The authors should provide the point-by-point response to the following issues. In particular, authors should provide clear reasoning as to why they did not address some of the following comments in the previous revisions. The next response should be directly answering to the following concerns.

(1) Discussion should be added regarding the criticism that NuRD asymmetric segregation is simply a result of daughter cell size asymmetry. It is perfectly fine that the NuRD asymmetry is due to the daughter cell size difference (still the nucleus within the bigger daughter would have more NuRD, which can determine the fate of daughter cells). Once the authors add this clarification, some criticisms about 'control' may become irrelevant.

We thank the reviewer for this suggestion. We will add the following text in the revised discussion on page 14, line 26:

“…We cannot rule out the possibility that NuRD asymmetric segregation results from daughter cell size asymmetry. According to this perspective, the nucleus in the larger daughter cell could possess more NuRD, potentially influencing the fate of the daughter cells. However, it is important to note that the nuclear protein histone or the MYST family histone acetyltransferase is equally segregated in daughter cells of different sizes.….”

(2) ZEN-4 is a kinesin that predominantly associates with the midzone microtubules and a midbody during mitosis. Given that midbodies can be asymmetrically inherited during cell division, ZEN-4 is not a good control for monitoring the inheritance of cytoplasmic proteins during asymmetric cell division. Other control proteins, such as a transcriptional factor that predominantly localizes in the cytoplasm during mitosis and enters into nucleus during interphase, are needed to clarify the concern.

We clarified the issue of ZEN-4 below:

The critique assumes that "midbodies can be asymmetrically inherited during cell division." However, this assumption does not apply to our study of Q cell asymmetric divisions. In our earlier research, we demonstrated that midbodies in Q cells are released post-division and subsequently engulfed by surrounding epithelial cells (Chai et al., Journal of Cell Biology, 2012). Moreover, we have shown that midbodies from the first cell division in C. elegans embryos are also released and engulfed by the P1 cell (Ou et al., Cell Research, 2013). Therefore, the notion of midbody asymmetric inheritance is irrelevant to this manuscript. Additionally, our manuscript already presents the example of the MYST family histone acetyltransferase, illustrating a nuclear protein that predominantly localizes in the cytoplasm during mitosis and symmetrically enters the nucleus during interphase.

As for pHluorin experiments, symmetric inheritance of GFP and mCherry is not an appropriate evidence to estimate the level of pHluorin during asymmmetric Q cell division. This issue remains unsolved.

We acknowledge the limitation of pHluorin in measuring the pH level in a living cell. Future studies could be performed to measure the dynamics of pH levels when advanced tools are available.

(3) Q-Q plot (quantile-quantile plot) in Figure S10 can be used for visually checking normality of the data, but it does not guarantee that the distribution of each sample is normal and has the standard deviation compared with the other samples. I recommend the authors to show the actual statistical comparison P-values for each case. The authors also need to show the number of replicate experiments for each figure panel.

We thank the reviewer for pointing this out. We will provide P-values for each case and the number of replicate experiments in the revised Figure 5-figure supplement 1 ( corresponding to Figure S10) and the figure legend.

The authors left inappropriate graphs in the revised manuscript. In Figure 3E, some error bars are disconnected and the other are stuck in the bars. In Figure S4C, LIN-53 in QR.a/p graph shows lines disconnected from error bars.

We thank the reviewer for pointing this out. We will correct these error bars.

I am bit confused with the error bars in Figure 2B. Each dot represents a fluorescent intensity ratio of either HDA-1 or LIN-53 between the two daughter cells in a single animal. Plots are shown with mean and SEM, but several samples (for example, the left end) exhibit the SEM error bar very close to a range of min and max. I might misunderstand this graph but am concerned that Figure 2B may contain some errors in representing these data sets. I would like to ask the authors to provide all values in a table format so that the reviewers could verify the statistical tests and graph representation.

We thank the reviewer for pointing this out. We apologize for the typo in Figure 2B figure legend. We will correct SEM to SD.

(4) The authors still do not provide evidence that the increase in sAnxV::GFP and Pegl-1gfp or the increase in H3K27ac at the egl-1 gene in hda-1(RNAi) and lin-53(RNAi) animals is not a consequence of global effects on development. Indeed, the images provided in Figure S7B demonstrate that there are global effects in these animals. no causal interactions have been demonstrated.

We cannot exclude the global effects and have discussed this issue in our previous manuscript on page 9, line 26:

“...Considering the pleiotropic phenotypes caused by loss of HDA-1, we cannot exclude the possibility that ectopic cell death might result from global changes in development, even though HDA-1 may directly contribute to the life-versus-death fate determination.”

(5) Figure 4: Due to the lack of appropriate controls for the co-IP experiment (Fig. 4), I remain unconvinced of the claim that the NuRD complex and V-ATPase specifically interact. Concerning the co-IP, the authors now mention that the co-IP was performed three times: "Assay was performed using three biological replicates. Three independent biological replicates of the experiment were conducted with similar results." However, the authors did not use ACT-4::GFP or GFP alone as controls for their co-IP as previously suggested. This is critical considering that the evidence for a specific HDA-1::GFP - V-ATPase interaction is rather weak (compare interactions between HDA-1::GFP and V-ATPase subunits in Fig 4B with those of HDA-1::GFP and subunits of NuRD in Fig S8B).

We conducted GFP pull-down experiments and MS spectrometric analysis for HDA-::GFP and ACT-4::GFP using identical protocols, yielding consistent results. We agree with the reviewer that in our Western blot, inclusion of ACT-4::GFP is a more effective negative control compared to empty beads.

(6) Based on Fig 5E, it appears that Bafilomycin treatment causes pleiotropic effects on animals (see differences in HDA-1::GFP signal in the three rows). The authors now state: "Although BafA1-mediated disruption of lysosomal pH homeostasis is recognized to elicit a wide array of intracellular abnormalities, we found no evidence of such pleiotropic effects at the organismal level with the dosage and duration of treatment employed in this study". However, the 'evidence' mentioned is not shown. It is critical that the authors provide this evidence.

We thank the Reviewer for pointing out this issue. We only checked the viability of the L1 larvae and morphology of animals at the organismal level with the BafA1 dosage and duration of treatment and did not notice any death of the animals and apparent abnormality in morphology (N > 20 for each treatment). However, as the reviewer pointed out, there can be some abnormalities at the cellular level. We thus revised this above description as the following, on page 11, line 27:

“…Although BafA1-mediated disruption of lysosomal pH homeostasis is recognized to elicit a wide array of intracellular abnormalities, we did not observe any larval deaths and apparent abnormality in morphology at the organismal level (N > 20 for each treatment) at the dose and duration of treatment employed in this study...”

The following is the authors’ response to the previous reviews.

eLife assessment

The authors propose that the asymmetric segregation of the NuRD complex in C. elegans is regulated in a V-ATPase-dependent manner, that this plays a crucial role in determining the differential expression of the apoptosis activator egl-1, and that it is therefore critical for the life/death fate decision in this species. If proven, the proposed model of the V-ATPase-NuRD-EGL-1-Apoptosis cascade would shed light onto the mechanisms underlying the regulation of apoptosis fate during asymmetric cell division, and stimulate further investigation into the intricate interplay between V-ATPase, NuRD, and epigenetic modifications. However, the strength of evidence for this is currently incomplete.

Public Review:

Xie et al. propose that the asymmetric segregation of the NuRD complex is regulated in a V-ATPase-dependent manner, and plays a crucial role in determining the differential expression of the apoptosis activator egl-1 and thus critical for the life/death fate decision.

While the model is very intriguing, the reviewers raised concerns regarding the rigor of the method. One issue is with statistics (either insufficient information or inadequate use of statistics), and second is the concern that the asymmetry observed may be caused by one cell dying (resulting in protein degradation, RNA degradation etc). We recommend that the authors address these issues.

We extend our sincere thanks to the Editors and Reviewers for their insightful comments on this study.

Major #1:

There are still many misleading statements/conclusions that are not rigorously tested or that are logically flawed. These issues must be thoroughly addressed for this manuscript to be solid.

(1) Asymmetry detected by scRNA seq vs. imaging may not represent the same phenomenon, thus should not be discussed as two supporting pieces of evidence for the authors' model, and importantly each method has its own flaw. First, for scRNA seq, when cells become already egl-1 positive, those cells may be already dying, and thus NuRD complex's transcripts' asymmetry may not have any significance. The data presented in FigS1D, E show that there are lots of genes (6487 out of 8624) that are decreased in dying cells. Thus, it is not convincing to claim that NuRD asymmetry is regulated by differential RNA amount.

We agree with the reviewer's comment. Indeed, scRNA-seq reveals phenomena different from those observed in protein imaging, and NuRD asymmetry may not be regulated by differential RNA levels. Seven years ago, when we started this project, NuRD asymmetry during asymmetric neuroblast division was unknown. We first found NuRD mRNA asymmetry using scRNA-seq and then NuRD protein asymmetry using fluorescence imaging. We have documented the whole process of discovering NuRD asymmetry, although the asymmetry of NuRD complex transcripts does not necessarily imply protein asymmetry. We have revised statements related to "NuRD asymmetry being regulated by differential RNA amounts" and discussed this issue in the revised manuscript on page 14, line 2:

" The transcript asymmetry detected by scRNA-seq may not correspond to the protein asymmetry detected by microscopic imaging. Our scRNA-seq data shows that 6487 out of 8624 genes were not detected in egl-1-positive cells, the putative apoptotic cells. Cells that are egl-1 positive may be undergoing apoptosis, rendering the asymmetry of NuRD complex transcripts insignificant in inferring protein asymmetry. Thus, the observed transcript asymmetry of the NuRD subunits between live and dead cells may be coincidental with NuRD protein asymmetry during asymmetric neuroblast division, rather than serving as a regulatory mechanism."

(2) Regarding NuRD protein's asymmetry, there are still multiple issues. Most likely explanation of their asymmetry is purely daughter size asymmetry. Because one cell is much bigger than the other (3 times larger), NuRD components, which are not chromatin associated, would be inherited to the bigger cell 3 times more than the smaller daughter. Then, upon nuclear envelope reformation, NuRD components will enter the nucleus, and there will be 3 times more NuRD components in the bigger daughter cell. It is possible that this is actually the underling mechanism to regulate gene expression differentially, but this possibility is not properly acknowledged. Currently, the authors use chromatin associated protein (Mys-1) as 'symmetric control', but this is not necessarily a fair comparison. For NuRD asymmetry to be meaningful, an example of protein is needed that is non-chromatin associated in mitosis, distributed to daughter cells proportional to daughter cell size, and re-enter nucleus after nuclear envelope formation to show symmetric distribution. And if daughter size asymmetry is the cause of NuRD asymmetry, other lineages that do not undergo apoptosis but exhibit daughter size asymmetry would also show NuRD asymmetry. The authors should comment on this (if such examples exist, it is fine in that in those cell types, NuRD asymmetry may be used for differential gene expression, not necessarily to induce cell death, but such comparison provides the explanation for NuRD asymmetry, and puts the authors finding in a better context).

For more than one decade, we have meticulously explored the relationship between protein asymmetry and cell size asymmetry during ACDs of Q cells. A notable example of even protein distribution is the cytokinetic kinesin ZEN-4, as documented in our 2012 publication in the Journal of Cell Biology (Chai et al., JCB, 2012). This study, primarily focusing on the fate of the midbody post-cell division, also showcased the dynamics of GFP-tagged ZEN-4 during ACDs of QR.a cells in movie S1. Intriguingly, beyond its role in the cytokinetic ring, we observed a uniform dispersal of ZEN-4 throughout the cytoplasm. Remarkably, following cell division, ZEN-4 transitions evenly into the nuclei of the daughter cells, a phenomenon with implications yet to be fully understood. One hypothesis is that ZEN-4's nuclear localization may prevent the formation of ectopic microtubule bundles in the cytosol during interphase. Below, we present a snapshot from our original movie, clearly showing the symmetrical distribution of ZEN-4 into the nuclei of the two daughter cells.

(3) For the analysis of protein asymmetry between two daughters in Fig S4C, the method of calibration is unclear, making it difficult to interpret the results.

In Figure S4C, we quantified the relative total fluorescence of the Q cell, with the quantification method illustrated in Figure S4A. To further clarify our quantification approach, we have updated Figure S4A and the "Live-Cell Imaging and Quantification" section in the Materials and Methods:

“…To determine the ratios of fluorescence intensities in the posterior to anterior half (P/A) of Q.a lineages or A/P of Q.p lineages, the cell in the mean intensity projection was divided into posterior and anterior halves. ImageJ software was used to measure the mean fluorescence intensities of two halves with background subtraction. The slide background's mean fluorescence intensity was measured in a region devoid of worm bodies. The background-subtracted mean fluorescence intensities of the two halves were divided to calculate the ratio. The same procedure was used to determine the fluorescence intensity ratios between two daughter cells. Total fluorescence intensity was the sum of the posterior and anterior fluorescence intensities or the sum of fluorescence intensities from two daughter cells (Figure S4A). …”

(4) As for pHluorin experiments, the authors were asked to test the changes in fluorescence observed are due to changes in pH or changes in the amount of pHluorin protein. They need to add a ratio-metric method in this manuscript. A brief mention to Page 12 line 12 is insufficient to clarify this issue.

We appreciate the concerns about potential changes in pH or pHluorin protein levels. While we cannot completely dismiss the impact of changes in the amount of pHluorin protein, it appears improbable that the asymmetry of pHluorin fluorescence is attributed to an asymmetric amount of pHluorin protein. This inference is supported by the observation that other fluorescent proteins, such as GFP or mCherry, did not exhibit any asymmetry during ACDs of Q cells. An example of GFP alone during the ACD of QL.p is illustrated in figure 5A from Ou and Vale, JCB, 2009. The fluorescence intensities in the large QL.pa cell and the small QL.aa are indistinguishable.

Major #2:

Some issues surrounding statistics must be resolved.

(1) Fig. 1FG, 2D, 3BDEG, 5BD and 6B used either one-sample t-test or unpaired two-tailed parametric t-test for statistical comparison. These t-tests require a verification of each sample fitting to a normal distribution. The authors need to describe a statistical test used to verify a normal distribution of each sample.

(2) Fig. 2D, 3D, and 3G have very small sample size (N=3-4, N=6, N=3, respectively), it is possible that a normal distribution cannot be verified. How can the authors justify the use of one-sample t-test and unpaired parametric t-test ?

(3) Statistical comparison in Fig. 2D and Fig. 6B should be re-assessed. For Fig. 2D, the authors need to compare the intensity ratio of HDA-1/LIN53 between sister cells dying within 35 min and those over 400 min. For Fig. 6B, they need to compare the intensity ratio of VHA-17 between DMSO- and BafA1- treated cells at the same time point after anaphase.

We appreciate the reviewer's advice on the statistical analysis of our data. In response, we performed normality tests on the datasets presented in Figures 1F, 1G, 3B, 5B, 5D, and 6B, all of which passed the tests (as demonstrated in Figure S10). We also acknowledge the reviewer's comment on the inadequate sample sizes in Figures 2D, 3D, 3E, and 3G for fitting a normal distribution. Therefore, we have revised our statistical analysis methods for these figures and updated both the figures and their legends. The revised statistical results support the primary conclusions of this study.

In response to the reviewer's observation regarding the small sample size in Figure 2D , which precluded normality verification, and the suggestion to compare sister cells that die within 35 minutes to those surviving over 400 minutes, we adapted our approach. We implemented the Kruskal-Wallis test to evaluate the differences among the groups. To assess the specific differences between each group and the 400 min MSpppaap group, we conducted the Dunn’s multiple comparisons test. The revised Figure 2D illustrates the updated statistical significance.

For Figure 3D, due to the small sample size precluding normality verification, we applied the Wilcoxon test with 1 as the theoretical median. The revised Figure 3D illustrates the updated statistical significance.

For Figure 3E, where the sample size also hindered normality verification, we conducted the Kruskal-Wallis test to evaluate the overall effect. Additionally, Dunn’s multiple comparisons test was utilized to examine the differences between groups. The revised Figure 3E illustrates the updated statistical significance.

For Figure 3G, the reviewer pointed out the small sample size and the limited statistical power due to having only three data points per group. To address this, we revised the figure to visually present each data point, aiming to more clearly illustrate the variation trends.

For Figure 6B, following the reviewer's suggestion, we compared the DMSO group directly with the Baf A1 group, updating Figure 6B to reflect this comparison as advised.

These adjustments have been made to ensure the statistical analyses are robust and appropriate given the sample sizes and to align with the reviewer's recommendations, enhancing the clarity and accuracy of our findings.

Recommendations for the authors:

We recommend using grey scale (instead of 'heatmap' representation) to show the protein distribution of interest. Heatmap does not help at all, because 'total protein amount per cell' (instead of signal intensity on each pixel) is what matters in the context of this paper. Heatmap presentation does not allow readers to integrate signal intensity with their eyes.

We thank the editor for pointing this out. We have changed heatmaps to inverted fluorescence images in grey scale.

Version 3

Abstract

Abstract

Asymmetric cell divisions (ACDs) generate two daughter cells with identical genetic information but distinct cell fates through epigenetic mechanisms. However, the process of partitioning different epigenetic information into daughter cells remains unclear. Here, we demonstrate that the nucleosome remodeling and deacetylase (NuRD) complex is asymmetrically segregated into the surviving daughter cell rather than the apoptotic one during ACDs in Caenorhabditis elegans. The absence of NuRD triggers apoptosis via the EGL-1-CED-9-CED-4-CED-3 pathway, while an ectopic gain of NuRD enables apoptotic daughter cells to survive. We identify the vacuolar H+– adenosine triphosphatase (V-ATPase) complex as a crucial regulator of NuRD’s asymmetric segregation. V-ATPase interacts with NuRD and is asymmetrically segregated into the surviving daughter cell. Inhibition of V-ATPase disrupts cytosolic pH asymmetry and NuRD asymmetry. We suggest that asymmetric segregation of V-ATPase may cause distinct acidification levels in the two daughter cells, enabling asymmetric epigenetic inheritance that specifies their respective life-versus-death fates.

Introduction

Asymmetric cell division (ACD) gives rise to two daughter cells that possess identical genetic material but distinct cell fates, playing a crucial role in both development and tissue homeostasis (Royall and Jessberger, 2021; Sunchu and Cabernard, 2020; Venkei and Yamashita, 2018; Wooten et al., 2020; Zion et al., 2020). Both extrinsic and intrinsic mechanisms determine distinct daughter cell fates after ACD. While extrinsic mechanisms, such as exposure to signaling cues from the local niche, have been extensively studied and are known to define stem cell fate (Morrison and Spradling, 2008), the intrinsic mechanisms are more complex and largely unresolved, despite several proteins, RNA molecules, and organelles having been implicated in the regulation of some types of ACD (Sunchu and Cabernard, 2020; Zion et al., 2020). Epigenetic mechanisms play a crucial role in guiding the two daughter cells toward establishing differential gene expression profiles, ultimately defining their unique cell identities (Allis and Jenuwein, 2016; Escobar et al., 2021; Stewart-Morgan et al., 2020). During the ACD of Drosophila male germline stem cells (GSCs), pre-existing and newly synthesized histones H3 and H4 are asymmetrically segregated towards the stem daughter cell and the differentiating daughter cell, respectively (Wooten et al., 2020). Nevertheless, the extent to which epigenetic information is asymmetrically inherited through ACD in other organisms and the mechanism by which this process occurs remains elusive.

Caenorhabditis elegans represents a valuable model for investigating ACD, given its invariant cell lineage and conserved mechanisms of ACD. During hermaphrodite development in C. elegans, 131 somatic cells undergo programmed cell death, with the majority produced through ACD that create a large cell programmed for survival and a small cell programmed to die (Sulston and Horvitz, 1977; Sulston et al., 1983a) The opposing cell fates of daughter cells, i.e., to live or die, offer a compelling experimental system for investigating how epigenetic inheritance determines life versus death decisions during ACD. It is noteworthy that 105 of the 131 apoptotic cells arise from neuronal lineages, of which the Q neuroblast represents a tractable system for studying ACD and apoptosis at single-cell resolution (Ellis and Horvitz, 1986; Hedgecock et al., 1983; Ou et al., 2010; Sulston and Horvitz, 1977). During the first larval stage, Q neuroblasts on the left (QL) and right (QR) undergo three rounds of asymmetric divisions, which produce three different neurons and two apoptotic cells (Q.aa and Q.pp), respectively (Figure 1A).

Asymmetric segregation of NuRD during ACDs of C. elegans Q neuroblast.

(A) Schematic of the Q neuroblast lineages. QL or QR neuroblast each generates three neurons and two apoptotic cells (Q.aa/Q.pp, X). QL produces PQR, PVM, and SDQL. QR produces AQR, AVM, and SDQR.

(B) A model of the NuRD complex composition (Bracken et al., 2019; Lai and Wade, 2011).

(C) Protein domain structure of the GFP-tagged HDAC1/2 (HDA-1) or mNeonGreen-tagged RBBP4/7 (LIN-53). CAF1C: histone-binding protein RBBP4 or subunit C of CAF1 complex; WD: WD40 repeat tryptophan-aspartate domain. Scale bar: 100 amino acids.

(D) Representative images of endogenous HDA-1::GFP, LIN-53::mNeonGreen and overexpressed MYS-1::GFP during ACDs of QR.a. In each panel, the top row shows merged images, the middle row shows mCherry-tagged plasma membrane and histone, and the bottom row shows inverted fluorescence images of GFP/mNeonGreen. The anterior of the cell is on the left. The GFP/mNeonGreen fluorescence intensity ratios between posterior and anterior chromatids, and between QR.ap and QR.aa nuclei, are shown in blue at the lower right corner of inverted fluorescence images. Other frames are in Figure S3A, and the full movies are in Supplementary Movie S3-S5. Scale bar: 2 µm.

(E) Quantification of HDA-1::GFP, LIN-53::mNeonGreen and MYS-1::GFP fluorescence intensity ratio between QR.ap and QR.aa nuclei. Data are presented as mean ± SEM. N = 10–12.

(F) Quantification of HDA-1::GFP (magenta), LIN-53::mNeonGreen (orange) and MYS-1::GFP (gray) fluorescence intensity ratios between the posterior and anterior half of QR.a or between QR.ap and QR.aa. Anaphase onset is defined as the last frame without chromatids segregation. Data are presented as mean ± SEM. N = 10–12. Statistical significance is determined by a one-sample t-test with 1 as the theoretical mean. ** p < 0.01, *** p < 0.001, ns: not significant.

(G) Quantification of HDA-1::GFP (left) and LIN-53::mNeonGreen (right) fluorescence intensity ratios between the large and small daughters of cells shown on the X-axis. Data are presented as mean ± SEM. N = 7–14. Statistical significance is determined by a one-sample t-test with 1 as the theoretical mean. ** p < 0.01, *** p < 0.001.

In the nematode, the classic apoptotic pathway initiates upon activation of the BH3-only protein EGL-1 in the cells that are fated to die. EGL-1 binds to the anti-apoptotic Bcl-2-like protein CED-9, which facilitates the release of the pro-apoptotic protein CED-4, leading to caspase CED-3 activation. Activated caspase promotes the exposure of phosphatidylserine (PS) on the surface of apoptotic cells, which triggers the phagocytosis of apoptotic cells via partially redundant signaling pathways such as the CED-1/CED-6/CED-7 pathway and the CED-2/CED-5/CED-12 pathway. Proper transcriptional regulation of the egl-1 gene is vital for life-versus-death decisions in C. elegans (Conradt and Horvitz, 1998; Nehme and Conradt, 2008). Although the cell-specific transcriptional regulator has been identified in certain cells, including the sexually dimorphic HSN and CEM neurons, M4 motor neuron, NSM neuron, P11.aaap cell, ABpl/rpppapp, and the tail spike cell (Hirose et al., 2010; Jiang and Wu, 2014; Nehme and Conradt, 2008), the general upstream regulators of egl-1 at the chromatin remodeling or epigenetic level are unclear. Two NSM neuroblasts undergo ACDs to each generate a larger NSM cell programmed to survive and a smaller NSM sister cell programmed to die (Sulston et al., 1983b). The PIG-1 kinase-dependent asymmetric partitioning of the Snail-like transcription factor, CES-1, represses egl-1 transcription in the larger NSM sister cells, thereby preventing their apoptosis (Hatzold and Conradt, 2008; Wei et al., 2020) However, the mechanism underlying the asymmetrical activation of egl-1 in other ACDs remains elusive.

In this study, we demonstrate the enrichment of the NuRD complex in cells that are predetermined to survive, and its role in suppressing the EGL-1-CED-9-CED-4-CED-3 apoptotic pathway through repression of the egl-1 gene. Furthermore, we report the interaction between the NuRD complex and the V-ATPase complex, and reveal that during asymmetric cell division, the localization of the V-ATPase complex and cytoplasmic pH are asymmetrical, thereby contributing to the polarized segregation of the NuRD complex.

Results

NuRD asymmetric segregation during neuroblast ACDs

In order to gain insights into the molecular distinctions between apoptotic and surviving cells in C. elegans embryos, we employed the SPLiT single-cell RNA sequencing (scRNA-seq) platform to compare their transcriptomes (Figure S1A-S1C) (Rosenberg et al., 2018). The expression of the critical somatic apoptosis-inducing gene egl-1 was used to distinguish the apoptotic and surviving cells (Conradt et al., 2016). Intriguingly, we identified transcripts encoding subunits of the nucleosome remodeling and deacetylase (NuRD; also known as Mi-2) complex in cells where egl-1 expression was indiscernible (Figure S1D-S1F, Supplementary Table 1). The NuRD complex represents an evolutionarily conserved protein complex associated with chromatin that couples the activities of chromatin-remodeling ATPases with histone deacetylases (Figure 1B) (Bracken et al., 2019; Lai and Wade, 2011). NuRD-mediated alterations in chromatin structure are vital for appropriate transcriptional regulation during cell fate determination and lineage commitment (Bracken et al., 2019; Lai and Wade, 2011), suggesting that NuRD may serve as a candidate epigenetic factor to specify cell fate between survival and death.

To investigate whether NuRD components are distributed asymmetrically between apoptotic and surviving daughter cells, we overexpressed GFP-tagged NuRD subunits within the C. elegans Q neuroblast lineages. Q.a neuroblast undergoes ACD to generate a large surviving Q.ap cell and a small apoptotic Q.aa cell (Figure 1A). We found that GFP-tagged NuRD subunits, including the histone deacetylase HDA-1, the histone binding protein LIN-53, the nucleosome remodeling factor CHD-3, and the MEP-1 protein were enriched asymmetrically in the surviving QR.ap cell during QR.a division (Figure S2, Supplementary Movie S1 and S2). Notably, NuRD asymmetric distribution occurred within several minutes after metaphase, while GFP protein translation and chromophore maturation require approximately 30 min in Q neuroblast (Ou and Vale, 2009), suggesting that the asymmetric enrichment of NuRD is likely the result of protein redistribution.

To investigate the dynamic distribution of endogenous NuRD during ACD, we generated a GFP knock-in (KI) strain for HDA-1 and an mNeonGreen (NG, green fluorescence) KI line for LIN-53 using CRISPR-Cas9 (Figure 1C). In QR.a cells, nuclear HDA-1 and LIN-53 were released into the cytoplasm and evenly distributed until metaphase (Figure 1D, Supplementary Movie S3 and S4). At anaphase, HDA-1 and LIN-53 became enriched in the posterior part of QR.a but became less detectable in the anterior part (Figure 1D, S3A-B, Supplementary Movie S3 and S4). We quantified the fluorescence intensity ratio between the posterior and anterior chromatids of QR.a, and between QR.ap and QR.aa nuclei, and found that nuclear HDA-1 or LIN-53 asymmetry gradually increased from 1.1-fold at anaphase onset to 1.5 or 1.8-fold upon completion of cytokinesis, respectively (Figure 1D-E). We also measured the ratios of fluorescence intensities between the posterior and anterior halves of QR.a, and between QR.ap and QR.aa (Figure S4A, see Materials and Methods). NuRD asymmetry became evident at ∼4 minutes and reached a plateau at ∼6 minutes after the anaphase onset (Figure 1D, 1F, S3A and S4B). QR.a spent ∼6 minutes from anaphase to the completion of cytokinesis (Chai et al., 2012; Ou et al., 2010), suggesting that QR.a cell establishes NuRD asymmetry during ACD.

Similar to QR.a , the dynamic and progressively developing NuRD asymmetry was also observed in the QL.a, QR.p and QL.p cell lineages, which generate apoptotic cells (Figure 1G, S3B and S4B). In contrast, an even distribution of HDA-1 and LIN-53 was observed in two surviving daughters of QR, QL, QL.pa and QR.pa cells, which generate two viable siblings (Figure 1G and S3B). To investigate whether other epigenetic factors are also asymmetrically segregated during Q cell ACDs, we tagged the MYST family histone acetyltransferase (MYS-1 in C. elegans) with GFP. We found that MYS-1::GFP was symmetrically segregated into the apoptotic and surviving daughter cells (Figure 1D-F, S3A and S4B, Supplementary Movie S5), indicating the specificity of the polarized NuRD partition. The total fluorescence of HDA-1, LIN-53, and MYS-1 remained constant during ACDs, suggesting that protein redistribution may establish NuRD asymmetry (Figure S4C). Inhibition of a PAR-1 family kinase PIG-1 led to symmetric Q cell divisions, producing extra neurons derived from some of their apoptotic daughters (Cordes et al., 2006). In pig-1 (gm344) mutants, HDA-1 and LIN-53 were evenly partitioned during ACDs (Figure S5, Supplementary Movie S6 and S7), indicating that asymmetric NuRD segregation depends on the PIG-1 kinase.

NuRD asymmetric segregation in embryonic cell lineages

Given that a significant portion of somatic apoptotic events in C. elegans-113 out of 131-occur during embryonic development (Sulston et al., 1983a), we employed live imaging and automated cell lineage tracing algorithms (Du et al., 2014) to monitor the asymmetry of NuRD distribution in embryos from the early two- or four-cell stage up to the 350-cell stage. Our analysis indicated that in 15 out of 17 embryonic ACDs that produce apoptotic daughter cells, the surviving daughter cells showed enrichment of HDA-1 and LIN-53, with an average enrichment ratio exceeding 1.5-fold (Figure 2A and Supplementary Table 2). Notably, this enrichment reached statistical significance in 6 out of 17 embryonic ACDs (Figure 2B). Intriguingly, the MSpppaa cell lineage’s two daughter cells received comparable levels of NuRD, yet its apoptotic daughter cell completes apoptosis approximately 400 minutes post-division (Hsieh et al., 2012; Sulston et al., 1983b), supporting the hypothesis that lower NuRD concentrations might promote apoptosis. Conversely, the apoptotic daughter cell of ABaraaaap completes apoptosis within 35 minutes (Sulston et al., 1983a), despite receiving a similar amount of NuRD as its non-apoptotic sibling. Therefore, while the regulation of apoptosis in certain embryonic cells appears to be governed by NuRD-independent mechanisms, NuRD asymmetric segregation is evident in several embryonic cell lineages.

NuRD Asymmetry in C. elegans embryonic cell lineages.

(A) The tree visualization depicts the segregation of HDA-1::GFP and LIN-53::mNeonGreen between sister cells during embryonic development. In this tree structure, vertical lines represent cells and horizontal lines denote cell divisions. Green vertical lines highlight cells with higher nuclear HDA-1::GFP and LIN-53::mNeonGreen fluorescence intensity than their apoptotic sister cells (average fluorescence intensity ratio between sister cell nuclei >1.5). Red arrowheads point to apoptotic cells. The placement of cells within the tree follows the Sulston nomenclature. See also Supplementary Table 2.

(B) Quantifications of HDA-1::GFP and LIN-53::mNeonGreen fluorescence intensity ratios between nuclei of live daughter cells and their apoptotic sister cells. The lineage names of 17 cells that divide to produce apoptotic daughter cells are shown below the X-axis. Data are shown as mean ± SEM. N = 3–4. The Dunn’s multiple comparisons test was used to assess statistical significance, with MSpppaa (magenta), whose apoptotic daughter cell completes apoptosis over 400 min after birth, as a control. *p < 0.05, **p < 0.01, ***p < 0.001.

Loss of the deacetylation activity of NuRD causes ectopic apoptosis

To investigate the role of NuRD in determining cell fate, we reduced the deacetylation activity of NuRD using RNA-mediated interference (RNAi). The hda-1 and lin-53 RNAi efficacy was confirmed by the reduced green fluorescence in the germlines of HDA-1::GFP and LIN-53::mNeonGreen KI animals, as well as the increased acetylation level of histone H3K27 (Figure S6). As reported before, RNAi of either hda-1 or lin-53 caused embryonic lethality (Shi and Mello, 1998). RNAi of lin-53 led to embryos arrested at the gastrulation stage before the generation of apoptotic cells, preventing us from analyzing its roles in somatic apoptosis. In contrast, hda-1 RNAi embryos were arrested between the late gastrulation stage and the bean stage, allowing us to track some of the apoptotic events. To quantify apoptotic events, we used a secreted Annexin V (sAnxV::GFP) sensor to label apoptotic cells with externalized phosphatidylserine (exPS) on the surface of the plasma membrane in ced-1(e1735) mutant embryos which inhibits engulfment of apoptotic cells (Conradt et al., 2016; Mapes et al., 2012; Zhou et al., 2001). The sAnxV::GFP labeled an average of 10 cells in ced-1(e1735) mutant embryos treated with control RNAi and an average of 14 cells in ced-1(e1735); hda-1 RNAi embryos (Figure 3A-B). To test whether the upregulated apoptosis in hda-1 RNAi embryos depends on the canonic apoptosis pathway, we introduced cell death-deficient mutations ced-3/ Caspase (n2433), ced-4/ APAF-1-like (n1162), ced-9/ BCL-2-like (n1950), and egl-1/ BH-3 only protein (n1084n3082) into ced-1 mutant, respectively. RNAi of hda-1 in these embryos only generated 0-2 exPS-positive cells (Figure 3A-B and S7A), indicating that the ectopic increase of apoptosis by HDA-1 inhibition requires the EGL-1-CED-9-CED-4-CED-3 pathway. Considering the pleiotropic phenotypes caused by the loss of HDA-1, we cannot exclude the possibility that ectopic cell death might result from global changes in development, even though HDA-1 may directly contribute to the life-versus-death fate determination.

RNAi of hda-1 induces ectopic apoptosis and increases H3K27 acetylation of the egl-1 gene.

(A) Representative inverted fluorescence images show Phsp::sAnxV::GFP from ced-1(e1735) and ced-1(e1735); ced-3(n2433) embryos between late gastrulation and bean stage, treated with control RNAi or hda-1 RNAi. Scale bars, 5 μm.

(B) Quantifications of cell corpse number in the embryos of indicated genotypes. Data are presented as mean ± SEM. N = 39–60. Statistical significance is determined by Student’s t test. ***p < 0.001.

(C) Scatter plot of the increased gene expression in both hda-1 RNAi and lin-53 RNAi animals. The egl-1 gene is marked in red. See also Supplementary Table 3.

(D) Quantitative real-time PCR (RT-PCR) measurement of egl-1 mRNA levels in the control, hda-1 or lin-53 RNAi embryos. Fold induction was calculated relative to levels in control RNAi embryos. Data of six biological replicates are presented as mean ± SEM. Statistical significance is determined by the Wilcoxon test as 1 as the theoretical median. *p < 0.05.

(E) Quantification of the number of cells expressing the Pegl-1::NLS::GFP reporter in embryos treated with control, hda-1 or lin-53 RNAi. Data are presented as mean ± SEM. N = 8–9. Statistical significance is determined by the Dunn’s multiple comparisons test.

*p < 0.05, ***p < 0.001, ns: not significant.

(F) Normalized ChIP-Seq signal profiles of the H3K27 acetylation level in the control, hda-1 or lin-53 RNAi embryos at the egl-1 locus. The Y-axis shows the average sequencing coverage of bins per million reads of three biological replicates normalized to the input control sample. Ectopic H3K27ac enrichments at the egl-1 locus of hda-1 RNAi and lin-53 RNAi embryos are highlighted in magenta and cyan.

(G) ChIP-qPCR analyses using H3K27ac antibody or IgG (the negative control) at selected elements of egl-1 indicated in (F) by red lines. Results are shown as the percentage of input DNA. Data from three biological replicates are presented.

NuRD RNAi upregulates the egl-1 expression by increasing its H3K27 acetylation

To understand how NuRD regulates apoptotic cell fates, we performed RNA-Seq analyses on WT, hda-1 RNAi and lin-53 RNAi embryos. In both RNAi conditions, we observed 287 or 575 genes that were upregulated or downregulated (|log2FD| > 2 and FDR < 0.05) (Supplementary Table 3). Notably, we found that the apoptosis-inducing gene egl-1 was among the upregulated genes (Figure 3C). Our quantitative RT-PCR results further confirmed that egl-1 transcripts were more abundant in hda-1 RNAi and lin-53 RNAi embryos than in control RNAi embryos (Figure 3D). We also used a Pegl-1::NLS::GFP transcriptional reporter to examine the egl-1 transcription activity during embryonic development, which allowed us to detect egl-1 expression while avoiding the adverse effects of egl-1 overexpression on apoptosis. We found that RNAi of hda-1 or lin-53 caused an increased number of Pegl-1::NLS::GFP positive cells (Figure 3E and S7B). These results indicate that the loss of NuRD abnormally activates the egl-1 gene expression.

Next, we investigated whether the loss of histone deacetylase activity in NuRD increased the H3K27ac levels at the egl-1 locus, as the histone H3K27ac levels are associated with transcription activation (Heinz et al., 2015). To test this, we performed ChIP-Seq experiments with the anti-H3K27ac antibody and found that the transcription start site (TSS) and the gene body region of egl-1 had higher levels of H3K27ac in hda-1 or lin-53 RNAi embryos compared to WT, while the H3K27ac levels of genes adjacent to egl-1 showed no significant changes (Figure 3F). This was supported by ChIP-qPCR in hda-1 RNAi embryos (Figure 3G). Therefore, inhibiting NuRD increased the H3K27ac levels at the egl-1 locus, leading to its upregulation and subsequent activation of apoptosis.

V-ATPase regulates asymmetric segregation of NuRD during somatic ACDs

To investigate how NuRD is asymmetrically segregated during ACD, we used affinity purification with an anti-GFP antibody and mass spectrometry (MS) to isolate NuRD binding partners from the lysate of HDA-1::GFP KI animals. Our analysis revealed all the previously known NuRD subunits from HDA-1::GFP KI animals (Figure S8, Supplementary Table 4) (Bracken et al., 2019; Lai and Wade, 2011), validating our experimental method. Interestingly, our co-immunoprecipitation and MS with HDA-1::GFP identified 12 subunits of the vacuolar-type H+-ATPase (V-ATPase) (Figure 4A-B), consistent with a previous report that used FLAG-tagged LIN-53 to identify 4 V-ATPase subunits (Muthel et al., 2019). We further confirmed the binding of HDA-1 to V-ATPase subunits in C. elegans by co-immunoprecipitation and western blot analysis with anti-V1A/VHA-13 antibodies (Figure 4C). V-ATPase is a proton pump that comprises a transmembrane domain (V0) responsible for proton transport across membranes and a peripheral ATP-hydrolytic domain (V1) that catalyzes ATP hydrolysis (Figure 4B). V-ATPases play a critical role in intracellular pH homeostasis and regulate numerous physiological processes, including apoptosis (Abbas et al., 2020; Chen et al., 2022; Ernstrom et al., 2012; Vasanthakumar and Rubinstein, 2020). However, the role of V-ATPase in the asymmetric segregation of epigenetic factors during ACD remains unknown.

HAD-1 interacts with subunits of V-ATPase.

(A) The plot compares counts of proteins co-precipitated with HDA-1::GFP with those with the control ACT-4 (actin)::GFP. The PSM (Peptide-Spectrum Match) is the number of identified peptide spectra matched for the protein. Blue dots represent the subunits of V-ATPase. See also Supplementary Table 4.

(B) Schematic model of the V-ATPase complex. The known worm subunits are indicated. The mean PSM of the encoded protein from two co-IP and MS repeats is shown in blue after the gene name.

(C) Western blot (WB) showing co-immunoprecipitation (co-IP) of V-ATPase V1 domain A subunit (V1A) with the HDA-1 from worm lysates. Assay was performed using three biological replicates. Three independent biological replicates of the experiment were conducted with similar results

V-ATPase regulates NuRD asymmetric segregation and cell fates.

(A) Dynamics of the cytosolic pH indicated by super-ecliptic pHluorin during QR.a division in DMSO- or BafA1-treated animals. In each panel, the top row shows mCherry-tagged plasma membrane and histone, and the bottom row shows inverted fluorescence images of super-ecliptic pHluorin. The anterior of the cell is on the left. Time 0 min is the onset of anaphase. Scale bar, 2 µm. See also Supplementary Movie S8.

(B) The super-ecliptic pHluorin fluorescence intensity ratio between QR.ap and QR.aa in DMSO control or BafA1-treated animals. Data are presented as mean ± SEM. N = 11–15. Statistical significance is determined by Student’s t test. **p < 0.01.

(C) Images of HDA-1::GFP distribution during QR.a cell division in DMSO- or BafA1-treated animals. In each panel, the top row shows mCherry-tagged plasma membrane and histone, and the bottom row shows inverted fluorescence images of HDA-1::GFP. The GFP fluorescence intensity ratios between the posterior and anterior chromatids, and between QR.ap and QR.aa nuclei are shown in blue at the lower left corner of inverted fluorescence images. Anterior of the cell is left. Scale bar: 2 µm. See also Supplementary Movie S9.

(D) Quantification of HDA-1::GFP fluorescence intensity ratios between the large and small daughter cell nuclei and the cell size ratios between the large and small daughters. The names of mother cells are shown on the X-axis. Data are presented as mean ± SEM. N = 9–12. Statistical significance is determined by Student’s t test. **p < 0.01, ns: not significant.

(E) Representative images showing fates of QR.aa after DMSO (top), BafA1 (middle) and BafA1 plus AID treatment (bottom). The Q cell plasma membrane and chromosome are labeled by mCherry. HDA-1::GFP is shown as inverted fluorescence images. Yellow arrowhead shows a short neurite-like outgrowth of QR.aa in BafA1-treated larvae. Frequencies of QR.aa survival and HDA-1 maintenance are shown on the right. Time 0 min is the birth of QR.aa. Scale bar: 5 µm.

To investigate the role of V-ATPase in NuRD asymmetric segregation in Q neuroblast, we used a pharmacologic inhibitor, bafilomycin A1 (BafA1), to inhibit V-ATPase proton pumping activity (Furuchi et al., 1993; Wang et al., 2021; Yoshimori et al., 1991). To assess the inhibitory effect of BafA1 on the proton translocation activity of V-ATPase in Q neuroblast, we monitored the cytosolic pH dynamics in dividing QR.a using the super-ecliptic pHluorin. The pHluorin is a pH-sensitive GFP reporter whose fluorescence can be quenched by the acidic pH (Miesenbock et al., 1998; Sankaranarayanan et al., 2000). Although BafA1-mediated disruption of lysosomal pH homeostasis is recognized to elicit a wide array of intracellular abnormalities, we found no evidence of such pleiotropic effects at the organismal level with the dosage and duration of treatment employed in this study. In DMSO-treated animals, the pHlourin fluorescence intensity remained constant in the posterior portion that forms QR.ap but was significantly reduced in the anterior portion that forms QR.aa, indicating that the cytoplasm of the future apoptotic cell became more acidic (Figure 5A-B and Supplementary Movie S8). This observation is consistent with the recognition that cytosolic acidification is a common feature of both death receptor-mediated and mitochondria-dependent apoptosis (Lagadic-Gossmann et al., 2004; Matsuyama et al., 2000; Matsuyama and Reed, 2000). Notably, BafA1 treatment reduced the pHluorin fluorescence intensity ratio between QR.ap and QR.aa from 1.6-fold to 1.3-fold (Figure 5A-B and Supplementary Movie S8), suggesting that BafA1 may disrupt the cytosolic pH asymmetry in dividing QR.a cells by inhibiting V-ATPase activity, although we cannot exclude the possibility that the changes in fluorescence could be due to changes in the amount of pHluorin protein.

We observed that BafA1 treatment resulted in the symmetric segregation of HDA-1 during QR.a division (Figure 5C-D and Supplementary Movie S9). Notably, neither DMSO nor BafA1 treatment affected the asymmetry in daughter cell size (Figure 5C-D and Supplementary Movie S9), suggesting that intracellular acidification, regulated by V-ATPase activity, specifically affects the NuRD asymmetric segregation without affecting the asymmetry in daughter cell size. We also investigated whether the small QR.aa cell carrying ectopic NuRD could escape apoptosis. In DMSO-treated animals, QR.aa cells underwent apoptosis, and 12 out of 13 QR.aa cells were engulfed and degraded by a neighboring epithelial cell within 120 min after birth. However, in BafA1-treated animals, QR.aa inherited similar levels of HDA-1::GFP as its sister cell, and 11 out of 12 QR.aa cells carrying ectopic NuRD survived for over 120 min and formed a short neurite-like outgrowth (Figure 5E). To confirm whether the long-lived QR.aa was a consequence of the ectopic gain of NuRD, we depleted HDA-1 within Q cell lineages using the auxin-inducible protein degradation (AID) system (Zhang et al., 2015). By introducing the degron sequence into the hda-1 locus and expressing the TIR1 F-box protein under the Q cell-specific egl-17 promoter, we showed that HDA-1::GFP fluorescence in Q cells was substantially decreased under auxin treatment (Figure 5E). Administration of auxin restored QR.aa apoptosis in BafA1-treated animals (Figure 5E). These results suggest that V-ATPase activity-dependent NuRD asymmetric segregation contributes to the specification of the live-death fate.

To understand how the V-ATPase regulates NuRD asymmetric segregation, we generated a transgenic strain expressing wrmScarlet-tagged VHA-17, which is the e subunit of the V0 domain. Using this strain, we were able to examine the dynamic distribution of V-ATPase during QR.a cell division. Like HDA-1, V-ATPase was also asymmetrically enriched in the surviving QR.ap portion (Figure 6A-B and Supplementary Movie S10). This observation suggests that V-ATPase and NuRD may co-segregate asymmetrically during ACD. We also found that BafA1 treatment disrupted V-ATPase asymmetric distribution (Figure 6A-B and Supplementary Movie S10), indicating the importance of the proton-pumping activity of V-ATPase in its asymmetric segregation. Therefore, our results suggest that V-ATPase may facilitate the asymmetric distribution of NuRD through its proton pumping activity.

V-ATPase distribution during ACDs and a model.

(A) Dynamics of VHA-17::wrmScarlet during QR.a cell division in DMSO- or BafA1-treated animals. VHA-17::wrmScarlet fluorescence is shown as inverted fluorescence images. Time 0 min is the onset of anaphase. Scale bar: 2 µm. See also Supplementary Movie S10.

(B) Quantification of VHA-17::wrmScarlet (magenta) and HDA-1::GFP (blue) fluorescence intensity ratios between the posterior and anterior half of QR.a or between QR.ap and QR.aa. Data are presented as mean ± SEM. N = 10–14. Student’s t test was used to compare the intensity ratio difference of VHA-17 and HDA-1 between DMSO- and BafA1-treated cells at 6 min after anaphase. **p < 0.01, ***p < 0.001.

(C) A proposed model. Asymmetric segregation of V-ATPase mediated the enrichment of its associated NuRD in the large daughter cell, where a high level of NuRD deacetylates egl-1 and suppresses its expression.

Discussion

Our findings indicate that the asymmetric segregation of the NuRD complex during ACD is regulated in a V-ATPase-dependent manner, and is critical for the differential expression of apoptosis activator egl-1 and the life-versus-death fate decision (Figure 6C). Specifically, we have provided evidence that the reduced level of NuRD in the apoptotic daughter cell leads to an increased level of H3K27ac, an epigenetic modification known to be associated with active gene expression, at the egl-1 locus, resulting in the upregulation of egl-1 expression and subsequent induction of apoptosis. Based on our results, we propose a dual-function model of V-ATPase in NuRD asymmetric segregation: (i) V-ATPase interacts with NuRD, both of which are asymmetrically segregated into the surviving cell; (ii) the asymmetrical activity of V-ATPase generates a more acidic cytoplasmic environment in the future apoptotic cell portion compared to the surviving portion, which might contribute to the asymmetric segregation of V-ATPase and NuRD. This model sheds light on the mechanisms underlying the regulation of apoptosis during ACD, and opens up new avenues for further investigation into the intricate interplay between V-ATPase, NuRD, and epigenetic modifications in the context of cell fate decisions.

The transcript asymmetry detected by scRNA-seq may not correspond to the protein asymmetry detected by microscopic imaging. Our scRNA-seq data shows that 6487 out of 8624 genes were not detected in egl-1-positive cells, the putative apoptotic cells. Cells that are egl-1 positive may be undergoing apoptosis, rendering the asymmetry of NuRD complex transcripts insignificant in inferring protein asymmetry. Thus, the observed transcript asymmetry of the NuRD subunits between live and dead cells may be coincidental with NuRD protein asymmetry during asymmetric neuroblast division, rather than serving as a regulatory mechanism.

The intrinsic mechanisms governing binary cell fate decisions involve asymmetric cortical localization of cell fate determinants, polarized partitioning of RNA species, unequal segregation of organelles, and biased distribution of damaged proteins or protein aggregates (Sunchu and Cabernard, 2020; Venkei and Yamashita, 2018). Despite this, little is known about the asymmetric segregation of epigenetic modification enzymes during ACDs, nor the functions of polarized V-ATPase distribution and cytoplasmic proton asymmetry during this process. Our findings provide new insights into the asymmetric segregation of cell-intrinsic factors and the previously unrecognized roles of V-ATPase and cytosolic acidification in this process.

Two mechanisms have been proposed to explain the V-ATPase-dependent asymmetric segregation of NuRD. Firstly, a polarized intracellular transportation system could selectively deliver NuRD-V-ATPase-containing organelles to the surviving daughter cell. Although V-ATPase is primarily known for its localization and function in the late endosome and lysosome, recent evidence suggests that V-ATPase subunits are synthesized and assembled on the endoplasmic reticulum (ER) (Abbas et al., 2020; Graham et al., 2003; Wang et al., 2020). Our findings demonstrate that V-ATPase colocalizes with the ER marker during ACDs, suggesting that NuRD-V-ATPase may be delivered as cargo on the ER or ER-derived vesicles (Figure S9). As reported in previous studies, a polarized microtubule transportation system, governed by specific motor proteins and the microtubule tracks, is responsible for the asymmetric segregation of signaling endosomes during ACDs of Drosophila sensory organ precursors (Derivery et al., 2015). A similar biased transportation system may play a role in the asymmetric segregation of NuRD-V-ATPase-containing organelles.

Secondly, the surviving daughter cell chromatids may have a greater ability to recruit NuRD than their apoptotic sister. This could be due to a higher number of yet unknown NuRD recruiting factors on chromatids or specific posttranslational modifications of chromosomal proteins that enhance the chromosomal recruitment of NuRD. Despite having an identical DNA sequence, sister chromatids can be distinguished by chromosomal proteins or modifications (Ranjan et al., 2019; Tran et al., 2012), which likely provides the specific molecular cues for asymmetric NuRD recruitment. The NuRD asymmetry on chromosomes (1.5∼1.8-fold between QR.ap and QR.aa nuclei) is greater than that between the two daughter cells (approximately 1.3-fold) (Figure 1D-1F), suggesting that polarized NuRD-V-ATPase transportation and distinct NuRD affinity from sister chromatids may act in concert to establish NuRD asymmetry.

The observation of asymmetric segregation of the NuRD complex during cell divisions suggests that the mother cell may play a role in initiating the fate specification of its daughter cells. This notion is supported by studies conducted on Drosophila male germline stem cells (GSCs) which demonstrated that the old and new histone H3-H4 or unequal amounts of histone H3 variant CENP-A were incorporated into sister chromatids prior to cell division (Ranjan et al., 2019; Tran et al., 2012). Although these patterns were not observed in cell types other than GSCs, asymmetric NuRD segregation was found to be common in C. elegans (Figure 1D-G, Figure 2A, and Figure S3). In 395 tracked embryonic cell divisions that generated two surviving daughter cells, NuRD was equally distributed in 390 cases, but in five cases, significant differences in HDA-1 and LIN-53 levels were observed between siblings (Figure 2A, Supplementary Table 2). This observation suggests that biased NuRD segregation may contribute to binary cell fate decisions beyond apoptosis. In the embryonic cerebral cortex of mammals, approximately 30% of newborn cells undergo programmed cell death (Voiculescu et al., 2000), which is a significantly higher rate than observed in C. elegans where only 12% of somatically born cells undergo apoptosis (131 apoptotic cells among 1090 somatically born cells). Thus, it is worth exploring whether brain stem cell divisions employ asymmetric NuRD segregation to specify daughter cell fates. Further studies on polarity establishment hold promise in elucidating the mechanisms underlying the asymmetric inheritance of epigenetic information.

Materials and Methods

Worm strains and culture

C. elegans were maintained as described in the standard protocol (Brenner, 1974). All strains were cultivated at 20 °C on nematode growth medium (NGM) agar plates seeded with Escherichia coli OP50 or HT115 (feeding RNAi assay). The wild-type strain was Bristol N2. Some strains were provided by the Caenorhabditis Genetics Center (CGC), funded by the NIH Office of Research Infrastructure Programs (P40 OD010440). The strains used in this study are listed in Extended Data Table 1.

Molecular biology

We performed genome editing experiments in C. elegans following the established methods (Dickinson et al., 2013; Friedland et al., 2013). We used the CRISPR design tool (https://zlab.bio/guide-design-resources) to select the target sites. The sgRNA sequences (Extended Data Table 2) were inserted into the pDD162 vector (Addgene #47549) by linearizing the vector with 15 base pairs (bp) overlapped primers using Phusion high-fidelity DNA polymerase (New England Biolabs, cat. no. MO531L). PCR products were digested using Dpn I (Takara, cat. no. 1235A) for 2 hours at 37°C and transformed into Trans5𝛼 bacterial chemically competent cells (TransGen Biotech, Cat no. CD201-01). The linearized PCR products with 15 bp overlapping ends were cyclized to generate plasmids by spontaneous recombination in bacteria. Homology recombination (HR) templates were constructed by cloning the 1.5 kb upstream and downstream homology arms into the pPD95.77 plasmid using the In-Fusion Advantage PCR cloning kit (Clontech, Cat. #639621). Fluorescence tags were inserted into the constructs with a flexible linker before the stop codons. Synonymous mutations were introduced to Cas9 target sites to avoid the cleavage of the homologous repair template by Cas9. The plasmids were listed in Extended Data Table 3.

Constructs that express GFP-tagged NuRD components were generated using the PCR SOEing method(Hobert, 2002). N2 genomic sequences (1.5-2 kb promoter plus coding region) were linked with gfp::unc-54 3’ UTR DNA fragments. Constructs that express wrmScarlet-tagged V-ATPase components with gfp::unc-54 3’ UTR were generated using the In-Fusion Advantage PCR cloning kit (Clontech, Cat. #639621). The primers were listed in Extended Data Table 3.

Genome editing and transgenesis

To generate a knock-in strain, we purified the sgRNA construct and the repair template plasmids with the PureLink Quick PCR purification Kit (Invitrogen, #K310001) and co-injected them into N2 animals with the pRF4 [rol-6 (su1006)] and Podr-1::dsRed selection markers. The F1 transgenic progenies were singled and screened by PCR and Sanger sequencing. Transgenic C. elegans was generated by the germline microinjection of DNA plasmids or PCR products into N2 with Podr-1::gfp or Pegl-17::myri-mCherry and Pegl-17::mCherry::his-24 plasmids. We maintained at least two independent transgenic lines with a constant transmission rate (>50%). Concentrations of DNA constructs used for generating knock-in were 50 ng/µl, and for transgenesis were 20 ng/µl. All the strains, primers, plasmids, and PCR products were listed in Extended Data Table 1-3, respectively.

Live-cell imaging and quantification

Live imaging of C. elegans embryos, larvae, or young adults followed our established protocols (Chai et al., 2012; Zhang et al., 2017). L1 larvae or young-adult worms were anesthetized with 0.1 mmol/L levamisole in M9 buffer, mounted on 3% (wt/vol) agarose pads maintained at 20°C. Our imaging system includes an Axio Observer Z1 microscope (Carl Zeiss MicroImaging) equipped with a Zeiss 100×/1.46 numerical aperture (NA) objective, an Andor iXon+ EM-CCD camera, and 488 and 561-nm lines of a Sapphire CW CDRH USB Laser System attached to a spinning disk confocal scan head (Yokogawa CSU-X1 Spinning Disk Unit). Images were acquired by μManager (https://www.micro-manager.org). Time-lapse images of Q cell divisions were acquired with an exposure time of 200 msec every 1 min. Recent advancements in optical and camera technologies permit the acquisition of Z-stacks without perturbing Q cell division or overall animal development. Z-stack images were acquired over a range of -1.6 to +1.6 μm from the focal plane, at intervals of 0.8 μm. The field-of-view spaned 160 μm × 160 μm, and the laser power, as measured at the optical fiber, was approximately 1 mW. ImageJ software (http://rsbweb.nih.gov/ij/) was used to perform image analysis and measurement. Image stacks were z-projected using the average projection for quantification and using the maximum projection for visual display. To follow the cell fate of QR.aa in WT, BafA1- or BafA1 and IAA-treated worms, time-lapse images were acquired with 300 msec for 561 nm and 100 msec for 488 nm exposure time at 5-minute intervals. To image Phsp-16.41::sAnxV::GFP positive cells in WT, mutant, or RNAi animals, we incubated the eggs at 33 °C for 30 min to induce the reporter expression 3 hours post-egg-laying. After we kept the eggs at 20 °C for 2 hours, we imaged them with an exposure time of 200 msec. To acquire time-lapse images of Pegl-1::NLS::GFP positive cells during embryonic development, we dissected 2-4 cell-stage embryos from gravid adults and imaged them with an exposure time of 300 msec every 5 min.

The quantifications of cellular fluorescence intensity ratios in Q cell lineages were described in Figure S4A. We used the mCherry-labeled plasma membrane to circumscribe Q cells (Region of Interest). To determine the ratios of fluorescence intensities in the posterior to anterior half (P/A) of Q.a lineages or A/P of Q.p lineages, the cell in the mean intensity projection was divided into posterior and anterior halves. ImageJ software was used to measure the mean fluorescence intensities of two halves with background subtraction. The slide background’s mean fluorescence intensity was measured in a region devoid of worm bodies. The background-subtracted mean fluorescence intensities of the two halves were divided to calculate the ratio. The same procedure was used to determine the fluorescence intensity ratios between two daughter cells. Total fluorescence intensity was the sum of the posterior and anterior fluorescence intensities or the sum of fluorescence intensities from two daughter cells (Figure S4A). The ROIs for measuring the mean fluorescence intensities of nuclei were delineated as areas containing mCherry-tagged histone within each cell. The background-subtracted mean fluorescence intensities of the two nuclei were divided to calculate the nuclear intensity ratios.

Feeding RNAi

RNAi bacteria were grown in LB media containing carbenicillin (50 µg/mL) and tetracycline hydrochloride (12.5 µg/mL) at 37°C overnight. RNAi bacteria were then seeded on NGM plates supplemented with 50 µg/mL carbenicillin and 1 mM Isopropyl β-D-thiogalactopyranoside (IPTG). We grew the seeded RNAi plates for 2 days at room temperature, allowing double-stranded RNA expression. To score phenotypes during embryogenesis or meiosis, we transferred the synchronized young adult worms to the culture plate containing the induced E. coli HT115 strains expressing the control (luciferase) or specific gene-targeted dsRNAs. We incubated them for 16-24 hours at 20°C and then scored the phenotypes. As for RT-qPCR and RNA-Seq experiments, parental animals of RNAi were used for laying eggs, and the progeny was collected 5 hours post-egg-laying. In ChIP-Seq experiments, we used worms of mixed stages under RNAi treatment for 24-36 hours. RNAi bacterial strains were from the Vidal library.

Immunoprecipitation and mass spectrometry

GFP transgenic or knock-in worms were raised on one hundred 90-mm NGM plates. Animals were collected and washed with M9 buffer three times. For each replicate, the lysate was made from 1∼2 ml packed worms with 3∼4 ml of 0.5-mm diameter glass beads using FastPrep-24 (MP Biomedicals) in lysis buffer [pH 7.4, 150 mM NaCl, 25 mM Tris-HCl, 10% glycerol, 1% NP-40, 1x cocktail of protease inhibitors from Roche (Complete, EDTA free), 40 mM NaF, 5 mM Na3VO4]. Worm lysates were then cleared by centrifugation at 14,000g for 30 min at 4 °C. For anti-GFP immunoprecipitation (IP), the supernatant was incubated with GFP-Trap A beads (Chromoteck, GTA20) for 1 hr. Beads were then washed three times with lysis buffer. To detect the interaction between the V-ATPase subunit and HDA-1, beads are boiled with 1X SDS-PAGE sample loading buffer for SDS–PAGE and western blot. Otherwise, proteins were eluted from beads with 300 μl 0.1 M glycine-HCl (pH 2.5) Add 15 μl 1.5 M Tris-HCl (pH 8.8) to the eluted protein, followed by precipitation using 100 μl trichloroacetic acid and re-dissolution in 60 μl 8 M Urea, 100 mM Tris-HCl (pH 8.5) for MS. MS Samples were then reduced in 5 mM TCEP, alkylated by 10 mM iodoacetamide, and diluted four-fold with 100 mM Tris-HCl (pH 8.5). The 0.2 μg trypsin was used to digest 10 μg protein in 1 mM CaCl2 and 20 mM methylamine overnight at 37°C. The resultant peptides were desalted by ZipTip pipette tips (Merck Millipore). For liquid chromatography-tandem mass spectrometry analysis, peptides were separated by an EASY-nLCII integrated nano-HPLC system (Proxeon, Odense, Denmark) directly interfaced to a Thermo Scientific Q Exactive mass spectrometer (Thermo Fisher Scientific). Peptides were loaded on an analytical fused-silica capillary column (150 mm in length, 75 mm in internal diameter; Upchurch) packed with 5 mm, 300 Å C-18 resin (Varian). Mobile phase A consisted of 0.1% formic acid, while mobile phase B consisted of 0.1% formic acid and 100% acetonitrile. The Q Exactive mass spectrometer was operated through Xcalibur 2.1.2 software in data-dependent acquisition mode. A single full-scan mass spectrum was used in the orbitrap (400–1,800 m/z in mass range, 60,000 in resolution) followed by ten data-dependent tandem mass spectrometry scans at 27% normalized collision energy (HCD). The tandem mass spectrometry spectra results were searched against the C. elegans proteome database by the Proteome Discoverer (version PD1.4; Thermo Fisher Scientific).

Generation of C. elegans embryonic cell suspensions for SPLiT-Seq

We prepared single cells from a C. elegans strain (smIs89) that expressed a Pegl-1::NLS::GFP reporter using the established protocols with minor modifications(Bianchi and Driscoll, 2006; Zhang et al., 2011). Synchronized gravid adults were transferred to NGM plates to lay eggs for 1 hour, and about 50 μl egg pellets were collected 3 hours post-laid. We washed eggs three times with 1 ml sterile egg buffer [118 mM NaCl, 48 mM KCl, 3 mM CaCl2, 3 mM MgCl2, 5 mM HEPES (pH 7.2)]. We rinsed the pellet with 250 μl 1 U/ml Chitinase three times. Eggshells were digested with 250 μl of 1 U/ml chitinase for 45 min on a rotter. We stopped the reaction by adding 750 µl of L-15/FBS (L-15 medium containing 10% fetal bovine serum). Eggs were pelleted by centrifugation at 900 g for 3 min. The supernatant was displaced with 500 µl L-15/FBS. Embryos were dissociated by gently pipetting up and down 60 times using an insulin syringe (29 G needle). The dissociated cells were separated from undissociated cell clumps by filtering through a 5 µm mesh filter, and we washed the filter with 500 µl L-15/FBS. Cells were collected by centrifuging 1 ml filtered suspension at 3220 g for 3 min at 4°C. We removed the supernatant and gently resuspended cells with 2 ml 1X PBS containing 5 μl SUPERaseIn RNase Inhibitor (Invitrogen) and 2.5 μl RiboLock RNase Inhibitor (Thermo Scientific).

SPLiT-Seq library preparation and sequencing

We performed SPLiT-Seq as previously described (Rosenberg et al., 2018) with minor modifications. After cell fixation, centrifugations were performed at 3220 g for 3 min according to the C. elegans cell size. The RT primers were poly-dT and random hexamer mixture with UMIs (Unique Molecular Identifiers). All the barcodes used in this study were described in the published protocol(Rosenberg et al., 2018). The first 24 barcodes were used in the first round of barcoding, named “Round1_01” to “Round1_24”, and 48 barcodes were used in the third round of barcoding, called “Round3_49” to “Round3_96”. Each cell barcode came from three rounds of barcoding. TruePrep DNA Library Prep Kit V2 constructed the sequencing library for Illumina (Vazyme Biotech #TD502) according to the manufacturer’s instructions. The library was sequenced on HiSeq systems (Illumina) using 150 nucleotides (nts) kits and paired- end sequencing.

SPLiT-Seq data processing

According to the results of FastQC, adaptors or low-quality nucleotides were trimmed by Trim Galore (v0.5.2) using the default parameters. For each paired-end sequencing read, a 10-bp UMI sequence and a 24-bp cell barcode were extracted from the Read 2 file by the tool “preprocess_splitseq.pl” of zUMIs (v0.0.6). The Read 1 was split by different cell barcodes in Read 2 and mapped to the modified C. elegans genome (WS263 with an artificial chromosome containing the C. elegans version of the GFP sequence) by zUMIs (v0.0.6) and STAR (v2.6.0c). Unique mapping reads were kept. The duplicated reads from the same transcript were excluded based on the UMI information in Read 2. Low-quality transcriptomes were removed from the analysis, yielding 442 single-cell transcriptomes, including transcript counts from 8629 genes. Among these transcriptomes, 41 transcriptomes were assigned to the putative apoptotic cell based on the expression of egl-1 or gfp. Among the total 8624 genes detected in SPLiT-seq, transcripts from 6487 genes were not detected in the putative apoptotic cell population. Enriched GO terms of those 6487 genes were identified using the Gene Ontology knowledgebase (Ashburner et al., 2000; Gene Ontology, 2021) (http://geneontology.org/).

Cell lineage tracing and quantification of reporter expression

Embryo mounting, live imaging, lineage tracing, and expression quantification were performed according to a previously described procedure with minor modifications (Bao and Murray, 2011; Du et al., 2014; Murray et al., 2008). Two- to four-cell stage embryos were collected from young adult worms and mounted between two coverslips in the egg buffer containing 20-30 20-μm polystyrene microspheres and sealed with melted Vaseline (Bao and Murray, 2011). 3D time-lapse imaging was performed at 20 °C ambient temperature using a spinning disk confocal microscope (Revolution XD) for 300-time points at a 75-second interval. For each time point, embryos were scanned for 30 Z focal planes with 1 µm spacing. 3D tiff stack images were processed with the StarryNite software for automated cell identification and tracing to reconstruct embryonic cell lineages using the ubiquitously–expressed mCherry fluorescence. The raw results of cell identification and tracing were subjected to extensive manual inspection and editing using the AceTree program to ensure high accuracy (Du et al., 2014). For each mentioned strain, the cell lineage was traced from the two- or four-cell stage to the 350-cell stage for three or four embryos (experimental replicates). Regions of the chromatids or nuclei are determined by the H2B::mCherry signal. The fluorescent intensity of HDA-1::GFP or LIN-53::mNeonGreen in each traced nucleus at each time point was measured as the average fluorescent intensity of all the pixels within each nucleus and then with the average fluorescent intensity of the local background subtracted (Murray et al., 2008).

RNA sequencing

Synchronized young adult worms were cultured on RNAi plates for 24 hours. Gravid adults were transferred to new RNAi plates to lay eggs for 1 hour. Eggs were harvested 5 hours post-laid and then lysed with TRIzol reagent (Invitrogen). We extracted the total RNA following the manufacturer’s protocol. The Qubit RNA High Sensitivity Assay Kit (Invitrogen) was used to quantify RNA concentration. The Agilent 2100 bioanalyzer system was used for the assessment of RNA quality. We used samples with an RNA integrity number (RIN) above 6.0 for sequencing library construction. We used identical input total RNA (50 ng to 500 ng) between control and samples for library preparation using the KAPA RNA HyperPrep Kit (KAPA Biosystems, Wilmington, MA, USA). Libraries were analyzed by the Agilent 2100 bioanalyzer system for quality control. The library samples were sequenced on an Illumina NovaSeq 6000 platform. Approximately 5 GB of raw data were generated from each sample with 150 bp paired-end read lengths.

RNA-Seq data analysis

FastQC assessed the quality score, adaptor content, and duplication rates of sequencing reads. Trim_galore (v0.6.0) was used to remove the low-quality bases and adaptor sequences with default parameters. After trimming, paired-end reads with at least 20 nucleotides in length were aligned to the C. elegans reference genome (WS263) using STAR (2.5.4b) with default parameters. Read counts representing gene expression levels were calculated with HTSeq (version 0.9.1) using default parameters. Uniquely mapped reads were included. Gene names were annotated on the Wormbase website (https://wormbase.org/tools/mine/simplemine.cgi). DESeq2 package in R programming language was used for differential expression analysis. Differentially expressed genes were defined as upregulated genes (log2-transformed fold change greater than 1) or downregulated genes (log2-transformed fold change less than -1) with a false discovery rate (FDR) less than 0.05. We used the ggplot2 package of R to plot figures.

Reverse transcription and quantitative real-time PCR

Young adult worms were cultured on RNAi plates for 24 hours. We transferred the gravid adults to new RNAi plates to lay eggs for 1 hour. Eggs were harvested 5 hours post-laid and then lysed with TRIzol reagent (Invitrogen). The RNeasy Mini kit (QIAGEN) prepared the total RNA from ∼20 μl egg pellets for each biological replicate. cDNA from the same amount of RNA between RNAi and control samples were synthesized in a 20 μl reaction volume by the PrimeScript™ RT reagent Kit with gDNA Eraser (TAKARA, Code No. RR047A). One μl of cDNA was used as the template in a 10 μl reaction volume of the PowerUp SYBR Green Master Mix (Applied Biosystems) with four technical replicates. Quantitative real-time PCR was performed in two independent experiments with three biological replicates each time using the Applied Biosystems QuantStudio 1 Real-Time PCR System and normalized to the cell division cycle-related GTPase encoding gene cdc-42. Data were analyzed using the standard curve method. Primers used in real-time PCR assays were listed in Extended Data Table 3.

Chromatin immunoprecipitation

For ChIP assays, chromatin immunoprecipitation was performed as previously described(Mukhopadhyay et al., 2008) with modifications. Mixture stage worms on RNAi plates were harvested and washed three times with M9 buffer. 0.4 ml of packed worms were obtained and frozen into small balls with liquid nitrogen for each ChIP replicate. We quickly ground the little worm balls into powders in liquid nitrogen. Worm powder was crosslinked in 4 ml crosslinking buffer (1.1% formaldehyde in PBS with protease and phosphatase inhibitors) with constant rotation for 15 min at room temperature. Fixed worm samples were quenched by incubation with 0.125 M glycine for 5 min and then washed in cold PBS-PIC (PBS buffer with protease inhibitor cocktail) three times. The tissue pellet resuspended in PBS-PIC was homogenized by 2-3 strokes in a Dounce homogenizer. The cell suspension was centrifuged at 2,000 g for 5 min at 4°C, and we removed the supernatant.

We performed nuclei preparation, chromatin digestion, chromatin immunoprecipitation, and DNA Purification using SimpleChIP® Enzymatic Chromatin IP Kit (Magnetic Beads) (Cell Signaling Technology, #9003). Samples were sonicated using a Qsonica sonicator (Qsonica Q800R) at 70% amplitude for 5 cycles of 30 sec on and 30 sec off with 4 repeats. The chromatin was digested using 6 μl of micrococcal nuclease (MNase) (CST catalog number 10011S) in 300 μl of buffer B (CST catalog number 7007) containing 0.5 mM DTT for 8 min at 37 °C. The quality of chromatin digestion was analyzed by the Agilent 2100 bioanalyzer system. Before IP, the chromatin sample concentration was assessed by the BCA Protein Assay Kit (Tiangen, cat no. PA115-02). 1 μg H3K27ac antibody (Abcam ab4729) or 2 μg IgG was used per IP with 0.5-1 mg input chromatin protein.

According to the manufacturer, we used 200 ng purified DNA samples to construct the ChIP–seq library using the VAHTS Universal DNA Library Prep Kit for Illumina V3 (Vazyme ND607). DNA libraries were sequenced on an Illumina NovaSeq 6000 platform, with 150-bp paired-end sequencing. Three biological replicates were collected for two independent ChIP–seq experiments. For ChIP–qPCR analysis of H3K27ac enrichment in the egl-1 locus, the same volume of cDNA from the IP experiment described above was used for qPCR analysis. Extended Data Table 3 listed the primers.

ChIP-Seq data processing

150-bp double-end reads were aligned to the C. elegans genome version WBcel235 using Bwa-mem2 version 2.1 with default parameters. Only uniquely mapped reads were kept. Samples were normalized by bins per million mapped reads (BPM) and averaged, and then treated samples were subtracted by input samples to calculate enrichment using deepTools2 (Ramirez et al., 2016). Data were visualized using IGV(Robinson et al., 2011).

Immunofluorescence

We performed immunofluorescence of C. elegans eggs using the freeze-cracking method. In brief, eggs were washed with M9 buffer three times and dripped onto poly-L-lysine coated slides. After freezing in liquid nitrogen for 5 minutes, the coverslip was swiftly removed, and embryos were fixed in methanol at -20 °C for 15 min. We washed the slides twice with PBST (PBS+ 0.05% Tween 20). We incubated slides with 1% BSA in PBST for 30 min at room temperature to block the unspecific bindings. The primary and secondary antibodies were diluted in PBST plus 1% BSA by 1:500-1:2,000 for primary antibodies (ab10799 for histone H3 and ab4729 for H3K27ac) and 1:5,000 for the secondary antibodies (Alexa-488 anti-mouse and Alexa546 anti-rabbit from Invitrogen). The primary antibody was incubated overnight at 4°C in a moisture chamber, and the secondary antibody was incubated for 1 hour at room temperature. After washing with PBST, we applied 1-2 drops of the Fluoroshield Mounting Medium containing DAPI (Abcam, ab104139) and coverslips to the samples.

Pharmacological and chemical treatments

To block V-ATPase proton pump activity in the L1-L2 larvae, which have cuticle layers that may compromise agent penetration, the Bafilomycin A1 treatment was designed based on the published protocols (Kumsta et al., 2017; Papandreou and Tavernarakis, 2017; Pivtoraiko et al., 2010). Briefly, gravid worms were placed in 1 drop of ddH2O containing 3% (v/v) pelleted OP50 and 100 μM bafilomycin A1 (Abcam, dissolved in DMSO) for 16 h at 20 ℃. After incubation, L1 larvae were washed with M9 buffer and recovered for 5 h on NG agar plates before imaging.

To achieve Q cell conditional HDA-1 depletion, the auxin-inducible degradation (AID) was performed as described previously (Zhang et al., 2015). The natural auxin (indole-3-acetic acid, IAA) was purchased from Alfa Aesar (#A10556). A 400 mM solution in ethanol was stored at 4°C as stock. Before imaging, L1 larvae were transferred to the S basal buffer supplemented with 3% (v/v) pelleted OP50 and 4 mM of auxin for a 1.5 h treatment.

Coimmunoprecipitation and Western blot

For immunoprecipitation of endogenous V-ATPase A subunit with HDA-1, HDA-1-GFP knock-in worms were lysed in ice-cold lysis buffer [pH 7.4, 150 mM NaCl, 25 mM Tris-HCl, 10% glycerol, 1% NP-40, 2x cocktail of protease inhibitors from Roche (Complete, EDTA free), 40 mM NaF, 5 mM Na3VO4]. The soluble fractions from worm lysates were immunoprecipitated with anti-GFP GFP-Trap A beads (Chromoteck, GTA20) for 1 hr at 4 °C. Immunoprecipitates were washed three times with lysis buffer containing 1x cocktail of protease inhibitors from Roche (Complete, EDTA free) and were then transferred to new tubes. Beads were washed again with 100 mM PB (pH 6.0, 6.5, 7.0, or 7.5) and boiled in 1X SDS loading buffer under 95 ℃ for 5min. The protein samples were analyzed by Western blotting.

Protein samples were resolved by SDS–PAGE and transferred to PVDF membranes. Each membrane was divided into two parts to be incubated with GFP (Abcam ab290, 1:4,000) or ATP6V1A (Abcam ab199326, 1:2,000) primary antibodies, followed by HRP-conjugated Goat Anti-Rabbit IgG (H&L) (Easybio, 1:5,000) secondary antibodies.

Quantification of the endogenous V-ATPase A subunit interacting with HDA-1 was performed using HDA-1::GFP bands as a calibration standard. Quantitative densitometry of chemiluminescent bands was performed using ImageJ software.

Data availability

The data sets generated and analyzed in this study are available in the NCBI Gene Expression Omnibus (GEO, http://www.ncbi.nlm.nih.gov/geo) under accession number GSE167379 with the secure token uxwleayqnjkvnkf.

Supporting information